---
title: "Processing targeted LC-MS/MS lipidomics and metabolomics with MRManalyzeR"
author:
- name: Matthew J. Smith
  affiliation: Karolinska Institutet, Stockholm, Sweden
  email: mattyjsmith123@gmail.com
date: "`r Sys.Date()`"
package: MRManalyzeR
output:
  BiocStyle::html_document:
    toc_float: true
    number_sections: true
vignette: >
  %\VignetteIndexEntry{Processing targeted LC-MS/MS lipidomics and metabolomics with MRManalyzeR}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup-knitr, include = FALSE}
knitr::opts_chunk$set(comment = "#>", fig.width = 6.5, fig.height = 4,
                      dpi = 100, message = FALSE, warning = FALSE)

# Tables are shown the way the HTML reports show them: searchable,
# sortable and scrollable rather than truncated with head().
dt <- function(x, rownames = TRUE, page = 5){
  DT::datatable(x, rownames = rownames, filter = "top",
                options = list(pageLength = page, scrollX = TRUE,
                               dom = "ftip"))
}
```

# Introduction

Targeted LC-MS/MS is analytically less complex than untargeted profiling. The
compound panel is defined in advance, every analyte has an authentic standard,
and the questions that remain are about **data quality** and the **accurate
reporting of known metabolites** rather than about feature discovery.
Normalisation to stable-isotope internal standards is commonplace, and reported
values are true concentrations only once they have been adjusted for the
internal-standard concentration each vial actually carried and for the volume
of sample the extract came from.

MRManalyzeR is built around exactly that: assessing data quality on the way to
accurate quantification, and giving a basic interpretation of the analytes that
come out. The output is deliberately visual -- quality control and statistics
are presented as plots with simple statistical summaries beside them, not as a
modelling framework. Everything is driven from a curated `.xlsx` workbook and a
single YAML file, so a run is reproducible and the workflow stays accessible to
analysts who do not write R.

The package reads Waters **TargetLynx** exports (the primary format) and
**Skyline** exports, and covers signal filtering, blank filtering,
normalisation, concentration adjustment, missing-value imputation and batch
correction, followed by quality control, statistics and two self-contained HTML
reports.

## How the package is organised

The workflow falls into four stages. Everything from stage 1 onwards is carried
in a single `struct::DatasetExperiment` -- the assay matrix, the sample
metadata and the feature metadata kept aligned in one object -- so every
function takes that object and returns it, and nothing has to be handed a
matching metadata table a second time.

| Stage | Question it answers | Functions |
|---|---|---|
| 1. Data parse | What did the instrument report? | `read_targetlynx()`, `read_skyline()`, `assemble_dataset()`, `load_config()`, `load_dataset()`, `combine_datasets()` |
| 2. Peak-matrix processing | What is the true concentration? | `filter_blanks()`, `normalise_matrix()`, `adjust_concentration()`, `impute_missing()`, `correct_batch()`, `subset_dataset()`, `process_dataset()` |
| 3. QC check | Can these numbers be trusted? | `add_cv_metrics()`, `summarise_features()`, `run_pca()`, `plot_drift()`, `plot_qq()`, `plot_pca()`, `plot_loadings()` |
| 4. Stats | What do these numbers mean? | `run_stats()`, `summarise_groups()`, `plot_boxplot()`, `plot_volcano()`, `plot_heatmap()`, `plot_correlation()` |

```
   project_config.yml            workbook.xlsx
           |                           |
           |                  read_targetlynx() / read_skyline()   [1]
           |                           |
           |                  wide sample x compound matrix
           |                           |
           |                    assemble_dataset()
           |                           |
           |                           v
           |                   DatasetExperiment
           |                           |
           |          filter_blanks -> normalise_matrix ->         [2]
           +--------> adjust_concentration -> impute_missing
                      then correct_batch
                                       |
                                       v
                          processed DatasetExperiment
                            |          |          |
             [3] QC check   |          |          |   [4] Stats
             add_cv_metrics |          |          |   summarise_groups
             run_pca        |          |          |   run_stats
             summarise_     |          |          |
               features     |          |          |
                            |          |          |
              tables + plots|          |          |tables + plots
                            v          v          v
              data_quality       .xlsx  .RDS      stats_report
                _report.html                      .html
                                       |
                                  _stats.xlsx
```

Every number in a figure comes from the same table that is written to the
spreadsheet -- `run_stats()`, `summarise_groups()` and `add_cv_metrics()`
compute once, the `plot_*()` functions consume. A figure and the sheet beside
it cannot disagree.

# Installation

```{r install, eval = FALSE}
if (!requireNamespace("BiocManager", quietly = TRUE))
    install.packages("BiocManager")
BiocManager::install("MRManalyzeR")
```

The development version:

```{r install-dev, eval = FALSE}
BiocManager::install("MJS-708/MRManalyzeR")
```

```{r load}
library(MRManalyzeR)
```

# Quick start

`run_example()` runs all four stages on the bundled dataset and writes its
outputs to a temporary directory. `render = FALSE` builds the peak matrix and
the xlsx/RDS outputs but skips the (slower) HTML report rendering; drop it to
get the reports as well. Running this successfully is the recommended check
that an installation works before pointing the package at real data.

```{r run-example}
res <- run_example(render = FALSE, open = FALSE)
list.files(attr(res, "results_dir"))
```

Everything after this section unpacks what that single call did.

# Stage 1: data parse

## The curated workbook

MRManalyzeR expects one `.xlsx` workbook holding the instrument export plus the
two curation sheets that describe it:

| Sheet | One row per | Purpose |
|---|---|---|
| `lcms_data` (or `skyline_data`) | peak | the instrument export itself |
| `feature_metadata` | compound | which compounds to report, and their annotation |
| `sample_metadata` | injection | which injections to include, and their design |

Keeping the curation next to the data is what makes a run reproducible: the
decisions about which compounds are reportable and which injections belong to
the study are recorded in the workbook, not in a script.

Two practical rules govern whether a run will work at all:

* **Names must match exactly.** `feature_metadata$Processing_name` has to match
  the compound names in the acquisition method, and `sample_metadata$Name` has
  to match the injection names in the data sheet -- including trailing spaces,
  which are the usual culprit.
* **Exclusions are recorded, not deleted.** An injection you do not want
  analysed gets `Include = "NO"` rather than being removed; a compound that
  should not be reported gets `Report = "NO"`. Re-injections are the common
  case, and marking them keeps the workbook a faithful record of the run.

```{r dataset}
xlsx  <- system.file("extdata", "example_data.xlsx",
                     package = "MRManalyzeR")
fdata <- openxlsx::read.xlsx(xlsx, sheet = "feature_metadata")
meta  <- openxlsx::read.xlsx(xlsx, sheet = "sample_metadata")

c(injections = nrow(meta), compounds = nrow(fdata))
```

The bundled workbook is a subset of the oxylipin LC-MS/MS data of Kolmert
*et al.* (2018) (see [References](#references)): 79 injections of mouse
bronchoalveolar lavage over a calibration ladder, extraction blanks, pooled QC
injections and 45 study samples.

```{r dataset-samples}
dt(as.data.frame(table(Sample_type = meta$Sample_type)),
   rownames = FALSE)
```

Because the panel is known in advance, `feature_metadata` can carry real
biology alongside the reporting decision -- the enzymatic pathway a compound
comes from, its precursor PUFA, its retention time. Those columns are not
decoration: they colour the PCA loadings, group the heatmap rows and define the
ion ratios, and they travel with the data as `variable_meta` all the way to the
output.

```{r dataset-features}
dt(fdata[fdata$Report == "YES",
         c("Compound", "Enzymatic_pathway", "PUFA", "RT", "Report")],
   rownames = FALSE)
```

## The TargetLynx export

TargetLynx does not export a matrix. Its "complete summary" (`Edit > Copy >
Complete Summary`, pasted straight into the `lcms_data` sheet) is a *stacked*
table -- one block per compound, each block its own little table of injections,
with the compound name in a header row between blocks. It is awkward to read by
eye and worse to read programmatically, which is the first problem the package
solves.

`read_targetlynx()` parses those blocks, pivots them into a wide
**sample x compound** matrix of the requested `datatype` column, and (with
`snr`) masks any peak whose signal-to-noise falls below the threshold:

```{r read}
m <- read_targetlynx(xlsx, datatype = "Area", snr = 3)
dim(m)
```

```{r read-table}
dt(round(m[, 1:6], 1))
```

Columns come back as the raw TargetLynx compound identifiers
(`Processing_name`). `feature_metadata` maps those to the canonical `Compound`
name used everywhere downstream, which is how an instrument method can be
renamed without touching the analysis.

`read_skyline()` returns the same contract from a Skyline molecule x sample
sheet, with `signal_filter = "LOD"` or `"LOQ"` in place of the S/N mask.

## The simulated dataset used from here on

The example above is a single chromatographic batch, injected from one
reconstitution volume, with no per-sample amount to normalise to -- so
normalisation, concentration adjustment and batch correction would all be
no-ops on it. The package therefore bundles a second, **entirely simulated**
workbook: a Skyline-layout export of 20 analytes across three enzymatic
pathways, 54 injections split over two chromatographic batches, with a
per-sample protein amount and a different reconstitution volume per batch.

Nothing in it is measured data. It is generated from a fixed seed by
`data-raw/make_example_synthetic.R` (which ships with the package source), and
what it contains was deliberately planted so each step can be seen recovering
something known: a response shift between the two batches, a treatment effect
on 5 of the 20 analytes -- recorded in `feature_metadata$Simulated_log2FC` --
and QC injections drawn with about 8 percent noise against 35 percent between
animals. The study design is balanced across the batches, so the batch effect
and the treatment effect are not confounded.

```{r syn-data}
sxlsx <- system.file("extdata", "example_synthetic.xlsx",
                     package = "MRManalyzeR")
sfeat <- openxlsx::read.xlsx(sxlsx, sheet = "feature_metadata")
smeta <- openxlsx::read.xlsx(sxlsx, sheet = "sample_metadata")

table(smeta$Chrom_Batch, smeta$Sample_type)
```

## Assembling a DatasetExperiment

`assemble_dataset()` aligns the matrix with its feature and sample metadata and
returns a `struct::DatasetExperiment`. **Every step from here on takes and
returns that object**, so the metadata travels with the data and no step has to
be handed a matching sample table again.

```{r step-assemble}
sm <- read_skyline(sxlsx, "skyline_data", sfeat, signal_filter = "LOD")
de <- assemble_dataset(sm, sfeat, smeta)
de
```

# Stage 2: peak-matrix processing

## Blank filter

`filter_blanks()` finds the extraction blanks via `sample_meta` and sets any
value below `mean(blanks) * blank_filter` to `NA`. A factor of 3 is a common
choice -- a peak has to be at least three times the blank level to be treated
as real.

```{r step-blanks}
de <- filter_blanks(de, blank_filter = 3, blank_head = "Sample_type",
                    blank_name = "Blank")
sum(is.na(de$data))
```

## Normalise

`normalise_matrix()` divides each sample by a known per-sample amount --
protein, tissue weight, or the volume of biofluid extracted -- so that
concentrations are expressed per unit of starting material:

```{r step-normalise}
de <- normalise_matrix(de, column = "protein_ug")
round(as.data.frame(de$data)[1:3, 1:3], 3)
```

## Concentration adjustment

A calibration curve returns the concentration *in the vial the instrument
sampled*, and it does so on the calibrant's terms: the curve was built at the
calibrant's internal-standard concentration, and it silently assumes your
sample carries the same one. `adjust_concentration()` corrects that assumption:

$$
\text{conc} \times \frac{\text{cal vol}}{\text{sample vol}}
            \times \frac{\text{sample IS}}{\text{cal IS}}
$$

Those two ratios are one thing, not two. Multiply them out and they collapse to
a single quantity:

$$
\frac{\text{cal vol}}{\text{sample vol}} \times
\frac{\text{sample IS}}{\text{cal IS}}
\;=\;
\frac{\text{sample IS} / \text{sample vol}}{\text{cal IS} / \text{cal vol}}
\;=\;
\frac{[\text{IS}] \text{ in the sample vial}}
     {[\text{IS}] \text{ in the calibrant vial}}
$$

So the correction is the ratio of internal-standard *concentrations*. More vial
volume per unit of IS means the IS is more dilute in your sample than in the
calibrant; a smaller IS peak inflates the analyte/IS response ratio, the curve
reads high, and this factor -- below 1 -- scales it back down.

Note what is *not* being corrected here. Reconstituting in more solvent dilutes
the analyte and the internal standard equally, so the response ratio does not
move and neither does the curve's answer. The reconstitution volume matters
only through the IS concentration, which is exactly what the expression above
captures. (This reasoning depends on quantification being IS-ratio based, as it
is for a targeted assay with stable-isotope standards. Pointing `adjust_conc`
at a raw `Area` datatype with no IS correction would break the assumption.)

Setting `starting_vol_col` adds a second step, back to the concentration in the
original, undried sample -- the plasma or lavage volume actually taken into
extraction -- so results are reported per unit of biofluid rather than per vial.
Following one analyte amount `A` through both steps:

```
  curve gives     conc_reported = (A / IS_sample) x [IS] in calibrant
  step 1 gives    conc_vial     = A / sample_vol
  step 2 gives    conc_original = A / starting_vol
```

`sample_vol` cancels between the two steps, and it should: the final number
depends only on how much analyte was there and the volume of biofluid it came
from, never on how much solvent you happened to reconstitute in.

```{r step-adjust}
de <- adjust_concentration(de,
                           sample_vol_col   = "sample_volume_uL",
                           cal_vol_col      = "cal_vol_uL",
                           sample_IS_col    = "sample_IS_vol_uL",
                           cal_IS_col       = "cal_IS_vol_uL",
                           starting_vol_col = "starting_vol_uL")
round(as.data.frame(de$data)[1:3, 1:3], 5)
```

## Impute missing values

`impute_missing()` replaces the remaining `NA`s in the biological samples with a
fraction of that compound's minimum observed value, treating a non-detect as
lying just below the lowest measurement rather than as a zero or a hole.
`scalar = 0.5` is the familiar half-minimum rule; a smaller fraction (0.2) is
more conservative if the panel has compounds sitting near the detection limit.
Blanks are left untouched -- they are diagnostics, not measurements.

Other strategies are available where they matter most, in the PCA
pre-processing (`run_pca(impute = )`), which offers `"none"`, `"min"`,
`"half_min"` and `"frac_min"` alongside missing-value filters that drop
features or samples above a missingness threshold instead of imputing them.
For targeted panels the choice is rarely critical: a compound missing in most
samples is usually better dropped than modelled.

```{r step-impute}
de <- impute_missing(de, scalar = 0.5, blank_head = "Sample_type",
                     blank_name = "Blank")
sum(is.na(de$data))
```

## Batch correction

Where an acquisition spans several days or columns, response can shift between
batches. `correct_batch()` scales each batch so its per-feature median matches
the grand median across the reference samples:

```{r step-batch}
batch     <- stats::setNames(de$sample_meta$Chrom_Batch,
                             rownames(as.data.frame(de$data)))
de_uncorr <- de

de <- correct_batch(de, qc_label = "QC",
                    factor_name = "Sample_type",
                    batch_head  = "Chrom_Batch")
```

Taking the first analyte as an example, the two batch medians are pulled
together:

```{r step-batch-medians}
med <- function(x) tapply(as.data.frame(x$data)[[1]], batch, median)

rbind(before = med(de_uncorr),
      after  = med(de))
```

Use pooled QC injections as the reference (`qc_label = "QC"`) whenever the
design has enough of them in every batch; falling back to `"Sample"` assumes
the batches are biologically comparable.

## The whole matrix in one call

`process_dataset()` composes stages 1 and 2 -- read, assemble, then filter /
normalise / adjust / impute within each acquisition batch, then batch-correct
-- driven by the same metadata sheets:

```{r step-process}
out <- process_dataset(
  sfeat, smeta,
  xlsx_path        = sxlsx,
  data_source      = "skyline",
  data_tab_names   = "skyline_data",
  signal_filter    = "LOD",
  blank_filter     = 3,
  normalize        = "protein_ug",
  adjust_conc      = TRUE,
  starting_vol_col = "starting_vol_uL",
  replace_MVs      = 0.5,
  batch_correction = TRUE,
  bc_qc_label      = "QC",
  bc_header        = "Chrom_Batch",
  blank_head       = "Sample_type")

dim(out[[1]]$data)
```

`out[[2]]` holds the features dropped along the way with the reason recorded --
the same table the data-quality report surfaces as "compounds excluded", and
the first place to look when a compound you expected is missing from the
output.

The processed dataset also ships ready-made, which is what the rest of this
vignette uses:

```{r syn-load}
syn <- load_dataset(system.file("extdata", "example_synthetic.RDS",
                                package = "MRManalyzeR"))
dim(syn$data)
```

# Stage 3: QC check

## How well was each compound measured?

`add_cv_metrics()` writes the per-compound quality metrics into
`variable_meta`, matched to the feature they describe, so annotation and
quality live in the same table. `run_MRManalyzeR()` applies it to every run.

```{r qc-cv}
syn <- add_cv_metrics(syn, qc_label = "QC")

cv <- as.data.frame(syn$variable_meta)[
  , c("Compound", "Enzymatic_pathway", "CV_QC", "CV_sample",
      "CV_sample_vs_QC")]

dt(cv[order(cv$CV_QC), ], rownames = FALSE, page = 10)
```

* **`CV_QC`** is the coefficient of variation across the pooled QC injections.
  Every QC is the same material, so this is **technical** variability -- the
  measurement noise of that compound in that run. It is the primary
  quantification filter: a common rule of thumb keeps compounds under 20-30
  percent CV~QC~ and reports the rest as detected but not reliably quantified.
  It is `NA` when a run contains no QC injections.
* **`CV_sample`** is the same statistic across the biological samples, so it
  carries the **biological** spread *plus* that same technical noise.
* **`CV_sample_vs_QC`** is `CV_sample / CV_QC`. A ratio comfortably above 1
  means the biological signal is larger than the noise it is measured against
  -- the compound has room to show a real difference. A ratio near 1 means a
  compound is varying no more between animals than between replicate injections
  of identical material: whatever passed the CV~QC~ filter, there is little
  biology left to find in it. Sorting the panel by this ratio is a quick way to
  see which analytes are worth interpreting before running any test.

## Analytical drift

`plot_drift()` puts one compound's response against acquisition order, coloured
by sample type so the pooled QCs can be tracked through the sequence. A trend,
a step or a widening spread in the QCs is instrumental, not biological.

```{r qc-drift}
plot_drift(syn, "Analyte_02", value_label = "ng/mL")
```

## Normality

`plot_qq()` asks whether a compound's distribution supports the parametric
tests. Points along the line mean the t-test is on safe ground; a curve means
the Wilcoxon result reported alongside it is the one to trust. Targeted panels
are frequently right-skewed, which is why the report renders a tab per
transform rather than picking one.

```{r qc-qq, fig.width = 5, fig.height = 4}
plot_qq(syn, "Analyte_02", transform = "log2", levels = "Sample")
```

## PCA

`run_pca()` applies the missing-value filtering, imputation, transform,
centring and scaling, and returns the fit together with a readable trace of
what it did. `plot_pca()` draws the scores.

```{r qc-pca}
pca <- run_pca(syn, impute = "frac_min", impute_frac = 0.2,
               transform = "log2", feat_na_max = 0.6,
               sample_na_max = 0.6)
pca$steps
```

The colouring is the whole point: the same scores answer a different question
depending on what you colour them by. The report draws one plot per column
listed under `qc_pca.color_by`.

```{r qc-pca-type, fig.height = 4.2}
plot_pca(pca, syn, colour_by = "Sample_type")
```

Colouring by chromatographic batch is the sharpest of these -- if the batches
separate, the variance driving component 1 is analytical rather than
biological. This dataset had a batch shift planted in it, so here is the same
PCA before and after `correct_batch()`:

```{r qc-pca-batch, fig.height = 3.4}
pca_by_batch <- function(x, title){
  p <- run_pca(x, impute = "frac_min", impute_frac = 0.2,
               transform = "log2")
  plot_pca(p, x, colour_by = "Chrom_Batch") +
    ggplot2::labs(title = title)
}

cowplot::plot_grid(
  pca_by_batch(de_uncorr, "Before batch correction"),
  pca_by_batch(de,        "After batch correction"),
  nrow = 1)
```

## Loadings

The scores say *whether* samples separate; `plot_loadings()` says *which
compounds* drove it. Colouring by the enzymatic pathway is what turns a ranked
list into something interpretable -- the palette is shared with the heatmap
annotation, so a class is the same colour in every figure.

```{r qc-loadings, fig.width = 7, fig.height = 5}
plot_loadings(pca, syn, colour_by = "Enzymatic_pathway")
```

## What was reported, and what was dropped

```{r qc-features}
dt(summarise_features(syn, out[[2]]), rownames = FALSE)
```

# Stage 4: statistics

## Selecting the biological samples

`subset_dataset()` slices by any `sample_meta` condition, which is how the
study samples are separated from the QCs and blanks before testing:

```{r stats-subset}
samples <- subset_dataset(syn,
                          conditions = list(Sample_type = "Sample"))

table(samples$sample_meta$Treatment, samples$sample_meta$Type)
```

## Group summaries

`summarise_groups()` is the table the box and bar plots are drawn from, and it
is written to the `summary` sheet of the stats workbook. Both SD and SE are
returned: SD describes the spread of the animals, SE how well the mean is
pinned down, and the choice belongs to the reader.

```{r stats-summary}
gs <- summarise_groups(samples, "Treatment")
dt(cbind(gs[, 1:3], round(gs[, 4:6], 3)), rownames = FALSE)
```

## Comparisons

`run_stats()` runs the comparisons, correlations, linear models and ion ratios
described by the `stats_report:` block of the YAML and returns tidy tables. The
method is chosen from the design rather than by the user: two levels get a
t-test and a Wilcoxon test, three or more get ANOVA and Kruskal-Wallis with
Tukey and pairwise-Wilcoxon post-hoc tests, and p-values are FDR-adjusted.

Reporting a parametric and a non-parametric test side by side is deliberate:
targeted panels are frequently run on small groups where normality cannot be
taken on trust, and agreement between the two is itself a piece of evidence.

```{r stats-run}
params <- list(
  sig_threshold = 0.05,
  comparisons = list(enabled = TRUE, entries = list(
    list(name    = "PBS_vs_HDM",
         compare = list(factor = "Treatment",
                        levels = c("PBS", "HDM"))))),
  ion_ratios = list(enabled = TRUE, ratios = list(
    list(num = "Analyte_02", den = "Analyte_05",
         label = "A02/A05"))))

st <- run_stats(samples, params)
```

Because the effects in this dataset were planted, the recovered fold changes
can be checked against the truth:

```{r stats-truth}
tt <- st$stats[st$stats$method == "t_test",
               c("feature", "log2FC", "p_value", "p_adj")]
truth <- as.data.frame(syn$variable_meta)[, c("Compound",
                                              "Simulated_log2FC")]
chk <- merge(tt, truth, by.x = "feature", by.y = "Compound")
chk <- chk[order(-abs(chk$Simulated_log2FC), chk$p_value), ]

dt(cbind(chk[, 1], round(chk[, -1], 4)), rownames = FALSE, page = 8)
```

The five analytes with a non-zero `Simulated_log2FC` should carry the largest
recovered effects and the smallest p-values; the other fifteen should scatter
around zero.

## Per-compound plots

```{r stats-box, fig.width = 5, fig.height = 4}
plot_boxplot(samples, "Analyte_02", "Treatment",
             value_label = "ng/mL")
```

`type = "bar"` draws the group means and error bars from `summarise_groups()`,
so the figure and the `summary` sheet cannot drift apart:

```{r stats-bar, fig.width = 5, fig.height = 4}
plot_boxplot(samples, "Analyte_02", "Treatment", type = "bar",
             error_bar = "SE", value_label = "ng/mL")
```

## Volcano

Effect size against significance. A large fold change on three noisy animals is
not a finding, and a tiny but reproducible shift is rarely interesting either;
the compounds worth attention sit in the upper corners.

```{r stats-volcano, fig.width = 5.5, fig.height = 5}
plot_volcano(st, "PBS_vs_HDM", sig_threshold = 0.05)
```

## Heatmap

Where the comparisons test one compound at a time, the heatmap shows the whole
panel at once -- which is how you see that a treatment moved an entire pathway
rather than a few compounds that happened to clear a threshold.

```{r stats-heatmap, fig.width = 7.5, fig.height = 6}
plot_heatmap(samples,
             color_samples_by  = "Treatment",
             group_features_by = "Enzymatic_pathway",
             transform         = "log2",
             title             = "All analytes")
```

## Feature correlations

Which compounds move *together*, which in a targeted panel usually means they
share a branch of the pathway -- so blocks along the diagonal are the pathway
structure showing itself.

```{r stats-corr, fig.width = 6.5, fig.height = 6}
cp <- list(correlations = list(enabled = TRUE, entries = list(
  list(name = "pairs", methods = "spearman",
       subsets = list(list())))))

plot_correlation(run_stats(samples, cp), samples,
                 group_by = "Enzymatic_pathway")
```

## Ion ratios

An ion ratio is an enzyme-activity surrogate: the ratio of a product to its
precursor tracks flux through a pathway more directly than either compound
alone, because it cancels the shared variation in how much substrate was there.
`run_stats()` returns the per-sample ratios with a group test attached, so they
reach the `ion_ratios` sheet as well as the report.

```{r stats-ratios}
ir <- st$ion_ratios[!is.na(st$ion_ratios$comparison), ]
dt(cbind(ir[, c("ratio", "comparison", "sample", "group")],
         round(ir[, c("value", "p_value")], 4)),
   rownames = FALSE)
```

# Driving it from a YAML config

In routine use none of the above is typed out. One YAML file describes the data
source, every processing step and both reports, and one call runs it:

```{r run-yaml, eval = FALSE}
res <- run_MRManalyzeR("path/to/your_config.yml")
```

The bundled reference configuration is a commented template to copy and edit:

```{r config-path}
system.file("extdata", "example_config.yml", package = "MRManalyzeR")
```

Its `PeakMatrixProcessing:` block is a one-to-one map of stage 2 -- each key
switches a step on or names the metadata column it should use:

```yaml
PeakMatrixProcessing:
  execute: True                  # False = skip, reuse existing RDS

  name_col: Name                 # sample id (matches the matrix rows)
  include_col: Include           # sample include-filter column
  include_value: "YES"

  tl_data:                       # TargetLynx source (or skyline_data:)
    enabled: True
    data_tab_names: [lcms_data]
    tl_headers: ["ID", "Name", "Area", "ng/mL", "Response", "S/N"]
    snr: 3                       # S/N threshold | False

  blank_filter: 3                # n x blank mean | False
  blank_head: 'Sample_type'
  blank_name: 'Blank'

  normalize: protein_ug          # False | sample_metadata column

  adjust_conc:
    enabled: True
    starting_vol_col: starting_vol_uL
    sample_vol_col:   sample_volume_uL
    cal_vol_col:      cal_vol_uL
    sample_IS_col:    sample_IS_vol_uL
    cal_IS_col:       cal_IS_vol_uL

  replace_MVs: 0.5               # False | scalar x per-feature min
  batch_correction: True
  bc_qc_label: "QC"
  bc_factor_name: "Sample_type"
  bc_header: "Chrom_Batch"
```

`data_quality_report:` and `stats_report:` configure stages 3 and 4 in the same
declarative way -- which column identifies QCs, which factors to compare, which
plots to draw. See the bundled file for the fully commented set.

Outputs are written next to your data: the processed matrix as `.xlsx` and
`.RDS`, a `_stats.xlsx` carrying `stats`, `correlations`, `linear_models`,
`ion_ratios` and `summary`, the two self-contained HTML reports, and a text
file recording the exact parameters used.

## Merging several acquisition panels

Where a study is split across methods -- a polar-acid and a polar-basic panel,
say -- each is processed on its own with `run_MRManalyzeR()` first, and the
resulting `.RDS` files are merged:

```{r run-combine, eval = FALSE}
file.copy(
  system.file("extdata", "example_combine_config.yml",
              package = "MRManalyzeR"),
  "combine_config.yml")

res <- run_MRManalyzeR_combine("combine_config.yml")
```

Two things decide whether a merge succeeds: `sample_id_col` must name an
identifier that is stable across the panels (not the acquisition `Name`, which
usually carries a method-specific prefix), and each dataset's `tag` prefixes
its feature names so that a compound measured on two methods does not collide.

# References {#references}

Kolmert J. *et al.* (2018). Prominent release of lipoxygenase generated
mediators in a murine house dust mite-induced asthma model.
*Prostaglandins & Other Lipid Mediators*.
doi:[10.1016/j.prostaglandins.2018.05.005](https://doi.org/10.1016/j.prostaglandins.2018.05.005)
-- source of the bundled example data.

Lloyd G.R., Jankevics A., Weber R.J.M. (2020). struct: an R/Bioconductor-based
framework for standardized metabolomics data analysis. *Bioinformatics*.
doi:[10.1093/bioinformatics/btaa227](https://doi.org/10.1093/bioinformatics/btaa227)
-- the `DatasetExperiment` container used throughout.

# Session info

```{r session-info}
sessionInfo()
```
