---
title: "Processing targeted LC-MS/MS lipidomics and metabolomics with MRManalyzeR"
author:
- name: Matthew J. Smith
  affiliation: Karolinska Institutet, Stockholm, Sweden
  email: mattyjsmith123@gmail.com
date: "`r Sys.Date()`"
package: MRManalyzeR
output:
  BiocStyle::html_document:
    toc_float: true
    number_sections: true
vignette: >
  %\VignetteIndexEntry{Processing targeted LC-MS/MS lipidomics and metabolomics with MRManalyzeR}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup-knitr, include = FALSE}
knitr::opts_chunk$set(comment = "#>", fig.width = 6.5, fig.height = 4,
                      dpi = 100)

# Tables are shown the way the HTML reports show them: searchable,
# sortable and scrollable rather than truncated with head().
dt <- function(x, rownames = TRUE, page = 5){
  DT::datatable(x, rownames = rownames, filter = "top",
                options = list(pageLength = page, scrollX = TRUE,
                               dom = "ftip"))
}
```

# Introduction

Targeted LC-MS/MS is analytically less complex than untargeted profiling. The
compound panel is defined in advance, every analyte has an authentic standard,
and the questions that remain are about **data quality** and the **accurate
reporting of known metabolites** rather than about feature discovery.
Normalisation to stable-isotope internal standards is commonplace, and reported
values are true concentrations only once they have been adjusted for the
volume each extract was reconstituted in and for the levels the calibration
curve was built on.

MRManalyzeR is built around exactly that: assessing data quality on the way to
accurate quantification, and giving a basic interpretation of the analytes that
come out. The output is deliberately visual -- quality control and statistics
are presented as plots with simple statistical summaries beside them, not as a
modelling framework. Everything is driven from a curated `.xlsx` workbook and a
single YAML file, so a run is reproducible and the workflow stays accessible to
analysts who do not write R.

The package reads Waters **TargetLynx** exports (the primary format) and
**Skyline** exports, and covers signal filtering, blank filtering,
normalisation, concentration adjustment, missing-value imputation and batch
correction, followed by statistics and two self-contained HTML reports.

This vignette walks the workflow step by step on two bundled datasets, then
shows the one-call YAML route that routine use actually takes.

## How the package is organised

The workflow falls into four stages. Everything from stage 1 onwards is carried
in a single `struct::DatasetExperiment` -- the assay matrix, the sample
metadata and the feature metadata kept aligned in one object -- so every
function takes that object and returns it, and nothing has to be handed a
matching metadata table a second time.

| Stage | Question it answers | Functions |
|---|---|---|
| 1. Data parse | What did the instrument report? | `read_targetlynx()`, `read_skyline()`, `assemble_dataset()`, `load_config()`, `load_dataset()`, `combine_datasets()` |
| 2. Peak-matrix processing | What is the true concentration? | `filter_blanks()`, `normalise_matrix()`, `adjust_concentration()`, `impute_missing()`, `correct_batch()`, `process_dataset()` |
| 3. QC check | Can these numbers be trusted? | `run_pca()`, per-compound CV metrics, the data-quality report |
| 4. Stats | What do these numbers mean? | `subset_dataset()`, `run_stats()`, the statistics report |

```
   project_config.yml            workbook.xlsx
           |                           |
           |                  read_targetlynx() / read_skyline()   [1]
           |                           |
           |                  wide sample x compound matrix
           |                           |
           |                    assemble_dataset()
           |                           |
           |                           v
           |                   DatasetExperiment
           |                           |
           |          filter_blanks -> normalise_matrix ->         [2]
           +--------> adjust_concentration -> impute_missing
                      then correct_batch
                                       |
                                       v
                          processed DatasetExperiment
                            |          |          |
             [3] QC check   |          |          |   [4] Stats
             CVs, drift,    |          |          |   run_stats():
             Q-Q, QC-PCA    |          |          |   comparisons,
                            |          |          |   correlations,
                            |          |          |   linear models
                            v          v          v
              data_quality       .xlsx  .RDS      stats_report
                _report.html                      .html
                                       |
                                  _stats.xlsx
```

The tables the statistics report shows and the tables written to
`_stats.xlsx` are the same objects, computed once by `run_stats()` -- a figure
and the spreadsheet beside it cannot disagree.

# Installation

```{r install, eval = FALSE}
if (!requireNamespace("BiocManager", quietly = TRUE))
    install.packages("BiocManager")
BiocManager::install("MRManalyzeR")
```

The development version:

```{r install-dev, eval = FALSE}
BiocManager::install("MJS-708/MRManalyzeR")
```

```{r load}
library(MRManalyzeR)
```

# Quick start

`run_example()` runs the whole workflow on the bundled dataset and writes its
outputs to a temporary directory. `render = FALSE` builds the peak matrix and
the xlsx/RDS outputs but skips the (slower) HTML report rendering; drop it to
get the reports as well.

```{r run-example, message = FALSE, warning = FALSE}
res <- run_example(render = FALSE, open = FALSE)
list.files(attr(res, "results_dir"))
```

Everything after this section unpacks what that single call did.

# The input: a curated workbook

MRManalyzeR expects one `.xlsx` workbook holding the instrument export plus the
two curation sheets that describe it:

| Sheet | One row per | Purpose |
|---|---|---|
| `lcms_data` (or `skyline_data`) | peak | the instrument export itself |
| `feature_metadata` | compound | which compounds to report, and their annotation |
| `sample_metadata` | injection | which injections to include, and their design |

Keeping the curation next to the data is what makes a run reproducible: the
decisions about which compounds are reportable and which injections belong to
the study are recorded in the workbook, not in a script.

```{r dataset}
xlsx  <- system.file("extdata", "example_data.xlsx",
                     package = "MRManalyzeR")
fdata <- openxlsx::read.xlsx(xlsx, sheet = "feature_metadata")
meta  <- openxlsx::read.xlsx(xlsx, sheet = "sample_metadata")

c(injections = nrow(meta), compounds = nrow(fdata))
```

The bundled workbook is a subset of the oxylipin LC-MS/MS data of Kolmert
*et al.* (2018) (see [References](#references)): 79 injections of mouse
bronchoalveolar lavage over a calibration ladder, extraction blanks, pooled QC
injections and 45 study samples.

```{r dataset-samples}
dt(as.data.frame(table(Sample_type = meta$Sample_type)),
   rownames = FALSE)
```

Because the panel is known in advance, `feature_metadata` can carry real
biology alongside the reporting decision -- the enzymatic pathway a compound
comes from, its precursor PUFA, its retention time. Those columns are not
decoration: they colour the PCA loadings, group the heatmap rows and define the
ion ratios in the statistics report, and they travel with the data as
`variable_meta` all the way to the output.

```{r dataset-features}
dt(fdata[fdata$Report == "YES",
         c("Compound", "Enzymatic_pathway", "PUFA", "RT", "Report")],
   rownames = FALSE)
```

## The TargetLynx export

TargetLynx does not export a matrix. It writes a *stacked* sheet -- one block
per compound, each block its own little table of injections, with the compound
name buried in a header row between blocks. It is awkward to read by eye and
worse to read programmatically, which is the first problem the package solves.

`read_targetlynx()` parses those blocks, pivots them into a wide
**sample x compound** matrix of the requested `datatype` column, and (with
`snr`) masks any peak whose signal-to-noise falls below the threshold:

```{r read}
m <- read_targetlynx(xlsx, datatype = "Area", snr = 3)
dim(m)
```

```{r read-table}
dt(round(m[, 1:6], 1))
```

Columns come back as the raw TargetLynx compound identifiers
(`Processing_name`). `feature_metadata` maps those to the canonical `Compound`
name used everywhere downstream, which is how an instrument method can be
renamed without touching the analysis.

`read_skyline()` returns the same contract from a Skyline molecule x sample
sheet, with `signal_filter = "LOD"` or `"LOQ"` in place of the S/N mask.

# The workflow, step by step

The example above is a single chromatographic batch, injected from one
reconstitution volume, with no per-sample amount to normalise to -- so
normalisation, concentration adjustment and batch correction would all be
no-ops on it. The package therefore bundles a second, deliberately messier
workbook for this section: a Skyline export of the same compounds split across
two chromatographic batches, with a per-sample protein amount and a different
reconstitution volume per batch.

```{r batch-data}
bxlsx <- system.file("extdata", "example_batch_data.xlsx",
                     package = "MRManalyzeR")
bfeat <- openxlsx::read.xlsx(bxlsx, sheet = "feature_metadata")
bmeta <- openxlsx::read.xlsx(bxlsx, sheet = "sample_metadata")

table(bmeta$Chrom_Batch, bmeta$Sample_type)
```

The peak areas in it are real; the batch shift, protein amounts and volumes are
simulated with a fixed seed so that each step below has something visible to do
(the script that builds it, `data-raw/make_example_batch_data.R`, ships with the
package source).

## Read

```{r step-read}
bm <- read_skyline(bxlsx, "skyline_data", bfeat, signal_filter = "LOD")
dim(bm)
```

## Assemble a DatasetExperiment

`assemble_dataset()` aligns the matrix with its feature and sample metadata and
returns a `struct::DatasetExperiment`. **Every step from here on takes and
returns that object**, so the metadata travels with the data and no step has to
be handed a matching sample table again:

```{r step-assemble}
de <- assemble_dataset(bm, bfeat, bmeta)
de
```

## Blank filter

`filter_blanks()` finds the extraction blanks via `sample_meta` and sets any
value below `mean(blanks) * blank_filter` to `NA`. A factor of 3 is a common
choice -- a peak has to be at least three times the blank level to be treated
as real.

```{r step-blanks}
de <- filter_blanks(de, blank_filter = 3, blank_head = "Sample_type",
                    blank_name = "Blank")
sum(is.na(de$data))
```

## Normalise

`normalise_matrix()` divides each sample by a known per-sample amount --
protein, tissue weight, or the volume of biofluid extracted -- so that
concentrations are expressed per unit of starting material:

```{r step-normalise}
de <- normalise_matrix(de, column = "protein_ug")
round(as.data.frame(de$data)[1:3, 1:3], 3)
```

## Concentration adjustment

A calibration curve returns the concentration *in the vial the instrument
sampled*, built at the calibrant's volumes. `adjust_concentration()` puts each
sample back on its own preparation:

$$
\text{conc} \times \frac{\text{cal vol}}{\text{sample vol}}
            \times \frac{\text{sample IS}}{\text{cal IS}}
$$

The first ratio corrects for **reconstitution**: an extract dried down and taken
up in 100 uL is twice as dilute as the calibrant reconstituted in 50 uL, and
that factor has to come back out. The second corrects for any difference in how
much internal standard was spiked into the sample versus the calibrant.

Setting `starting_vol_col` adds a further correction back to the concentration
in the original, undried sample -- the plasma or lavage volume actually taken
into extraction -- so results are reported per unit of biofluid rather than per
vial.

```{r step-adjust}
de <- adjust_concentration(de,
                           sample_vol_col   = "sample_volume_uL",
                           cal_vol_col      = "cal_vol_uL",
                           sample_IS_col    = "sample_IS_vol_uL",
                           cal_IS_col       = "cal_IS_vol_uL",
                           starting_vol_col = "starting_vol_uL")
round(as.data.frame(de$data)[1:3, 1:3], 5)
```

## Impute missing values

`impute_missing()` replaces the remaining `NA`s in the biological samples with a
fraction of that compound's minimum observed value, treating a non-detect as
lying just below the lowest measurement rather than as a zero or a hole.
`scalar = 0.5` is the familiar half-minimum rule; a smaller fraction (0.2) is
more conservative if the panel has compounds sitting near the detection limit.
Blanks are left untouched -- they are diagnostics, not measurements.

Other strategies are available where they matter most, in the PCA
pre-processing (`run_pca(impute = )`), which offers `"none"`, `"min"`,
`"half_min"` and `"frac_min"` alongside missing-value filters that drop
features or samples above a missingness threshold instead of imputing them.
For targeted panels the choice is rarely critical: a compound that is missing
in most samples is usually better dropped than modelled.

```{r step-impute}
de <- impute_missing(de, scalar = 0.5, blank_head = "Sample_type",
                     blank_name = "Blank")
sum(is.na(de$data))
```

## Batch correction

Where an acquisition spans several days or columns, response can shift between
batches. `correct_batch()` scales each batch so its per-feature median matches
the grand median across the reference samples:

```{r step-batch, message = FALSE, warning = FALSE}
batch     <- stats::setNames(de$sample_meta$Chrom_Batch,
                             rownames(as.data.frame(de$data)))
de_uncorr <- de

de <- correct_batch(de, qc_label = "Sample",
                    factor_name = "Sample_type",
                    batch_head  = "Chrom_Batch")
```

Taking the first compound as an example, the two batch medians are pulled
together:

```{r step-batch-medians}
med <- function(x) tapply(as.data.frame(x$data)[[1]], batch, median)

rbind(before = med(de_uncorr),
      after  = med(de))
```

Use pooled QC injections as the reference (`qc_label = "QC"`) whenever the
design has enough of them in every batch; falling back to `"Sample"` assumes
the batches are biologically comparable.

## The whole matrix in one call

`process_dataset()` composes exactly those steps -- read, assemble, then filter
/ normalise / adjust / impute within each acquisition batch, then batch-correct
-- driven by the same metadata sheets:

```{r step-process, message = FALSE, warning = FALSE}
out <- process_dataset(
  bfeat, bmeta,
  xlsx_path        = bxlsx,
  data_source      = "skyline",
  data_tab_names   = "skyline_data",
  signal_filter    = "LOD",
  blank_filter     = 3,
  normalize        = "protein_ug",
  adjust_conc      = TRUE,
  starting_vol_col = "starting_vol_uL",
  replace_MVs      = 0.5,
  batch_correction = TRUE,
  bc_qc_label      = "Sample",
  bc_header        = "Chrom_Batch",
  blank_head       = "Sample_type")

dim(out[[1]]$data)
```

`out[[2]]` holds the features that were dropped along the way with the reason
recorded -- the same table the data-quality report surfaces as "compounds
excluded".

# Outputs

## The processed matrix

The processed dataset is written as `.xlsx` (for anyone downstream who does not
use R) and as `.RDS` (the `DatasetExperiment`, for anyone who does). Rows are
injections, columns are the reported compounds:

```{r out-matrix}
de_full <- res$datasetExperiment
dim(de_full$data)
```

```{r out-matrix-table}
dt(round(as.data.frame(de_full$data)[, 1:6], 2))
```

## Feature metadata: how well was each compound measured?

The per-compound quality metrics are written back into `variable_meta`, matched
to the feature they describe, so annotation and quality live in the same table:

```{r out-cv}
vm <- as.data.frame(de_full$variable_meta)
cv <- vm[, c("Compound", "Enzymatic_pathway", "CV_QC",
             "CV_sample", "CV_sample_vs_QC")]

dt(cv[order(cv$CV_QC), ], rownames = FALSE, page = 10)
```

* **`CV_QC`** is the coefficient of variation across the pooled QC injections.
  Every QC is the same material, so this is **technical** variability -- the
  measurement noise of that compound in that run. It is the primary
  quantification filter: a common rule of thumb keeps compounds under 20-30
  percent CV~QC~ and reports the rest as detected but not reliably quantified.
  It is `NA` when a run contains no QC injections.
* **`CV_sample`** is the same statistic across the biological samples, so it
  carries the **biological** spread *plus* that same technical noise.
* **`CV_sample_vs_QC`** is `CV_sample / CV_QC`, putting the two side by side.
  A ratio comfortably above 1
  means the biological signal is larger than the noise it is measured against
  -- the compound has room to show a real difference. A ratio near 1 means a
  compound is varying no more between animals than it does between replicate
  injections of identical material: whatever passed the CV~QC~ filter, there is
  little biology left to find in it. Sorting the panel by this ratio is a quick
  way to see which analytes are worth interpreting before running any test.

## The reports

The data-quality report is the visual half of that assessment. It covers the
run summary and excluded compounds, the CV table above, per-compound intensity
against injection order (analytical drift), Q-Q plots for normality under
several transforms, and a QC-PCA. Its central question is whether the QC
injections cluster tightly and sit stable across the run -- if they do, the
separation seen between study groups can be trusted.

Drift is read directly off injection order, with the QCs tracked through the
sequence:

```{r qc-drift, message = FALSE, warning = FALSE}
smeta <- as.data.frame(de_full$sample_meta)
dmat  <- as.data.frame(de_full$data)

target <- names(sort(colSums(!is.na(dmat)), decreasing = TRUE))[1]
drift  <- data.frame(order = as.numeric(smeta$Injection_order),
                     type  = smeta$Sample_type,
                     value = dmat[[target]])

ggplot2::ggplot(drift,
                ggplot2::aes(order, value, colour = type)) +
  ggplot2::geom_point(size = 1.8, na.rm = TRUE) +
  ggplot2::scale_colour_viridis_d(end = 0.85) +
  ggplot2::labs(x = "Injection order", y = target, colour = NULL) +
  ggplot2::theme_bw()
```

The QC-PCA asks the same question in the multivariate space the statistics will
use. `run_pca()` takes the dataset, applies the missing-value filtering,
imputation, transform, centring and scaling, and returns the fit together with a
readable trace of what it did:

```{r qc-pca, message = FALSE, warning = FALSE}
pca <- run_pca(de_full, impute = "frac_min", impute_frac = 0.2,
               transform = "log2", feat_na_max = 0.6,
               sample_na_max = 0.6)
pca$steps
```

```{r qc-pca-plot, message = FALSE, warning = FALSE}
scores <- data.frame(
  pca$pr$x[, 1:2],
  type = smeta[rownames(pca$pr$x), "Sample_type"])

ggplot2::ggplot(scores, ggplot2::aes(PC1, PC2, colour = type)) +
  ggplot2::geom_point(size = 2) +
  ggplot2::scale_colour_viridis_d(end = 0.85) +
  ggplot2::labs(colour = NULL) +
  ggplot2::theme_bw()
```

The report draws one PCA per factor named under `qc_pca.color_by`, because the
same scores answer different questions depending on what you colour them by.
Colouring by chromatographic batch is the sharpest of these: if the batches
separate, the variance driving component 1 is analytical, not biological. The
example above is a single-batch run, so use the two-batch workbook -- before and
after `correct_batch()`:

```{r qc-pca-batch, message = FALSE, warning = FALSE, fig.height = 3.4}
pca_by_batch <- function(x, title){
  p <- run_pca(x, impute = "frac_min", impute_frac = 0.2,
               transform = "log2")
  df <- data.frame(p$pr$x[, 1:2],
                   batch = batch[rownames(p$pr$x)])
  ggplot2::ggplot(df, ggplot2::aes(PC1, PC2, colour = batch)) +
    ggplot2::geom_point(size = 2) +
    ggplot2::scale_colour_viridis_d(end = 0.75) +
    ggplot2::labs(title = title, colour = "Chrom_Batch") +
    ggplot2::theme_bw()
}

cowplot::plot_grid(
  pca_by_batch(de_uncorr, "Before batch correction"),
  pca_by_batch(de,        "After batch correction"),
  nrow = 1)
```

In the rendered HTML reports these plots are interactive (hover a point for the
sample behind it) and every table is a searchable, sortable, downloadable
`DT::datatable()` like the ones in this vignette.

# Statistics

`subset_dataset()` slices the dataset by any `sample_meta` condition, which is
how the biological samples are separated from the QCs and blanks before
testing:

```{r stats-subset}
samples <- subset_dataset(de_full,
                          conditions = list(Sample_type = "Sample"))

table(samples$sample_meta$Treatment, samples$sample_meta$Type)
```

`run_stats()` then runs the comparisons, correlations and linear models
described by the `stats_report:` block of the YAML and returns tidy tables. The
method is chosen from the design rather than by the user: two levels get a
t-test and a Wilcoxon test, three or more get ANOVA and Kruskal-Wallis with
Tukey and pairwise-Wilcoxon post-hoc tests, and p-values are FDR-adjusted
within each comparison.

```{r stats-run, message = FALSE, warning = FALSE}
comparison <- list(
  name    = "PBS_vs_HDM_in_WT",
  subset  = list(Type = "WT"),
  compare = list(factor = "Treatment",
                 levels = c("PBS", "HDM")))

params <- list(
  sig_threshold = 0.05,
  comparisons   = list(enabled = TRUE,
                       entries = list(comparison)))

st <- run_stats(samples, params)
```

```{r stats-table}
tab <- st$stats[, c("feature", "method", "mean_g1", "mean_g2",
                    "log2FC", "p_value", "p_adj", "significant")]
tab[, 3:7] <- round(tab[, 3:7], 4)

dt(tab[order(tab$p_value), ], rownames = FALSE, page = 10)
```

Reporting both a parametric and a non-parametric test side by side is
deliberate: targeted panels are frequently run on small groups where normality
cannot be taken on trust, and agreement between the two is itself a piece of
evidence.

The statistics report renders these tables alongside the visual half of the
same analysis: per-compound box and bar plots by study factor, PCA with group
ellipses and annotated loadings, volcano plots per comparison, heatmaps
(all features, top-changing features, and split by compound class),
feature-feature correlation heatmaps, linear-model coefficient tables and ion
ratios for enzyme-activity surrogates. The tables are also appended to a
`_stats.xlsx` next to the processed matrix.

# Driving it from a YAML config

In routine use none of the above is typed out. One YAML file describes the data
source, every processing step and both reports, and one call runs it:

```{r run-yaml, eval = FALSE}
run_MRManalyzeR("path/to/your_config.yml")
```

The bundled reference configuration is a commented template to copy and edit:

```{r config-path}
system.file("extdata", "example_config.yml", package = "MRManalyzeR")
```

Its `PeakMatrixProcessing:` block is a one-to-one map of the steps above -- each
key switches a step on or names the metadata column it should use:

```yaml
PeakMatrixProcessing:
  execute: True                  # False = skip, reuse existing RDS

  name_col: Name                 # sample id (matches the matrix rows)
  include_col: Include           # sample include-filter column
  include_value: "YES"

  tl_data:                       # TargetLynx source (or skyline_data:)
    enabled: True
    data_tab_names: [lcms_data]
    tl_headers: ["ID", "Name", "Area", "ng/mL", "Response", "S/N"]
    snr: 3                       # S/N threshold | False

  blank_filter: 3                # n x blank mean | False
  blank_head: 'Sample_type'
  blank_name: 'Blank'

  normalize: protein_ug          # False | sample_metadata column to divide by

  adjust_conc:
    enabled: True
    starting_vol_col: starting_vol_uL   # False to skip the reconstitution step
    sample_vol_col:   sample_volume_uL
    cal_vol_col:      cal_vol_uL
    sample_IS_col:    sample_IS_vol_uL
    cal_IS_col:       cal_IS_vol_uL

  replace_MVs: 0.5               # False | scalar x per-feature min
  batch_correction: True
  bc_qc_label: "QC"
  bc_factor_name: "Sample_type"
  bc_header: "Chrom_Batch"
```

The `data_quality_report:` and `stats_report:` blocks configure the two reports
in the same declarative way -- which column identifies QCs, which factors to
compare, which plots to draw. See the bundled file for the fully commented set.

Outputs are written next to your data: the processed matrix as `.xlsx` and
`.RDS`, a `_stats.xlsx` of comparisons, correlations and linear models, and the
two self-contained HTML reports. `run_MRManalyzeR_combine()` is the equivalent
entry point for merging several acquisition panels into a single analysis.

# References {#references}

Kolmert J. *et al.* (2018). Prominent release of lipoxygenase generated
mediators in a murine house dust mite-induced asthma model.
*Prostaglandins & Other Lipid Mediators*.
doi:[10.1016/j.prostaglandins.2018.05.005](https://doi.org/10.1016/j.prostaglandins.2018.05.005)
-- source of the bundled example data.

Lloyd G.R., Jankevics A., Weber R.J.M. (2020). struct: an R/Bioconductor-based
framework for standardized metabolomics data analysis. *Bioinformatics*.
doi:[10.1093/bioinformatics/btaa227](https://doi.org/10.1093/bioinformatics/btaa227)
-- the `DatasetExperiment` container used throughout.

# Session info

```{r session-info}
sessionInfo()
```
