---
title: "Processing targeted LC-MS lipidomics and metabolomics with MRManalyzeR"
author: "Matthew J. Smith"
date: "`r Sys.Date()`"
package: MRManalyzeR
output:
  BiocStyle::html_document:
    toc_float: true
vignette: >
  %\VignetteIndexEntry{Processing targeted LC-MS lipidomics and metabolomics with MRManalyzeR}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup-knitr, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
```

# Overview

MRManalyzeR processes and analyses targeted lipidomics and metabolomics data
exported from Waters TargetLynx (or Skyline). The processing pipeline is a
sequence of small, composable step functions -- read the workbook, filter,
adjust, impute, assemble, batch-correct -- followed by the statistical
analyses. The single entry point `run_MRManalyzeR()` drives the whole thing
from one YAML config, but every step is a public function you can call on its
own. This vignette walks through each step, then shows the one-YAML shortcut.

```{r load}
library(MRManalyzeR)
```

# The pipeline, step by step

## 1. Read the workbook

`read_targetlynx()` (TargetLynx) and `read_skyline()` (Skyline) both return a
wide *sample x compound* matrix, optionally masking values below a
signal-to-noise (`snr`) or LOD/LOQ threshold. Here we read the bundled
oxylipin example workbook:

```{r read}
xlsx <- system.file("extdata", "example_data.xlsx", package = "MRManalyzeR")
m <- read_targetlynx(xlsx, datatype = "Area", snr = 3)
m[1:4, 1:4]
```

## 2. Blank filter

`filter_blanks()` sets any value below `mean(blank) * blank_filter` to `NA`:

```{r filter-blanks}
demo <- data.frame(PGE2 = c(100, 90, 5), PGD2 = c(50, 60, 1),
                   row.names = c("S1", "S2", "Blank1"))
filter_blanks(demo, blank_samples = "Blank1", blank_filter = 3)
```

## 3. Normalise

`normalise_matrix()` divides each sample by a known per-sample amount (protein,
tissue weight, ...), matched by name:

```{r normalise}
m2 <- data.frame(PGE2 = c(10, 20), PGD2 = c(30, 40), row.names = c("S1", "S2"))
sm <- data.frame(Name = c("S1", "S2"), protein_ug = c(2, 4))
normalise_matrix(m2, sm, column = "protein_ug")
```

## 4. Concentration adjustment

`adjust_concentration()` applies the vial-level calibration correction
(volumes + internal standard) and, optionally, a starting-volume correction:

```{r adjust-conc}
sv <- data.frame(Name = c("S1", "S2"),
                 sample_volume_uL = c(100, 100), cal_vol_uL = c(100, 100),
                 sample_IS_vol_uL = c(10, 10),   cal_IS_vol_uL = c(10, 10))
adjust_concentration(m2, sv)
```

## 5. Impute missing values

`impute_missing()` replaces `NA`s with a fraction of the per-feature minimum:

```{r impute}
mv <- data.frame(PGE2 = c(10, 5, NA), PGD2 = c(2, NA, 4),
                 row.names = c("S1", "S2", "S3"))
impute_missing(mv, blank_samples = character(0), scalar = 0.5)
```

## 6. Assemble a DatasetExperiment

`assemble_dataset()` aligns the matrix with the feature and sample metadata
into a `struct::DatasetExperiment`, the object the rest of the package works on:

```{r assemble}
fm <- data.frame(Compound = c("PGE2", "PGD2"), Report = "YES",
                 row.names = c("PGE2", "PGD2"))
smeta <- data.frame(Name = c("S1", "S2"), Group = c("ctrl", "trt"),
                    row.names = c("S1", "S2"))
de <- assemble_dataset(m2, fm, smeta)
de
```

## 7. Batch correction

`correct_batch()` scales each batch so its per-feature median matches the grand
median across the reference samples:

```{r batch, message = FALSE, warning = FALSE}
bm  <- data.frame(PGE2 = c(10, 12, 20, 24), PGD2 = c(5, 6, 10, 12),
                  row.names = c("S1", "S2", "S3", "S4"))
bfm <- data.frame(Compound = c("PGE2", "PGD2"), row.names = c("PGE2", "PGD2"))
bsm <- data.frame(Sample_type = "Sample", Chrom_Batch = c("b1", "b1", "b2", "b2"),
                  row.names = c("S1", "S2", "S3", "S4"))
bde <- struct::DatasetExperiment(data = bm, sample_meta = bsm, variable_meta = bfm)
correct_batch(bde, qc_label = "Sample", factor_name = "Sample_type",
              batch_head = "Chrom_Batch")
```

## 8. Statistics

`run_stats()` runs the comparisons, correlations and linear models described by
a `stats_report:` parameter list and returns tidy tables:

```{r stats, message = FALSE, warning = FALSE}
sde <- struct::DatasetExperiment(
  data = data.frame(PGE2 = c(10, 12, 11, 20, 24, 22),
                    PGD2 = c(5, 6, 5, 11, 13, 12), row.names = paste0("S", 1:6)),
  sample_meta = data.frame(Sample_ID = paste0("S", 1:6),
                           Group = rep(c("ctrl", "trt"), each = 3),
                           row.names = paste0("S", 1:6)),
  variable_meta = data.frame(Compound = c("PGE2", "PGD2"),
                             row.names = c("PGE2", "PGD2")))
params <- list(comparisons = list(enabled = TRUE, entries = list(
  list(name = "ctrl_vs_trt",
       compare = list(factor = "Group", levels = c("ctrl", "trt"))))))
run_stats(sde, params)$stats
```

## 9. PCA

`run_pca()` pre-processes a matrix (missing-value handling, transform, scaling)
and runs `prcomp()`, returning the fit plus a trace of the steps applied:

```{r pca}
X <- matrix(abs(rnorm(60, 100, 20)), nrow = 12,
            dimnames = list(paste0("S", 1:12), paste0("F", 1:5)))
res <- run_pca(X, transform = "log2")
res$steps
```

# Composing the pipeline

`process_dataset()` wires the read + per-batch (filter / normalise /
concentration / impute) + assemble + batch-correct steps together, driven by the
feature and sample metadata:

```{r process, eval = FALSE}
fdata <- openxlsx::read.xlsx(xlsx, sheet = "feature_metadata")
meta  <- openxlsx::read.xlsx(xlsx, sheet = "sample_metadata")
out <- process_dataset(fdata, meta, xlsx_path = xlsx,
                       data_source = "targetlynx", datatype = "Area",
                       snr = 3, blank_filter = FALSE, bc_header = "Chrom_Batch")
out[[1]]   # the processed DatasetExperiment
```

`run_example()` runs that whole pipeline on the bundled dataset and writes the
outputs. With `render = FALSE` it builds the matrix without rendering the (slower)
HTML reports:

```{r run-example, message = FALSE}
res <- run_example(render = FALSE, open = FALSE)
list.files(attr(res, "results_dir"))
```

# Driving it from a YAML config

In practice the entire run -- data source, every processing step, the QC and
statistics reports -- is described by a single YAML file, and you call one
function. The bundled reference config is:

```{r config-path}
system.file("extdata", "example_config.yml", package = "MRManalyzeR")
```

Its top-level blocks are `paths`, `PeakMatrixProcessing` (data source + each
processing step), `data_quality_report`, and `stats_report`. Copy it, edit it
for your study, and run:

```{r run-yaml, eval = FALSE}
run_MRManalyzeR("path/to/your_config.yml")
```

Outputs (the processed matrix as `.xlsx` + `.RDS`, a `_stats.xlsx`, and two
self-contained HTML reports) are written next to your data.
`run_MRManalyzeR_combine()` is the equivalent entry point for merging several
acquisition panels into one analysis.

# Session info

```{r session-info}
sessionInfo()
```
