---
title: "MRManalyzeR — Data Quality Report"
output:
  html_document:
    toc: true
    toc_float: true
    self_contained: true
    code_folding: hide
    css: !expr system.file("rmd","IMP_ReportCSS.css", package="MRManalyzeR")
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo        = FALSE,
  warning     = FALSE,
  message     = FALSE,
  fig.align   = "center",
  fig.retina  = 3
)

library(dplyr)
library(tidyr)
library(ggplot2)
```

```{r unpack-params}
# `report_pars` is injected by run_MRManalyzeR().
# Falls back to legacy uva_pars / mva_pars if running outside that wrapper.
if(!exists("report_pars") || is.null(report_pars))
  report_pars = if(exists("uva_pars")) uva_pars else mva_pars

`%||%` = function(a, b) if(is.null(a)) b else a

# Universal params
summary_head           = report_pars$summary_head
qc_real                = report_pars$qc_real
qc_label               = report_pars$qc_label
qc_head                = report_pars$qc_head
qc_head_ID             = report_pars$qc_head_ID
global_boxpot_factor   = report_pars$global_boxpot_factor
intensity_plot_levels  = report_pars$intensity_plot_levels
intensity_plot_head    = report_pars$intensity_plot_head
intensity_plot_head_ID = report_pars$intensity_plot_head_ID
yaxis_quant            = report_pars$yaxis_quant %||% 0.95

units      = pmp_params$units %||% pmp_params$datatype %||%
             project_paths$datatype %||% ""
bc_header  = pmp_params$bc_header

# Performance / layout knobs
plot_intensity_per_compound = isTRUE(report_pars$plot_intensity_per_compound %||% TRUE)
plot_qq_per_compound        = isTRUE(report_pars$plot_qq_per_compound        %||% TRUE)
# qq_transform may be a single value or a list — list form creates one tab per transform.
qq_transforms               = report_pars$qq_transform %||% c("none", "log2")
if(length(qq_transforms) == 1) qq_transforms = as.character(qq_transforms)
fig_scale                   = as.numeric(report_pars$fig_scale %||% 1)

# QC PCA config
qc_pca_include = isTRUE(report_pars$qc_pca$include  %||% FALSE)
qc_pca_color   = report_pars$qc_pca$color_by        %||% character(0)
# Default INCLUDES the QC sample type so the QC PCA shows QCs alongside
# samples (drift / batch / clustering checks). Override via
# stats_report.qc_pca.levels: in YAML.
qc_pca_levels  = report_pars$qc_pca$levels          %||%
                   unique(c("Sample", qc_label))
qc_pca_head    = report_pars$qc_pca$head            %||% "Sample_type"
qc_pca_label   = report_pars$qc_pca$label           %||% "none"
qc_pca_label_factor = report_pars$qc_pca$label_factor %||% "Sample_ID"
qc_pca_ellipse = report_pars$qc_pca$ellipse         %||% "none"

# Preprocessing for QC PCA (with sensible defaults)
qc_pca_pp = report_pars$qc_pca$preprocess %||% list()
qc_pca_impute        = qc_pca_pp$impute        %||% "min"
qc_pca_impute_frac   = as.numeric(qc_pca_pp$impute_frac %||% 0.5)
qc_pca_transform     = qc_pca_pp$transform     %||% "log2"
qc_pca_center        = isTRUE(qc_pca_pp$center %||% TRUE)
qc_pca_scale         = isTRUE(qc_pca_pp$scale  %||% TRUE)
qc_pca_feat_na_max   = as.numeric(qc_pca_pp$feat_na_max   %||% 1)
qc_pca_sample_na_max = as.numeric(qc_pca_pp$sample_na_max %||% 1)

# Interactive plotting (plotly hover tooltips). Default ON. Set
# `data_quality_report.interactive_plots: False` in YAML to fall back to png.
interactive_plots = isTRUE(report_pars$interactive_plots %||% TRUE) &&
                    requireNamespace("plotly", quietly = TRUE)

.show_plot = function(p, lock_aspect = FALSE, legend_below = FALSE,
                      max_width = NULL, pixel_height = NULL){
  if(interactive_plots){
    w = plotly::ggplotly(p, tooltip = "text", height = pixel_height)
    if(isTRUE(lock_aspect))
      w = plotly::layout(w, yaxis = list(scaleanchor = "x", scaleratio = 1))
    if(isTRUE(legend_below)){
      w = plotly::layout(w,
                         margin = list(b = 170, t = 40),
                         legend = list(orientation = "h",
                                       x = 0.5, xanchor = "center",
                                       y = -0.40, yanchor = "top"))
      # Continuous colorbars (Injection_order, numeric Chrom_Batch) live on
      # each trace's `marker.colorbar`. Walk the figure data and only modify
      # traces that already carry a colorbar — discrete-color traces don't,
      # so they remain untouched (avoids the phantom-colorbar bug).
      for(.i in seq_along(w$x$data)){
        .mk = w$x$data[[.i]]$marker
        if(isTRUE(.mk$showscale) || !is.null(.mk$colorscale)){
          if(is.null(w$x$data[[.i]]$marker$colorbar))
            w$x$data[[.i]]$marker$colorbar = list()
          w$x$data[[.i]]$marker$colorbar$orientation = "h"
          w$x$data[[.i]]$marker$colorbar$x          = 0.5
          w$x$data[[.i]]$marker$colorbar$xanchor    = "center"
          w$x$data[[.i]]$marker$colorbar$y          = -0.30
          w$x$data[[.i]]$marker$colorbar$yanchor    = "top"
          w$x$data[[.i]]$marker$colorbar$len        = 0.7
          w$x$data[[.i]]$marker$colorbar$thickness  = 14
        }
      }
    }
    w = plotly::config(w, displaylogo = FALSE,
                       modeBarButtonsToRemove = c("lasso2d", "select2d"))
    tags = htmltools::as.tags(w, standalone = FALSE)
    deps = htmltools::findDependencies(tags)
    if(length(deps)) knitr::knit_meta_add(list(deps))
    html_str = as.character(tags)
    if(!is.null(max_width))
      html_str = sprintf('<div style="max-width:%s;margin:0 auto;">%s</div>',
                         max_width, html_str)
    cat(html_str)
  } else {
    print(p)
  }
}

# Pre-built theme reused across the per-compound loop
compound_theme = ggplot2::theme_classic() +
  ggplot2::theme(
    legend.position = "bottom",
    axis.title      = ggplot2::element_text(face = "bold", color = "black", size = 12),
    axis.text       = ggplot2::element_text(face = "bold", color = "black", size = 8)
  )
```

# 1. Aim

Inspect data quality for acquired compounds. Check QCs, drift over injection
order, and confounding factors (batch, sample type) via PCA.

# 2. Data import and summary

```{r import}
importData      = combined_datamatrices
metadata        = importData$sample_meta
data_matrix     = importData$data
compoundInfo    = importData$variable_meta
compound_names  = compoundInfo$Compound

DT::datatable(
  metadata %>%
    dplyr::group_by(.data[[summary_head]]) %>%
    dplyr::summarize(`Number of samples` = dplyr::n(), .groups = "drop")
)
```

# 3. Quality CVs

```{r cvs}
# Per-feature CV in QC vs samples (when QC samples present)
qc_idx     = which(metadata[[qc_head]] == qc_label)
sample_idx = setdiff(seq_len(nrow(metadata)), qc_idx)

cv_fn = function(x) sd(x, na.rm = TRUE) / mean(x, na.rm = TRUE) * 100

cv_table = data.frame(
  Compound  = colnames(data_matrix),
  CV_QC     = if(length(qc_idx) >= 2) round(vapply(data_matrix, function(x) cv_fn(x[qc_idx]),     numeric(1)), 2) else NA_real_,
  CV_sample = if(length(sample_idx) >= 2) round(vapply(data_matrix, function(x) cv_fn(x[sample_idx]), numeric(1)), 2) else NA_real_
)

DT::datatable(cv_table, filter = "top",
              options = list(pageLength = 25, scrollX = TRUE))
```

# 4. Per-compound intensity vs injection order

`r if(!plot_intensity_per_compound) "_Skipped — set_ \x60plot_intensity_per_compound: True\x60 _in YAML to enable._"`

```{r intensity-dropdown, results="asis", eval=plot_intensity_per_compound, fig.width=10*fig_scale, fig.height=5*fig_scale}
# Subset to requested sample types
keep = metadata[[intensity_plot_head]] %in% intensity_plot_levels
inj_data = data_matrix[keep, , drop = FALSE]
inj_data$injection_order = if("Injection_order" %in% colnames(metadata))
  metadata$Injection_order[keep] else seq_len(nrow(inj_data))
# Colour-by factor; coalesce NA values to the Sample_type label so e.g.
# QC injections (which usually have no Treatment) show up as "QC" rather
# than legend group "NA" when colouring by Treatment.
inj_data$type = as.character(metadata[[intensity_plot_head_ID]][keep])
.fallback     = as.character(metadata[[intensity_plot_head]][keep])
inj_data$type[is.na(inj_data$type)] = .fallback[is.na(inj_data$type)]
inj_data$.sample_id = if("Sample_ID" %in% colnames(metadata))
  as.character(metadata$Sample_ID[keep]) else rownames(inj_data)

make_safe_id = function(x) gsub("[^A-Za-z0-9_\\-]", "_", x)
safe_ids = vapply(compound_names, make_safe_id, character(1))
# Alphabetise dropdown options for easier scanning (T3.5-7); plot blocks
# below still iterate in import order.
.alpha_idx        = order(compound_names)
.compound_alpha   = compound_names[.alpha_idx]
.safe_ids_alpha   = safe_ids[.alpha_idx]

cat('
<label for="compound_select" style="font-weight:600;">Select compound:</label>
<select id="compound_select" style="margin: 0 0 12px 8px; padding: 4px 6px;"
        onchange="showCompoundPlot(this.value)">
')
for(i in seq_along(.compound_alpha)){
  sel = if(i == 1) " selected" else ""
  cat(sprintf('<option value="%s"%s>%s</option>\n',
              .safe_ids_alpha[i], sel, .compound_alpha[i]))
}
cat('</select>
<script>
function showCompoundPlot(id){
  var plots = document.querySelectorAll(".compound_plot_block");
  for (var i=0; i<plots.length; i++) { plots[i].style.display = "none"; }
  var el = document.getElementById("plot_" + id);
  if (el) {
    el.style.display = "block";
    if (window.Plotly){
      var pls = el.querySelectorAll(".plotly-graph-div");
      for (var j=0; j<pls.length; j++) Plotly.Plots.resize(pls[j]);
    }
  }
}
</script>
')

for(i in seq_along(compound_names)){
  compound = compound_names[i]
  cid      = safe_ids[i]
  axis_limit = quantile(inj_data[[compound]], prob = yaxis_quant, na.rm = TRUE) * 1.3
  inj_data$.tooltip = sprintf("Sample: %s<br>%s: %s<br>Order: %s<br>%s: %.3g",
                              inj_data$.sample_id, intensity_plot_head_ID,
                              as.character(inj_data$type),
                              as.character(inj_data$injection_order),
                              compound, inj_data[[compound]])

  p = ggplot2::ggplot(inj_data,
                      ggplot2::aes(injection_order, .data[[compound]])) +
        ggplot2::geom_point(ggplot2::aes(color = type, text = .tooltip),
                            size = 3, alpha = 0.6) +
        ggplot2::scale_color_brewer(palette = "Set1", na.value = "grey60") +
        ggplot2::coord_cartesian(ylim = c(0, axis_limit)) +
        ggplot2::labs(x = "Injection order", y = units, title = compound) +
        compound_theme

  # Show the alphabetically-first compound by default (matches dropdown selection)
  display = if(safe_ids[i] == .safe_ids_alpha[1]) "block" else "none"
  cat(sprintf('<div class="compound_plot_block" id="plot_%s" style="display:%s;">\n',
              cid, display))
  .show_plot(p, legend_below = TRUE)
  cat("</div>\n")
}
```

# 5. Q-Q plot per compound (normality check) {.tabset}

`r if(!plot_qq_per_compound) "_Skipped — set_ \x60plot_qq_per_compound: True\x60 _in YAML to enable._"`

`r if(plot_qq_per_compound) sprintf("Sample-level normality assessment per compound. Filter applied to drop blanks/QCs/calibration: \x60%s\x60 _∈ {%s}_ → _%d samples kept (of %d in metadata). One tab per transform listed in_ \x60qq_transform\x60_:_ %s_. Caption shows Shapiro-Wilk W and p (3 ≤ n ≤ 5000)._", intensity_plot_head, paste(intensity_plot_levels, collapse=", "), sum(metadata[[intensity_plot_head]] %in% intensity_plot_levels), nrow(metadata), paste(qq_transforms, collapse=", "))`

```{r qq-dropdown, results="asis", eval=plot_qq_per_compound, fig.width=6*fig_scale, fig.height=5*fig_scale}
keep_qq = metadata[[intensity_plot_head]] %in% intensity_plot_levels
qq_data = data_matrix[keep_qq, , drop = FALSE]

.qq_apply = function(v, mode){
  if(mode == "log2") return(suppressWarnings(log2(v + 1)))
  if(mode == "sqrt") return(suppressWarnings(sqrt(pmax(v, 0))))
  v
}

make_safe_id_qq = function(x) gsub("[^A-Za-z0-9_\\-]", "_", x)
# Iterate compounds in import order for the plot blocks; the dropdown lists
# them alphabetically (T3.5-7) for easier scanning.
safe_ids_qq      = vapply(compound_names, make_safe_id_qq, character(1))
alpha_idx        = order(compound_names)
compound_alpha   = compound_names[alpha_idx]
safe_ids_alpha   = safe_ids_qq[alpha_idx]

for(tr in qq_transforms){
  tr_safe = make_safe_id_qq(tr)
  cat(sprintf("\n## %s {-}\n\n", tr))

  cat(sprintf('
<label for="qq_select_%s" style="font-weight:600;">Select compound:</label>
<select id="qq_select_%s" style="margin: 0 0 12px 8px; padding: 4px 6px;"
        onchange="showQQPlot_%s(this.value)">
', tr_safe, tr_safe, tr_safe))
  for(i in seq_along(compound_alpha)){
    sel = if(i == 1) " selected" else ""
    cat(sprintf('<option value="%s"%s>%s</option>\n',
                safe_ids_alpha[i], sel, compound_alpha[i]))
  }
  cat(sprintf('</select>
<script>
function showQQPlot_%s(id){
  var plots = document.querySelectorAll(".qq_plot_block_%s");
  for (var i=0; i<plots.length; i++) { plots[i].style.display = "none"; }
  var el = document.getElementById("qq_%s_" + id);
  if (el) el.style.display = "block";
}
</script>
', tr_safe, tr_safe, tr_safe))

  for(i in seq_along(compound_names)){
    compound = compound_names[i]
    cid      = safe_ids_qq[i]
    v_raw    = qq_data[[compound]]
    v        = .qq_apply(v_raw[!is.na(v_raw)], tr)

    sw_lbl = ""
    if(length(v) >= 3 && length(v) <= 5000){
      sw = tryCatch(stats::shapiro.test(v), error = function(e) NULL)
      if(!is.null(sw))
        sw_lbl = sprintf("  Shapiro-Wilk W = %.3f, p = %s",
                         sw$statistic,
                         format.pval(sw$p.value, digits = 3, eps = 1e-4))
    }

    if(length(v) < 3){
      p = ggplot2::ggplot() + ggplot2::theme_void() +
            ggplot2::labs(title = sprintf("%s — too few non-NA values (n=%d)",
                                          compound, length(v)))
    } else {
      df_qq = data.frame(value = v)
      y_lab = switch(tr,
                     log2 = sprintf("log2(%s + 1) sample quantiles", units),
                     sqrt = sprintf("sqrt(%s) sample quantiles", units),
                     sprintf("%s sample quantiles", units))

      p = ggplot2::ggplot(df_qq, ggplot2::aes(sample = value)) +
            ggplot2::stat_qq(size = 2, alpha = 0.7) +
            ggplot2::stat_qq_line(color = "#d62728", linewidth = 0.6) +
            ggplot2::labs(x = "Theoretical normal quantiles",
                          y = y_lab,
                          title = sprintf("%s   (n = %d)%s",
                                          compound, length(v), sw_lbl)) +
            compound_theme
    }

    # Each compound's plot div uses the dropdown safe_id (alpha-list value);
    # its leading style remembers which compound is the alpha-first one.
    display = if(safe_ids_qq[i] == safe_ids_alpha[1]) "block" else "none"
    cat(sprintf('<div class="qq_plot_block_%s" id="qq_%s_%s" style="display:%s;">\n',
                tr_safe, tr_safe, cid, display))
    print(p)
    cat("</div>\n")
  }
}
```

# 6. QC PCA (drift / batch / confounder inspection)

`r if(!qc_pca_include) "_Skipped — set_ \x60qc_pca.include: True\x60 _in YAML to enable._"`

`r if(qc_pca_include) sprintf("PCA computed once on the filtered subset; one plot per coloring listed in_ \x60qc_pca.color_by\x60_:_ %s.", paste(qc_pca_color, collapse=", "))`

```{r qc-pca, results="asis", eval=qc_pca_include && length(qc_pca_color) > 0, fig.width=7*fig_scale, fig.height=6*fig_scale}
# Subset to requested sample types
keep = metadata[[qc_pca_head]] %in% qc_pca_levels
X = as.matrix(data_matrix[keep, , drop = FALSE])
smeta_pca = metadata[keep, , drop = FALSE]

# Ensure shared rownames on X and smeta_pca so we can re-align after the
# PCA pipeline drops samples via `sample_na_max`.
.qc_rn = if(!is.null(rownames(X)) && !anyDuplicated(rownames(X))) {
  rownames(X)
} else {
  paste0("qc_pca_", seq_len(nrow(X)))
}
rownames(X)         = .qc_rn
rownames(smeta_pca) = .qc_rn

res_pca = run_pca_pipeline(X,
                           impute        = qc_pca_impute,
                           impute_frac   = qc_pca_impute_frac,
                           transform     = qc_pca_transform,
                           center        = qc_pca_center,
                           scale         = qc_pca_scale,
                           feat_na_max   = qc_pca_feat_na_max,
                           sample_na_max = qc_pca_sample_na_max)

# Realign metadata to the samples that survived sample_na_max filtering.
if(!is.null(res_pca$pr))
  smeta_pca = smeta_pca[rownames(res_pca$pr$x), , drop = FALSE]

# Show what was actually applied
cat("**Preprocessing pipeline applied:**\n\n")
for(s in res_pca$steps) cat(sprintf("- %s\n", s))
cat(sprintf("\n_Final matrix: %d samples × %d features._\n\n",
            res_pca$n_samples, res_pca$n_features))

if(is.null(res_pca$pr)){
  cat(sprintf("\n_PCA could not be fit on cleaned matrix (n_samples=%d, n_features=%d)._\n",
              res_pca$n_samples, res_pca$n_features))
} else {
  pr     = res_pca$pr
  scores = as.data.frame(pr$x[, 1:min(2, ncol(pr$x))])
  if(ncol(scores) < 2) scores$PC2 = 0
  vexp = (pr$sdev^2) / sum(pr$sdev^2) * 100
  vx = sprintf("PC1 (%.1f%%)", vexp[1])
  vy = sprintf("PC2 (%.1f%%)", if(length(vexp) >= 2) vexp[2] else 0)

  # One tab per coloring listed in qc_pca.color_by
  cat("\n## QC PCA scatter plots {.tabset .tabset-fade}\n\n")

  for(col_name in qc_pca_color){
    cat(sprintf("\n### Coloured by %s {-}\n\n", col_name))

    if(!col_name %in% colnames(smeta_pca)){
      cat(sprintf("_Column `%s` not in sample_meta — skipped._\n\n", col_name))
      next
    }

    sc = scores
    sc$.color = smeta_pca[[col_name]]
    sc$.label = smeta_pca[[qc_pca_label_factor]] %||% rownames(sc)

    is_num = is.numeric(sc$.color) ||
             (suppressWarnings(!any(is.na(as.numeric(sc$.color)))) &&
              col_name %in% c("Injection_order", "Run_order"))
    if(is_num) sc$.color = as.numeric(sc$.color)

    sc$.tooltip = sprintf("%s: %s<br>%s: %s<br>PC1: %.3g<br>PC2: %.3g",
                          qc_pca_label_factor, sc$.label,
                          col_name, as.character(sc$.color),
                          sc$PC1, sc$PC2)

    p = ggplot2::ggplot(sc, ggplot2::aes(PC1, PC2, text = .tooltip)) +
          ggplot2::geom_point(ggplot2::aes(color = .color), size = 3, alpha = 0.85) +
          ggplot2::labs(x = vx, y = vy, color = col_name) +
          ggplot2::theme_classic(base_size = 13) +
          ggplot2::theme(legend.position = "bottom")

    # Continuous (Injection_order, Run_order) → viridis_c;
    # categorical (Sample_type, Chrom_Batch as factor) → Set1 brewer.
    p = if(is_num) p + ggplot2::scale_color_viridis_c()
        else       p + ggplot2::scale_color_brewer(palette = "Set1",
                                                   na.value = "grey60")

    if(qc_pca_ellipse == "group" && !is_num && length(unique(sc$.color)) > 1)
      p = p + ggplot2::stat_ellipse(ggplot2::aes(group = .color),
                                    type = "norm", level = 0.95, linewidth = 0.4)

    if(qc_pca_label == "all"){
      p = p + ggplot2::geom_text(ggplot2::aes(label = .label),
                                 size = 2.5, vjust = -1, show.legend = FALSE)
    } else if(qc_pca_label == "outliers"){
      zx = scale(sc$PC1)[, 1]; zy = scale(sc$PC2)[, 1]
      out_idx = abs(zx) > 2.5 | abs(zy) > 2.5
      if(any(out_idx))
        p = p + ggplot2::geom_text(data = sc[out_idx, ],
                                   ggplot2::aes(label = .label),
                                   size = 2.5, vjust = -1, show.legend = FALSE)
    }

    .show_plot(p, lock_aspect = TRUE, legend_below = TRUE,
               max_width = "650px", pixel_height = 650)
    cat("\n\n")
  }
}
```

# 7. Final considerations {.tabset}

## Compounds included {-}

```{r}
DT::datatable(
  combined_datamatrices$variable_meta %>% dplyr::select(Compound, Comment),
  caption = "Summary of compounds included"
)
```

## Compounds excluded {-}

```{r}
if(!is.null(removed_features) && nrow(removed_features) > 0){
  DT::datatable(
    removed_features %>% dplyr::select(Compound, Comment),
    caption = "Summary of compounds excluded"
  )
} else {
  cat("_None recorded (typical when PeakMatrixProcessing.execute was False)._")
}
```

# 8. Session info

```{r session}
tryCatch(sessionInfo(), error = function(e){
  cat("sessionInfo() failed (likely a malformed DESCRIPTION in an installed package):\n")
  cat(conditionMessage(e), "\n\n")
  cat("R version:", R.version$version.string, "\n")
  cat("Platform :", R.version$platform, "\n")
  cat("Loaded packages:\n")
  cat(paste(" ", sort(loadedNamespaces()), collapse = "\n"), "\n")
})
```
