---
title: "MRManalyzeR — Statistics Report"
output:
  html_document:
    toc: true
    toc_float: true
    self_contained: true
    code_folding: hide
    css: !expr system.file("rmd","IMP_ReportCSS.css", package="MRManalyzeR")
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo        = FALSE,
  warning     = FALSE,
  message     = FALSE,
  fig.align   = "center",
  fig.retina  = 3
)

library(dplyr)
library(tidyr)
library(ggplot2)
```

```{r unpack-params}
if(!exists("report_pars") || is.null(report_pars))
  report_pars = if(exists("mva_pars")) mva_pars else uva_pars

`%||%` = function(a, b) if(is.null(a)) b else a

# Stats blocks
comparisons   = report_pars$comparisons   %||% list()
correlations  = report_pars$correlations  %||% list()
linear_models = report_pars$linear_models %||% list()

p_adjust      = report_pars$p_adjust      %||% "BH"
sig_threshold = report_pars$sig_threshold %||% 0.05

# Plot config
fig_scale       = as.numeric(report_pars$fig_scale %||% 1)
boxplot_faceted = isTRUE(report_pars$boxplot$faceted %||% TRUE)
pca_include     = isTRUE(report_pars$pca$include   %||% TRUE)
pca_label       = report_pars$pca$label            %||% "none"
pca_label_factor = report_pars$pca$label_factor    %||% "Sample_ID"
pca_ellipse     = report_pars$pca$ellipse          %||% "group"
pca_ellipse_conf = as.numeric(report_pars$pca$ellipse_confidence %||% 0.95)

# PCA preprocessing (with sensible defaults)
pca_pp = report_pars$pca$preprocess %||% list()
pca_impute        = pca_pp$impute      %||% "min"
pca_impute_frac   = as.numeric(pca_pp$impute_frac %||% 0.5)
pca_transform     = pca_pp$transform   %||% "log2"
pca_center        = isTRUE(pca_pp$center %||% TRUE)
pca_scale         = isTRUE(pca_pp$scale  %||% TRUE)

# Component pair to plot (default PC1 vs PC2). Length-2 vector of integers.
pca_components    = report_pars$pca$components %||% c(1, 2)
pca_pc_x          = as.integer(pca_components[1])
pca_pc_y          = as.integer(pca_components[length(pca_components)])

# Optional loadings plot below scores; coloured by a feature_metadata column.
pca_loadings_show = isTRUE(report_pars$pca$loadings$include %||% TRUE)
pca_loadings_color_by = report_pars$pca$loadings$color_by %||% NULL

# Optional extra subset for the GLOBAL PCA tab (multi-key, AND'd together).
# This is on top of the biological_filter, e.g. pca.subset: { Lobe: Left }
# to inspect only the left-lobe samples in the global PCA.
pca_global_subset = report_pars$pca$subset %||% NULL

volcano_include   = isTRUE(report_pars$volcano$include %||% TRUE)
volcano_top_n     = as.integer(report_pars$volcano$top_n_label %||% 10)
# Volcano y-axis: use raw or BH-adjusted p (default = unadjusted, per request)
volcano_use_adj   = isTRUE(report_pars$volcano$use_adjusted %||% FALSE)

# Heatmap (z-score, samples x features)
heatmap_include            = isTRUE(report_pars$heatmap$include %||% TRUE)
heatmap_color_samples_by   = report_pars$heatmap$color_samples_by  %||%
                             (report_pars$global_summary_factor %||% "Sample_type")
heatmap_group_features_by  = report_pars$heatmap$group_features_by %||% NULL
heatmap_transform          = report_pars$heatmap$transform %||% "log2"
heatmap_scale              = report_pars$heatmap$scale     %||% "z"
heatmap_cluster_rows       = isTRUE(report_pars$heatmap$cluster_rows %||% FALSE)
heatmap_cluster_within     = isTRUE(report_pars$heatmap$cluster_within_groups %||% TRUE)

# Biological-sample filter (drops blanks/QCs from global summary + global PCA).
# Default: keep only Sample_type == "Sample". Set head: ~ to disable.
bio_filter        = report_pars$biological_filter %||%
                    list(head = "Sample_type", levels = c("Sample"))
bio_filter_head   = bio_filter$head
bio_filter_levels = bio_filter$levels

# General per-compound boxplot (independent of comparisons)
boxplot_enabled = isTRUE(report_pars$boxplot$enabled %||% FALSE)
boxplot_factors = report_pars$boxplot$factors %||%
                   c(report_pars$boxplot_factor1, report_pars$boxplot_factor2)
boxplot_factors = boxplot_factors[!is.null(boxplot_factors) & nzchar(boxplot_factors)]
boxplot_display = report_pars$boxplot$display %||% "tabs"      # combined | tabs
boxplot_levels  = report_pars$boxplot$levels  %||% report_pars$boxplot_levels %||% c("Sample")
boxplot_head    = report_pars$boxplot$head    %||% report_pars$boxplot_head   %||% "Sample_type"

# Per-comparison boxplot toggle (faceted across comparisons)
plot_boxplot_per_comparison = isTRUE(report_pars$plot_boxplot_per_compound %||% TRUE)
yaxis_quant = report_pars$yaxis_quant %||% 0.95

# Sample-level summary bar plots
global_summary_factor = report_pars$global_summary_factor %||%
                        report_pars$global_boxpot_factor %||%
                        "Sample_type"

units = pmp_params$units %||% pmp_params$datatype %||% project_paths$datatype %||% ""

# Pre-built theme reused across per-compound loops
boxplot_theme = ggplot2::theme_classic() +
  ggplot2::theme(
    axis.text.x     = ggplot2::element_text(angle = 90, vjust = 0.5, hjust = 1),
    legend.position = "none"
  )
```

```{r logo}
logo_path = system.file("rmd", "imp_logo.jpg", package = "MRManalyzeR")
if(nzchar(logo_path)){
  htmltools::img(
    src   = knitr::image_uri(logo_path),
    alt   = "logo",
    style = "position:absolute; top:0; right:0; padding:70px;"
  )
}
```

# 1. Aim

Statistical analysis of processed `r units` data. Pre-defined comparisons,
correlations, and linear models from the YAML config; multivariate (PCA +
heatmap) inspection per comparison.

**Missing-value policy** (which sections impute and which don't):

| Section | NA handling |
|---|---|
| Comparisons (t-test / Wilcoxon / ANOVA / Kruskal / Tukey) | **No imputation** — raw values used; NAs dropped per test by `t.test()` / `wilcox.test()` / `aov()`. |
| Boxplots (per-compound + by-factor) | **No imputation** — raw values plotted; NAs simply absent from the boxes/jitter. |
| Correlations | **No imputation** — `pairwise.complete.obs` (each pair uses its own complete cases). |
| Linear models | `lm()` drops rows with NA in the feature or any predictor (per feature). |
| PCA (global + per-comparison) | Imputed per `pca.preprocess.impute` (default in this config: `r pca_impute``r if(pca_impute=="frac_min") sprintf(" with frac=%g", pca_impute_frac) else ""`), then transform → centre → scale. |
| Heatmap | After `r heatmap_transform` transform and per-feature z-scoring, residual NAs are set to 0 (= per-feature mean in z-units), then capped at ±2.5. |
| Volcano | Inherits the comparisons table, so no imputation in the underlying p-values; rows missing log2FC or p are dropped from the plot. |

# 2. Data import

```{r import}
importData     = combined_datamatrices
metadata       = importData$sample_meta
data_matrix    = importData$data
compoundInfo   = importData$variable_meta
compound_names = compoundInfo$Compound

# Biological-sample subset: dropped from sections that should not see blanks/QCs
# (global summary, global PCA). Per-comparison sections do their own subsetting
# via de_subset(), so they're unaffected.
if(!is.null(bio_filter_head) && bio_filter_head %in% colnames(metadata)){
  bio_de = de_subset(importData,
                     conditions = setNames(list(bio_filter_levels),
                                           bio_filter_head))
  bio_meta   = bio_de$sample_meta
  bio_matrix = bio_de$data
} else {
  bio_de     = importData
  bio_meta   = metadata
  bio_matrix = data_matrix
}

DT::datatable(
  metadata %>%
    dplyr::group_by(.data[[report_pars$summary_head %||% "Sample_type"]]) %>%
    dplyr::summarize(`Number of samples` = dplyr::n(), .groups = "drop")
)
```

`r if(!is.null(bio_filter_head)) sprintf("_Biological-sample filter applied to global summary + global PCA: \x60%s\x60 ∈ {%s} → %d samples (of %d)._", bio_filter_head, paste(bio_filter_levels, collapse=", "), nrow(bio_meta), nrow(metadata))`

# 3. Global summary {.tabset}

Mean intensity per compound (biological subset: `r nrow(bio_meta)` of
`r nrow(metadata)` samples), with `r global_summary_factor` levels dodged
side-by-side. Compounds in TargetLynx import order. Plot height auto-scales
with the number of features.

```{r global-helpers, include=FALSE}
.feature_bar = function(d, smeta, factor_name, scale, units){
  feats = colnames(d)
  grp   = smeta[[factor_name]]

  rows = lapply(feats, function(f){
    by_grp = tapply(d[[f]], grp, mean, na.rm = TRUE)
    data.frame(feature = f,
               .grp    = names(by_grp),
               value   = unname(by_grp),
               stringsAsFactors = FALSE)
  })
  m = dplyr::bind_rows(rows)

  if(scale == "log2") m$value = suppressWarnings(log2(m$value + 1))
  if(scale == "sqrt") m$value = suppressWarnings(sqrt(pmax(m$value, 0)))

  # Reverse so first imported feature sits at the top of the y-axis
  m$feature = factor(m$feature, levels = rev(feats))

  xlab = switch(scale,
                raw  = units,
                log2 = paste0("log2(", units, " + 1)"),
                sqrt = paste0("sqrt(", units, ")"))

  ggplot2::ggplot(m, ggplot2::aes(value, feature, fill = .grp)) +
    ggplot2::geom_col(position = ggplot2::position_dodge(width = 0.8),
                      width = 0.7) +
    ggplot2::scale_fill_brewer(palette = "Set1", na.value = "grey70") +
    ggplot2::labs(x = xlab, y = "Compound", fill = factor_name) +
    ggplot2::theme_classic(base_size = 11) +
    ggplot2::theme(
      axis.text.y        = ggplot2::element_text(size = 6),
      legend.position    = "bottom",
      panel.grid.major.x = ggplot2::element_line(color = "grey90", linewidth = 0.3)
    )
}
```

## 3.1 Raw (`r units`) {-}

```{r global-feature-raw, fig.width=7*fig_scale, fig.height=max(8, ncol(bio_matrix)*0.18)*fig_scale}
print(.feature_bar(bio_matrix, bio_meta, global_summary_factor, "raw", units))
```

## 3.2 log2 (`r units`) {-}

```{r global-feature-log2, fig.width=7*fig_scale, fig.height=max(8, ncol(bio_matrix)*0.18)*fig_scale}
print(.feature_bar(bio_matrix, bio_meta, global_summary_factor, "log2", units))
```

# 4. Comparisons

```{r comparisons-summary, results="asis"}
if(length(comparisons) == 0){
  cat("_No comparisons defined in YAML._\n")
} else {
  cat("Configured comparisons:\n\n")
  for(comp in comparisons){
    sub_str = if(is.null(comp$subset)) "(all samples)" else
      paste(sprintf("%s = %s", names(comp$subset), unlist(comp$subset)),
            collapse = ", ")
    cat(sprintf("- **%s** — subset: `%s`; compare `%s`: %s\n",
                comp$name, sub_str, comp$compare$factor,
                paste(comp$compare$levels, collapse = " vs ")))
  }
}
```

## 4.1 P-value table {-}

**Methods.** For each comparison the test method is auto-selected by group
count: 2 levels → t-test + Wilcoxon; ≥3 levels → ANOVA + Kruskal-Wallis,
plus Tukey HSD post-hoc per pair. P-values are Benjamini-Hochberg adjusted
across **all** comparisons × methods × features. Full statistic + df +
group means columns are in the `stats` tab of the xlsx.

```{r stats-table}
if(!is.null(stats_tables) && nrow(stats_tables$stats) > 0){
  DT::datatable(
    stats_tables$stats %>%
      dplyr::select(comparison, method, feature, group1, group2,
                    FC, log2FC, p_value, p_adj, significant) %>%
      dplyr::mutate(FC      = round(FC, 4),
                    log2FC  = round(log2FC, 4),
                    p_value = signif(p_value, 4),
                    p_adj   = signif(p_adj, 4)),
    filter = "top",
    options = list(pageLength = 25, scrollX = TRUE)
  )
} else {
  cat("_No comparison results._\n")
}
```

## 4.2 Boxplots per compound (faceted across comparisons) {-}

`r if(!plot_boxplot_per_comparison) "_Skipped — set_ \x60plot_boxplot_per_compound: True\x60 _in YAML to enable._"`

```{r comp-boxplots, results="asis", eval=plot_boxplot_per_comparison && length(comparisons) > 0, fig.width=10*fig_scale, fig.height=6*fig_scale}
# Build one long data frame across all comparisons, then facet.
# Combine user `subset:` with the comparison level restriction in one
# de_subset() call (avoid post-hoc DE mutation which trips SE's assay check).
long_rows = list()
for(comp in comparisons){
  cond = comp$subset
  cond = if(is.null(cond)) list() else as.list(cond)
  cond[[comp$compare$factor]] = comp$compare$levels
  de_sub = de_subset(importData, conditions = cond)
  if(nrow(de_sub$sample_meta) == 0) next

  d = de_sub$data
  d$.group      = factor(de_sub$sample_meta[[comp$compare$factor]],
                         levels = comp$compare$levels)
  d$.comparison = comp$name
  long_rows[[length(long_rows) + 1]] = d
}
long = dplyr::bind_rows(long_rows)

make_safe_id = function(x) gsub("[^A-Za-z0-9_\\-]", "_", x)
safe_ids = vapply(compound_names, make_safe_id, character(1))

is_detected = vapply(compound_names, function(cmp){
  cc = which(colnames(long) == cmp)
  length(cc) > 0 && !all(is.na(unlist(long[, cc])))
}, logical(1))
detected_compounds = compound_names[is_detected]
detected_ids       = safe_ids[is_detected]

# Alphabetise dropdown order (T3.5-7); plot blocks below iterate in import order.
.alpha_idx        = order(detected_compounds)
detected_alpha    = detected_compounds[.alpha_idx]
detected_ids_alpha = detected_ids[.alpha_idx]

cat('
<label for="boxplot_select" style="font-weight:600;">Select compound:</label>
<select id="boxplot_select" style="margin: 0 0 12px 8px; padding: 4px 6px;"
        onchange="showBoxplot(this.value)">
')
for(i in seq_along(detected_alpha)){
  sel = if(i == 1) " selected" else ""
  cat(sprintf('<option value="%s"%s>%s</option>\n',
              detected_ids_alpha[i], sel, detected_alpha[i]))
}
cat('</select>
<script>
function showBoxplot(id){
  var plots = document.querySelectorAll(".cmp_boxplot_block");
  for (var i=0; i<plots.length; i++) { plots[i].style.display = "none"; }
  var el = document.getElementById("cmp_boxplot_" + id);
  if (el) el.style.display = "block";
}
</script>
')

for(i in seq_along(detected_compounds)){
  compound = detected_compounds[i]
  cid      = detected_ids[i]
  axis_limit = quantile(long[[compound]], prob = yaxis_quant, na.rm = TRUE) * 1.3

  p = ggplot2::ggplot(long, ggplot2::aes(x = .group, y = .data[[compound]],
                                         fill = .group)) +
        ggplot2::geom_boxplot(alpha = 0.3, outlier.shape = NA) +
        ggplot2::geom_jitter(width = 0.15, alpha = 0.6, size = 2) +
        ggplot2::ylim(c(0, axis_limit)) +
        ggplot2::labs(x = NULL, y = units, title = compound) +
        ggplot2::facet_wrap(~ .comparison, scales = "free_x", nrow = 1) +
        boxplot_theme

  # Show alphabetically-first detected compound by default (matches dropdown)
  display = if(detected_ids[i] == detected_ids_alpha[1]) "block" else "none"
  cat(sprintf('<div class="cmp_boxplot_block" id="cmp_boxplot_%s" style="display:%s;">\n',
              cid, display))
  print(p)
  cat("</div>\n")
}
```

## 4.3 Boxplots by factor(s) {-}

`r if(!boxplot_enabled) "_Skipped — set_ \x60boxplot.enabled: True\x60 _in YAML to enable._"`

`r if(boxplot_enabled) sprintf("Per-compound boxplot grouped by_ \x60%s\x60_. Display mode:_ \x60%s\x60_._", paste(boxplot_factors, collapse=", "), boxplot_display)`

```{r factor-boxplots, results="asis", eval=boxplot_enabled && length(boxplot_factors) > 0, fig.width=10*fig_scale, fig.height=6*fig_scale}
# Filter to requested sample types
keep = metadata[[boxplot_head]] %in% boxplot_levels
plot_meta = metadata[keep, , drop = FALSE]
plot_data = data_matrix[keep, , drop = FALSE]

# Detect compounds with any data
is_detected_fb = vapply(compound_names, function(cmp){
  cc = which(colnames(plot_data) == cmp)
  length(cc) > 0 && !all(is.na(unlist(plot_data[, cc])))
}, logical(1))
detected_fb     = compound_names[is_detected_fb]
make_safe_id_fb = function(x) gsub("[^A-Za-z0-9_\\-]", "_", x)
detected_ids_fb = vapply(detected_fb, make_safe_id_fb, character(1))

# Build per-tab grouping vectors
tab_groupings = list()
if(boxplot_display == "combined" || length(boxplot_factors) == 1){
  lab = paste(boxplot_factors, collapse = " × ")
  tab_groupings[[lab]] = do.call(paste, c(plot_meta[boxplot_factors], sep = "_"))
} else {
  for(f in boxplot_factors)
    tab_groupings[[f]] = as.character(plot_meta[[f]])
  if(length(boxplot_factors) >= 2){
    lab = paste(boxplot_factors, collapse = " × ")
    tab_groupings[[lab]] = do.call(paste, c(plot_meta[boxplot_factors], sep = "_"))
  }
}

cat('\n### {.tabset .tabset-fade}\n\n')

for(tab_name in names(tab_groupings)){
  cat(sprintf('\n#### %s {-}\n\n', tab_name))

  grouping = tab_groupings[[tab_name]]
  tab_safe = make_safe_id_fb(tab_name)

  # Alphabetised dropdown (T3.5-7); plot blocks below in import order.
  alpha_idx_fb       = order(detected_fb)
  detected_alpha_fb  = detected_fb[alpha_idx_fb]
  detected_ids_alpha_fb = detected_ids_fb[alpha_idx_fb]

  cat(sprintf('
<label for="fbox_select_%s" style="font-weight:600;">Select compound:</label>
<select id="fbox_select_%s" style="margin: 0 0 12px 8px; padding: 4px 6px;"
        onchange="showFactorBoxplot_%s(this.value)">
', tab_safe, tab_safe, tab_safe))
  for(i in seq_along(detected_alpha_fb)){
    sel = if(i == 1) " selected" else ""
    cat(sprintf('<option value="%s"%s>%s</option>\n',
                detected_ids_alpha_fb[i], sel, detected_alpha_fb[i]))
  }
  cat(sprintf('</select>
<script>
function showFactorBoxplot_%s(id){
  var plots = document.querySelectorAll(".fbox_block_%s");
  for (var i=0; i<plots.length; i++) { plots[i].style.display = "none"; }
  var el = document.getElementById("fbox_%s_" + id);
  if (el) el.style.display = "block";
}
</script>
', tab_safe, tab_safe, tab_safe))

  for(i in seq_along(detected_fb)){
    compound = detected_fb[i]
    cid      = detected_ids_fb[i]
    axis_limit = quantile(plot_data[[compound]], prob = yaxis_quant, na.rm = TRUE) * 1.3
    df_plot = data.frame(group = grouping, value = plot_data[[compound]])

    p = ggplot2::ggplot(df_plot, ggplot2::aes(group, value, fill = group)) +
          ggplot2::geom_boxplot(alpha = 0.3, outlier.shape = NA) +
          ggplot2::geom_jitter(width = 0.15, alpha = 0.6, size = 2) +
          ggplot2::ylim(c(0, axis_limit)) +
          ggplot2::scale_fill_brewer(palette = "Set1", na.value = "grey60") +
          ggplot2::labs(x = NULL, y = units, title = compound) +
          boxplot_theme

    # Show alphabetically-first compound by default (matches dropdown)
    display = if(detected_ids_fb[i] == detected_ids_alpha_fb[1]) "block" else "none"
    cat(sprintf('<div class="fbox_block_%s" id="fbox_%s_%s" style="display:%s;">\n',
                tab_safe, tab_safe, cid, display))
    print(p)
    cat("</div>\n")
  }
}
```

## 4.4 PCA per comparison {.tabset .tabset-fade}

`r if(!pca_include) "_Skipped — set_ \x60pca.include: True\x60 _in YAML to enable._"`

For each comparison: filter samples by the comparison's `subset:`, run PCA on
that slice, colour points by the comparison's `compare.factor`. The first
tab is a **global** PCA across the biological subset, coloured by
`r global_summary_factor`.

**Components plotted:** PC`r pca_pc_x` × PC`r pca_pc_y`
(set via `stats_report.pca.components: [x, y]`).

**Preprocessing pipeline (per-PCA actual steps echoed inside each tab):**
`impute=`*`r pca_impute`*`r if(pca_impute=="frac_min") sprintf(" (frac=%g)", pca_impute_frac) else ""`,
`transform=`*`r pca_transform`*,
`center=`*`r pca_center`*, `scale=`*`r pca_scale`*.

```{r pca-helpers, include=FALSE}
# Local helper: outlier flag on chosen PCs
.pca_outliers = function(x, y){
  zx = scale(x)[, 1]; zy = scale(y)[, 1]
  abs(zx) > 2.5 | abs(zy) > 2.5
}

# Render one PCA tab: preprocessing trace, scores plot, optional loadings.
.pca_panel = function(res_pca, smeta, vmeta, color_factor,
                      pc_x, pc_y, label_mode, label_factor,
                      ellipse_mode, ellipse_conf,
                      loadings_show, loadings_color_by){

  # ---- preprocessing trace (S5) ----
  cat("**Preprocessing pipeline applied:**\n\n")
  for(s in res_pca$steps) cat(sprintf("- %s\n", s))
  cat(sprintf("\n_Final matrix: %d samples × %d features._\n\n",
              res_pca$n_samples, res_pca$n_features))

  pr = res_pca$pr
  if(is.null(pr)){
    cat(sprintf("_PCA could not be fit (n_samples=%d, n_features=%d)._\n\n",
                res_pca$n_samples, res_pca$n_features))
    return(invisible(NULL))
  }

  # ---- validate component selection (S4) ----
  pc_max = ncol(pr$x)
  if(pc_x > pc_max || pc_y > pc_max){
    cat(sprintf("_Requested PC%d × PC%d but only %d components available — falling back to PC1 × PC%d._\n\n",
                pc_x, pc_y, pc_max, min(2, pc_max)))
    pc_x = 1; pc_y = min(2, pc_max)
  }
  vexp = (pr$sdev^2) / sum(pr$sdev^2) * 100
  pcx_lab = sprintf("PC%d", pc_x); pcy_lab = sprintf("PC%d", pc_y)

  # ---- scores plot ----
  scores = data.frame(
    pcx   = pr$x[, pc_x],
    pcy   = pr$x[, pc_y],
    group = smeta[[color_factor]],
    stringsAsFactors = FALSE
  )
  scores$label = if(!is.null(label_factor) && label_factor %in% colnames(smeta))
                   as.character(smeta[[label_factor]]) else rownames(scores)

  is_num = is.numeric(scores$group)

  p_scores = ggplot2::ggplot(scores, ggplot2::aes(pcx, pcy, color = group)) +
        ggplot2::geom_point(size = 3, alpha = 0.85) +
        ggplot2::labs(x = sprintf("%s (%.1f%%)", pcx_lab, vexp[pc_x]),
                      y = sprintf("%s (%.1f%%)", pcy_lab, vexp[pc_y]),
                      color = color_factor) +
        ggplot2::theme_classic(base_size = 13) +
        ggplot2::theme(legend.position = "bottom")
  # Continuous (numeric) → viridis_c; categorical → Set1 brewer (red/blue/green/...)
  p_scores = if(is_num) p_scores + ggplot2::scale_color_viridis_c()
             else       p_scores + ggplot2::scale_color_brewer(palette = "Set1",
                                                               na.value = "grey60")

  if(ellipse_mode == "group" && !is_num && length(unique(scores$group)) > 1)
    p_scores = p_scores + ggplot2::stat_ellipse(ggplot2::aes(group = group),
                                                 type = "norm", level = ellipse_conf,
                                                 linewidth = 0.4)
  if(label_mode == "all"){
    p_scores = p_scores + ggplot2::geom_text(ggplot2::aes(label = label),
                                              size = 2.5, vjust = -1, show.legend = FALSE)
  } else if(label_mode == "outliers"){
    out_idx = .pca_outliers(scores$pcx, scores$pcy)
    if(any(out_idx))
      p_scores = p_scores + ggplot2::geom_text(data = scores[out_idx, ],
                                                ggplot2::aes(label = label),
                                                size = 2.5, vjust = -1, show.legend = FALSE)
  }

  # ---- loadings plot (S7) ----
  # When loadings are off, just print scores at full canvas size and return.
  if(!isTRUE(loadings_show)){
    print(p_scores); cat("\n\n"); return(invisible(NULL))
  }

  feats = rownames(pr$rotation)
  has_color = !is.null(loadings_color_by) && loadings_color_by %in% colnames(vmeta)
  color_vec = if(has_color){
    cv = as.character(vmeta[[loadings_color_by]][match(feats, vmeta$Compound)])
    cv[is.na(cv)] = "NA"; cv
  } else {
    rep("(all)", length(feats))
  }

  if(has_color){
    cat(sprintf("**Loadings** _(colour =_ \x60%s\x60_; bars grouped by colour, sorted by loading within each group)_\n\n",
                loadings_color_by))
  } else if(!is.null(loadings_color_by)){
    cat(sprintf("**Loadings** _(colour column \x60%s\x60 not in feature_metadata)_\n\n",
                loadings_color_by))
  } else {
    cat("**Loadings**\n\n")
  }

  # Per-PC ordering: group features by colour value (first-seen order
  # = variable_meta import order), then sort within each group by loading
  # descending. Each PC has its own ordering, so encode the y-axis as a
  # composite "feature__PC" factor and strip the suffix in the label
  # function. This lets the whole loadings figure live in ONE ggplot with
  # a single shared legend.
  groups_in_order = unique(color_vec)

  build_pc_df = function(loading_vec, pc_lab){
    ord_feats = character(0)
    for(g in groups_in_order){
      idx_g = which(color_vec == g)
      if(!length(idx_g)) next
      sorted = idx_g[order(loading_vec[idx_g], decreasing = TRUE)]
      ord_feats = c(ord_feats, feats[sorted])
    }
    data.frame(
      feature = feats,
      feat_pc = paste0(feats, "__", pc_lab),
      loading = loading_vec,
      .color  = factor(color_vec, levels = groups_in_order),
      PC      = pc_lab,
      .order_levels = I(list(rev(paste0(ord_feats, "__", pc_lab)))),
      stringsAsFactors = FALSE
    )
  }
  ld_x_df = build_pc_df(pr$rotation[, pc_x], pcx_lab)
  ld_y_df = build_pc_df(pr$rotation[, pc_y], pcy_lab)
  ld_long = rbind(ld_x_df, ld_y_df)

  # All composite levels in the order required for ggplot to honour
  # per-facet ordering (concatenate both PC orderings; ggplot will pick
  # the matching ones in each facet).
  combined_levels = c(ld_x_df$.order_levels[[1]], ld_y_df$.order_levels[[1]])
  ld_long$feat_pc = factor(ld_long$feat_pc, levels = combined_levels)
  ld_long$PC      = factor(ld_long$PC, levels = c(pcx_lab, pcy_lab))

  p_loadings = ggplot2::ggplot(ld_long,
                               ggplot2::aes(loading, feat_pc, fill = .color)) +
        ggplot2::geom_col() +
        ggplot2::geom_vline(xintercept = 0, color = "grey50", linewidth = 0.3) +
        ggplot2::facet_wrap(~ PC, scales = "free", nrow = 1) +
        # Strip the "__PCk" suffix when rendering y-axis labels
        ggplot2::scale_y_discrete(labels = function(x) sub("__[^_]+$", "", x)) +
        ggplot2::scale_fill_viridis_d(end = 0.9, na.value = "grey60",
                                      drop = FALSE) +
        ggplot2::labs(x = "Loading", y = NULL,
                      fill = loadings_color_by %||% "") +
        ggplot2::theme_classic(base_size = 10) +
        ggplot2::theme(
          axis.text.y     = ggplot2::element_text(size = 6),
          strip.text      = ggplot2::element_text(face = "bold"),
          legend.position = "bottom"
        ) +
        ggplot2::guides(fill = ggplot2::guide_legend(nrow = 3, byrow = TRUE))

  # Combine: small square scores plot on top, full-width faceted loadings below.
  if(requireNamespace("cowplot", quietly = TRUE)){
    scores_centered = cowplot::plot_grid(
      NULL,
      p_scores + ggplot2::theme(aspect.ratio = 1,
                                plot.margin = ggplot2::margin(5, 5, 5, 5)),
      NULL, ncol = 3, rel_widths = c(1, 2, 1)
    )
    combined = cowplot::plot_grid(scores_centered, p_loadings,
                                  ncol = 1, rel_heights = c(0.35, 1))
    print(combined)
  } else {
    print(p_scores); cat("\n\n")
    print(p_loadings)
  }
  cat("\n\n")
}
```

```{r comp-pca, results="asis", eval=pca_include, fig.width=10*fig_scale, fig.height=max(7, ncol(bio_de$data)*0.18)*fig_scale}
# ---- Global PCA tab (biological subset, coloured by global_summary_factor) ----
# Optionally narrow further via pca.subset: (multi-key, AND'd) on top of the
# biological_filter that already produced bio_de.
pca_global_de = if(length(pca_global_subset)){
  de_subset(bio_de, conditions = as.list(pca_global_subset))
} else {
  bio_de
}

global_subset_lbl = if(length(pca_global_subset)){
  sprintf(" — subset: %s",
          paste(sprintf("%s=%s", names(pca_global_subset),
                        unlist(pca_global_subset)), collapse = ", "))
} else {
  ""
}
cat(sprintf("\n### Global (%d samples; coloured by %s)%s {-}\n\n",
            nrow(pca_global_de$sample_meta),
            global_summary_factor, global_subset_lbl))

res_pca_global = run_pca_pipeline(as.matrix(pca_global_de$data),
                                  impute      = pca_impute,
                                  impute_frac = pca_impute_frac,
                                  transform   = pca_transform,
                                  center      = pca_center,
                                  scale       = pca_scale)

.pca_panel(res_pca_global,
           smeta = pca_global_de$sample_meta,
           vmeta = pca_global_de$variable_meta,
           color_factor      = global_summary_factor,
           pc_x              = pca_pc_x,
           pc_y              = pca_pc_y,
           label_mode        = pca_label,
           label_factor      = pca_label_factor,
           ellipse_mode      = pca_ellipse,
           ellipse_conf      = pca_ellipse_conf,
           loadings_show     = pca_loadings_show,
           loadings_color_by = pca_loadings_color_by)

# ---- Per-comparison tabs ----
if(length(comparisons) == 0){
  cat("_No comparisons defined._\n\n")
}

for(comp in comparisons){
  cat(sprintf("\n### %s {-}\n\n", comp$name))

  cond = comp$subset
  cond = if(is.null(cond)) list() else as.list(cond)
  cond[[comp$compare$factor]] = comp$compare$levels
  de_sub = de_subset(importData, conditions = cond)

  if(nrow(de_sub$data) < 3 || ncol(de_sub$data) < 2){
    cat("_Insufficient data._\n\n"); next
  }

  res_pca = run_pca_pipeline(as.matrix(de_sub$data),
                             impute      = pca_impute,
                             impute_frac = pca_impute_frac,
                             transform   = pca_transform,
                             center      = pca_center,
                             scale       = pca_scale)

  .pca_panel(res_pca,
             smeta = de_sub$sample_meta,
             vmeta = de_sub$variable_meta,
             color_factor      = comp$compare$factor,
             pc_x              = pca_pc_x,
             pc_y              = pca_pc_y,
             label_mode        = pca_label,
             label_factor      = pca_label_factor,
             ellipse_mode      = pca_ellipse,
             ellipse_conf      = pca_ellipse_conf,
             loadings_show     = pca_loadings_show,
             loadings_color_by = pca_loadings_color_by)
}
```

## 4.5 Volcano per comparison {.tabset .tabset-fade}

`r if(!volcano_include) "_Skipped — set_ \x60volcano.include: True\x60 _in YAML to enable._"`

**Methods.** One volcano per 2-group comparison directly from the t-test
rows. For ≥3-group comparisons we plot one volcano per Tukey HSD pair.
X = log2 fold-change (mean_g2 / mean_g1); Y = -log10 of `r if(volcano_use_adj) "BH-adjusted" else "raw"` p
(toggle via `stats_report.volcano.use_adjusted`).
Top `r volcano_top_n` features by significance are labelled.
Significance threshold dashed line at p `r if(volcano_use_adj) "(BH-adjusted)" else "(raw)"` = `r sig_threshold`.

```{r comp-volcano, results="asis", eval=volcano_include && !is.null(stats_tables) && nrow(stats_tables$stats) > 0, fig.width=7*fig_scale, fig.height=6*fig_scale}
# Volcano sources:
#   2-group comparisons -> use the t_test rows directly (have log2FC + p_adj)
#   3+-group comparisons -> use the tukey post-hoc rows (one volcano per pair)
.volc_p_col = if(volcano_use_adj) "p_adj" else "p_value"
.volc_p_lbl = if(volcano_use_adj) sprintf("BH-adjusted p (%s)", p_adjust) else "raw p"

volc_data = stats_tables$stats %>%
  dplyr::filter(method %in% c("t_test", "tukey"),
                !is.na(log2FC), !is.na(.data[[.volc_p_col]]))

if(nrow(volc_data) == 0){
  cat("_No volcano-eligible rows._\n")
} else {
  for(comp_name in unique(volc_data$comparison)){
    sub = volc_data %>% dplyr::filter(comparison == comp_name)
    pair_groups = if("tukey" %in% sub$method)
      unique(sub %>% dplyr::filter(method == "tukey") %>%
               dplyr::transmute(pair = paste(group1, "vs", group2)) %>%
               dplyr::pull(pair))
    else NA

    for(pair_lbl in (if(is.na(pair_groups[1])) NA else pair_groups)){
      this = if(is.na(pair_lbl)) sub else
        sub %>% dplyr::filter(method == "tukey",
                              paste(group1, "vs", group2) == pair_lbl)

      cat(sprintf("\n### %s%s {-}\n\n", comp_name,
                  if(is.na(pair_lbl)) "" else paste0(" — ", pair_lbl)))

      this$neglog10p = -log10(this[[.volc_p_col]])
      this$sig = ifelse(this[[.volc_p_col]] < sig_threshold, "sig", "ns")

      # Always label all significant features; if more than top_n_label sig
      # features exist, label only the top_n_label most significant.
      sig_rows = this %>% dplyr::filter(sig == "sig") %>%
                          dplyr::arrange(.data[[.volc_p_col]])
      to_label = if(nrow(sig_rows) <= volcano_top_n) sig_rows
                 else utils::head(sig_rows, volcano_top_n)

      p = ggplot2::ggplot(this, ggplot2::aes(log2FC, neglog10p, color = sig)) +
            ggplot2::geom_point(alpha = 0.7, size = 2) +
            ggplot2::scale_color_manual(values = c(sig = "#d62728", ns = "grey60")) +
            ggplot2::geom_hline(yintercept = -log10(sig_threshold),
                                linetype = "dashed", color = "grey40") +
            ggplot2::geom_vline(xintercept = 0, linetype = "dashed", color = "grey40") +
            ggplot2::labs(x = "log2 fold-change",
                          y = sprintf("-log10(%s)", .volc_p_lbl),
                          color = NULL,
                          title = sprintf("%s%s",
                                          comp_name,
                                          if(is.na(pair_lbl)) "" else paste0(" — ", pair_lbl))) +
            ggplot2::theme_classic(base_size = 13) +
            ggplot2::theme(legend.position = "bottom")

      if(nrow(to_label) > 0){
        # ggrepel is now an Imports dep so it's always available; but keep
        # a geom_text fallback in case of unusual install paths.
        if(requireNamespace("ggrepel", quietly = TRUE)){
          p = p + ggrepel::geom_text_repel(data = to_label,
                                           ggplot2::aes(label = feature),
                                           color = "#d62728",
                                           fontface = "bold",
                                           size = 3.2,
                                           max.overlaps = Inf,
                                           force = 4,
                                           min.segment.length = 0,
                                           segment.color = "#d62728",
                                           segment.size = 0.3,
                                           show.legend = FALSE)
        } else {
          p = p + ggplot2::geom_text(data = to_label,
                                     ggplot2::aes(label = feature),
                                     color = "#d62728", fontface = "bold",
                                     size = 3.2, vjust = -0.7,
                                     show.legend = FALSE)
        }
      }
      print(p)
      cat("\n\n")
    }
  }
}
```

## 4.6 Heatmap (samples × features) {-}

`r if(!heatmap_include) "_Skipped — set_ \x60heatmap.include: True\x60 _in YAML to enable._"`

`r if(heatmap_include) sprintf("Sample × feature heatmap of the biological subset (%d samples × %d features). Cells coloured by per-feature %s of %s data. Sample rows annotated by_ \x60%s\x60_; feature columns%s.", nrow(bio_matrix), ncol(bio_matrix), heatmap_scale, heatmap_transform, heatmap_color_samples_by, if(!is.null(heatmap_group_features_by)) sprintf(" annotated and grouped by_ \x60%s\x60", heatmap_group_features_by) else "")`

```{r heatmap, eval=heatmap_include, fig.width=max(10, ncol(bio_matrix)*0.18)*fig_scale, fig.height=max(5, nrow(bio_matrix)*0.22 + 2)*fig_scale}
if(!requireNamespace("pheatmap", quietly = TRUE)){
  cat("_Install \x60pheatmap\x60 to render the heatmap section: \x60install.packages('pheatmap')\x60._\n")
} else {

  M = as.matrix(bio_matrix)

  # Row labels: prefer biological Sample_ID over the acquisition file Name
  # (T3.5-6). Keep rownames in sync with bio_meta order.
  if("Sample_ID" %in% colnames(bio_meta))
    rownames(M) = as.character(bio_meta$Sample_ID)

  # 1. Optional transform
  if(heatmap_transform == "log2"){
    M[M < 0 & !is.na(M)] = 0
    M = log2(M + 1)
  } else if(heatmap_transform == "sqrt"){
    M[M < 0 & !is.na(M)] = 0
    M = sqrt(M)
  }

  # 2. Drop zero-variance features (would produce NaN after z-scoring)
  vars   = apply(M, 2, stats::var, na.rm = TRUE)
  keep_f = is.finite(vars) & vars > 0
  if(sum(!keep_f) > 0)
    message(sprintf("[heatmap] Dropping %d zero-variance features.", sum(!keep_f)))
  M = M[, keep_f, drop = FALSE]

  # 3. Per-feature z-score (default) so features with different dynamic ranges
  #    are comparable.
  if(heatmap_scale == "z"){
    M = scale(M, center = TRUE, scale = TRUE)
  }
  # Replace any residual NA with 0 so pheatmap doesn't choke
  M[is.na(M)] = 0
  # Cap extremes so the colour scale is informative
  cap = 2.5
  M[M >  cap] =  cap
  M[M < -cap] = -cap

  # 4. Sample annotation (one column = the colour-by factor)
  ann_row = data.frame(row.names = rownames(M))
  ann_row[[heatmap_color_samples_by]] = bio_meta[[heatmap_color_samples_by]]

  # 5. Feature annotation + grouping (if a feature_meta column was given)
  ann_col  = NULL
  gaps_col = NULL
  vmeta    = bio_de$variable_meta
  if(!is.null(heatmap_group_features_by) &&
     heatmap_group_features_by %in% colnames(vmeta)){

    feats     = colnames(M)
    fgroups   = as.character(vmeta[[heatmap_group_features_by]][match(feats, vmeta$Compound)])
    fgroups[is.na(fgroups)] = "NA"

    # Order columns: by group (preserve first-seen order), then optionally
    # cluster within each group block.
    group_levels = unique(fgroups)
    ordered_cols = character(0)
    for(g in group_levels){
      in_g = feats[fgroups == g]
      if(heatmap_cluster_within && length(in_g) > 1){
        sub_M = M[, in_g, drop = FALSE]
        d = stats::dist(t(sub_M))
        if(all(is.finite(d))){
          hc = stats::hclust(d, method = "average")
          in_g = in_g[hc$order]
        }
      }
      ordered_cols = c(ordered_cols, in_g)
    }
    M = M[, ordered_cols, drop = FALSE]

    # Re-derive groups in new order, build annotation + gap positions
    fgroups2 = as.character(vmeta[[heatmap_group_features_by]][match(ordered_cols, vmeta$Compound)])
    fgroups2[is.na(fgroups2)] = "NA"
    ann_col = data.frame(row.names = ordered_cols)
    ann_col[[heatmap_group_features_by]] = factor(fgroups2,
                                                  levels = group_levels)

    block = rle(fgroups2)$lengths
    gaps_col = cumsum(block); gaps_col = gaps_col[-length(gaps_col)]
  }

  # 6. Sort rows by sample-colour group so the row annotation forms blocks
  ord_s = order(as.character(ann_row[[heatmap_color_samples_by]]))
  M       = M[ord_s, , drop = FALSE]
  ann_row = ann_row[ord_s, , drop = FALSE]
  block_r = rle(as.character(ann_row[[heatmap_color_samples_by]]))$lengths
  gaps_row = cumsum(block_r); gaps_row = gaps_row[-length(gaps_row)]

  # 7. Diverging blue→white→red palette (matches request)
  pal = grDevices::colorRampPalette(c("#2166ac", "white", "#b2182b"))(101)
  brk = seq(-cap, cap, length.out = 102)

  # Annotation palette: Set1 brewer (red/blue/green/...). Use explicit
  # red/blue when only two sample groups so it visually anchors the
  # treatment contrast; otherwise let pheatmap auto-pick from a Set1 sweep.
  .qual = function(n){
    if(requireNamespace("RColorBrewer", quietly = TRUE) && n <= 9 && n >= 3)
      return(RColorBrewer::brewer.pal(n, "Set1"))
    if(requireNamespace("RColorBrewer", quietly = TRUE) && n > 9)
      return(grDevices::colorRampPalette(RColorBrewer::brewer.pal(9, "Set1"))(n))
    grDevices::hcl.colors(max(n, 2), palette = "Dark 3")
  }
  build_ann_colors = function(ann){
    if(is.null(ann)) return(NULL)
    out = lapply(ann, function(v){
      lv = unique(as.character(v[!is.na(v)]))
      if(length(lv) == 2) setNames(c("#377eb8", "#e41a1c"), lv)   # blue, red
      else                setNames(.qual(length(lv)), lv)
    })
    names(out) = colnames(ann)
    out
  }
  ann_colors = c(build_ann_colors(ann_row), build_ann_colors(ann_col))

  pheatmap::pheatmap(
    M,
    cluster_rows      = heatmap_cluster_rows,
    cluster_cols      = FALSE,             # columns pre-ordered above
    gaps_row          = if(length(gaps_row)) gaps_row else NULL,
    gaps_col          = if(length(gaps_col)) gaps_col else NULL,
    annotation_row    = ann_row,
    annotation_col    = ann_col,
    annotation_colors = if(length(ann_colors)) ann_colors else NA,
    color             = pal,
    breaks            = brk,
    fontsize_row      = 8,
    fontsize_col      = 6,
    border_color      = NA,
    legend_breaks     = c(-cap, -1, 0, 1, cap),
    legend_labels     = c(sprintf("≤ -%g", cap), "-1", "0", "1", sprintf("≥ %g", cap)),
    main              = sprintf("Z-score heatmap (%s of %s)",
                                heatmap_scale, heatmap_transform)
  )
}
```

# 5. Correlations (feature × feature) {.tabset}

Pairwise feature correlations within each subset. Long table written to the
`correlations` tab in the xlsx.

## 5.1 Long table {-}

```{r corr-table}
if(!is.null(stats_tables) && nrow(stats_tables$correlations) > 0){
  DT::datatable(
    stats_tables$correlations %>%
      dplyr::mutate(dplyr::across(c(estimate, statistic), ~ round(., 4)),
                    dplyr::across(c(p_value, p_adj), ~ signif(., 4))),
    filter = "top",
    options = list(pageLength = 25, scrollX = TRUE)
  )
} else {
  cat("_No correlations defined._\n")
}
```

## 5.2 Heatmaps {-}

```{r corr-heatmaps, results="asis", eval=!is.null(stats_tables) && nrow(stats_tables$correlations) > 0, fig.width=8*fig_scale, fig.height=7*fig_scale}
# One heatmap per (correlation × subset × method).
# Optional clustering controlled by `correlations[i].cluster:` in YAML; we
# read it back from `report_pars$correlations` (the YAML list) by name.
corr_yaml = report_pars$correlations %||% list()
get_yaml_opt = function(name, opt, default){
  for(c in corr_yaml) if(identical(c$name, name) && !is.null(c[[opt]])) return(c[[opt]])
  default
}

cdf = stats_tables$correlations
combos = unique(cdf[, c("correlation", "subset", "method")])

for(i in seq_len(nrow(combos))){
  cn = combos$correlation[i]; sb = combos$subset[i]; mt = combos$method[i]
  cluster_yn   = isTRUE(get_yaml_opt(cn, "cluster",      TRUE))
  show_values  = isTRUE(get_yaml_opt(cn, "show_values",  FALSE))

  cat(sprintf("\n### %s — %s — %s {-}\n\n", cn, sb, mt))

  sub = cdf[cdf$correlation == cn & cdf$subset == sb & cdf$method == mt, ,
            drop = FALSE]
  if(nrow(sub) == 0){ cat("_(empty)_\n\n"); next }

  # Reshape lower-triangle long → full square matrix.
  # Use first-seen order (= import order from .pairwise_cor_long); the
  # cluster step below may reorder it.
  feats = unique(c(sub$feature_a, sub$feature_b))
  M = matrix(NA_real_, nrow = length(feats), ncol = length(feats),
             dimnames = list(feats, feats))
  for(k in seq_len(nrow(sub))){
    a = sub$feature_a[k]; b = sub$feature_b[k]; r = sub$estimate[k]
    M[a, b] = r; M[b, a] = r
  }
  # Self-self cells are dropped from the long table by default — leave them
  # as NA so the heatmap renders the diagonal in `na.value` grey, signalling
  # "self-correlation, omitted". (When include_self: True they're filled.)

  if(cluster_yn && length(feats) > 2){
    d  = stats::as.dist(1 - abs(M))
    hc = tryCatch(stats::hclust(d, method = "average"), error = function(e) NULL)
    if(!is.null(hc)) feats = feats[hc$order]
  }

  long = expand.grid(feature_a = feats, feature_b = feats,
                     stringsAsFactors = FALSE)
  long$r = M[cbind(long$feature_a, long$feature_b)]
  long$feature_a = factor(long$feature_a, levels = feats)
  long$feature_b = factor(long$feature_b, levels = rev(feats))

  p = ggplot2::ggplot(long, ggplot2::aes(feature_a, feature_b, fill = r)) +
        ggplot2::geom_tile(color = "white", linewidth = 0.1) +
        ggplot2::scale_fill_gradient2(low = "#2166ac", mid = "white",
                                      high = "#b2182b", midpoint = 0,
                                      limits = c(-1, 1), na.value = "grey90") +
        ggplot2::coord_fixed() +
        ggplot2::theme_classic(base_size = 10) +
        ggplot2::theme(
          axis.text.x = ggplot2::element_text(angle = 90, vjust = 0.5,
                                              hjust = 1, size = 6),
          axis.text.y = ggplot2::element_text(size = 6),
          legend.position = "right"
        ) +
        ggplot2::labs(x = NULL, y = NULL, fill = mt,
                      title = sprintf("%s — %s — %s", cn, sb, mt))

  if(show_values && length(feats) <= 30){
    p = p + ggplot2::geom_text(ggplot2::aes(label = sprintf("%.2f", r)),
                               size = 2)
  }
  print(p); cat("\n\n")
}
```

# 6. Linear models

For each feature we fit `feature ~ formula` (one `lm()` per feature).
Each row of the table below is **one coefficient × feature**.

**Reading the terms** (using a model like `~ Treatment + Sex` with reference
levels `Treatment = Ctrl`, `Sex = Female`):

- **`Intercept (ref: Treatment=Ctrl; Sex=Female)`** — the model's expected mean
  value of the feature for a sample at all reference levels (a Ctrl female).
  Often biologically uninteresting on raw intensity data; mostly a baseline.
- **`Treatment: HDM_Curdlan (vs Ctrl)`** — how much higher (positive estimate)
  or lower (negative) the feature is in HDM_Curdlan vs Ctrl, **after adjusting
  for Sex**. This is the treatment effect controlling for sex imbalance.
- **`Sex: Male (vs Female)`** — how much the feature differs in males vs females,
  **after adjusting for Treatment**.
- **Continuous predictors** (e.g. `~ Treatment + Age`) get a single term —
  the per-unit slope of the feature on that variable.
- **Interactions** (e.g. `~ Treatment * Sex`) add a `Treatment:Sex` term that
  asks "does the treatment effect itself differ between males and females".

**`significant` (p_adj < `r sig_threshold`)**: the coefficient is statistically
distinguishable from zero after Benjamini-Hochberg correction across all
features × terms in this run. Biologically: that predictor has a measurable
additive effect on this lipid's intensity, beyond what the other terms in the
formula already explain. A significant `Treatment:HDM_Curdlan` therefore says
the lipid genuinely shifts with treatment in a way that isn't sex confounding.

`term_raw` keeps the original `lm()` coefficient name (e.g.
`TreatmentHDM_Curdlan`) for joining with downstream scripts.

```{r lm-table}
if(!is.null(stats_tables) && nrow(stats_tables$linear_models) > 0){
  DT::datatable(
    stats_tables$linear_models %>%
      dplyr::select(model, feature, term, term_raw,
                    estimate, std_error, statistic,
                    p_value, p_adj, significant) %>%
      dplyr::mutate(dplyr::across(c(estimate, std_error, statistic), ~ round(., 4)),
                    dplyr::across(c(p_value, p_adj), ~ signif(., 4))),
    filter = "top",
    options = list(pageLength = 25, scrollX = TRUE)
  )
} else {
  cat("_No linear models defined._\n")
}
```

# 7. Session info

```{r session}
sessionInfo()
```
