---
title: "MRManalyzeR — Statistics Report"
output:
  html_document:
    toc: true
    toc_float: true
    self_contained: true
    code_folding: hide
    css: !expr system.file("rmd","IMP_ReportCSS.css", package="MRManalyzeR")
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo        = FALSE,
  warning     = FALSE,
  message     = FALSE,
  fig.align   = "center",
  fig.retina  = 3
)

library(dplyr)
library(tidyr)
library(ggplot2)
```

```{r unpack-params}
if(!exists("report_pars") || is.null(report_pars))
  report_pars = if(exists("mva_pars")) mva_pars else uva_pars

`%||%` = function(a, b) if(is.null(a)) b else a

# -- Sample identification (shared across all sections) --
# One column classifies every row; sample_labels names the value(s) marking
# biological samples (vs blanks / QCs / calibration standards).
# sample_id_head is the per-sample identifier used for hover tooltips and
# heatmap row labels (and the default for pca.label_factor).
sample_type_head = report_pars$sample_type_head %||% "Sample_type"
sample_labels    = report_pars$sample_labels    %||% "Sample"
sample_id_head   = report_pars$sample_id_head   %||% "Sample_ID"

# Stats blocks
# Each section is a block with a section-level `enabled:` toggle and an
# `entries:` list; .section_entries() honours the toggle and returns the
# entries (or list() when disabled/absent).
comparisons   = .section_entries(report_pars$comparisons,   "comparisons")
correlations  = .section_entries(report_pars$correlations,  "correlations")
linear_models = .section_entries(report_pars$linear_models, "linear_models")

p_adjust      = report_pars$p_adjust      %||% "BH"
sig_threshold = report_pars$sig_threshold %||% 0.05

# Plot config
fig_scale       = as.numeric(report_pars$fig_scale %||% 1)
boxplot_faceted = isTRUE(report_pars$boxplot$faceted %||% TRUE)
pca_include     = isTRUE(report_pars$pca$enabled   %||% TRUE)
pca_label       = report_pars$pca$label            %||% "none"
pca_label_factor = report_pars$pca$label_factor    %||% sample_id_head
pca_ellipse     = report_pars$pca$ellipse          %||% "group"
pca_ellipse_conf = as.numeric(report_pars$pca$ellipse_confidence %||% 0.95)

# PCA preprocessing (with sensible defaults)
pca_pp = report_pars$pca$preprocess %||% list()
pca_impute        = pca_pp$impute      %||% "min"
pca_impute_frac   = as.numeric(pca_pp$impute_frac %||% 0.5)
pca_transform     = pca_pp$transform   %||% "log2"
pca_center        = isTRUE(pca_pp$center %||% TRUE)
pca_scale         = isTRUE(pca_pp$scale  %||% TRUE)
pca_feat_na_max   = as.numeric(pca_pp$feat_na_max   %||% 1)
pca_sample_na_max = as.numeric(pca_pp$sample_na_max %||% 1)

# Component pair to plot (default PC1 vs PC2). Length-2 vector of integers.
pca_components    = report_pars$pca$components %||% c(1, 2)
pca_pc_x          = as.integer(pca_components[1])
pca_pc_y          = as.integer(pca_components[length(pca_components)])

# Optional loadings plot below scores; coloured by a feature_metadata column.
pca_loadings_show = isTRUE(report_pars$pca$loadings$enabled %||% TRUE)
pca_loadings_color_by = report_pars$pca$loadings$color_by %||% NULL

# Optional extra subset for the GLOBAL PCA tab (multi-key, AND'd together).
# This is on top of the biological_filter, e.g. pca.subset: { Lobe: Left }
# to inspect only the left-lobe samples in the global PCA.
pca_global_subset = report_pars$pca$subset %||% NULL

volcano_include   = isTRUE(report_pars$volcano$enabled %||% TRUE)
volcano_top_n     = as.integer(report_pars$volcano$top_n_label %||% 10)
# Volcano y-axis: use raw or BH-adjusted p (default = unadjusted, per request)
volcano_use_adj   = isTRUE(report_pars$volcano$use_adjusted %||% FALSE)

# Heatmap (z-score, samples x features). Shared render settings live at the
# heatmap: root and are INHERITED by the general / top_changing / split_by_class
# sub-blocks; any can be OVERRIDDEN per block (resolved via .resolve_hm below).
heatmap_include            = isTRUE(report_pars$heatmap$enabled %||% TRUE)
heatmap_color_samples_by   = report_pars$heatmap$color_samples_by  %||%
                             report_pars$global_summary_factor %||% sample_type_head
heatmap_group_features_by  = report_pars$heatmap$group_features_by %||% NULL
heatmap_transform          = report_pars$heatmap$transform %||% "log2"
heatmap_scale              = report_pars$heatmap$scale     %||% "z"
heatmap_cluster_rows       = isTRUE(report_pars$heatmap$cluster_rows %||% FALSE)
heatmap_cluster_within     = isTRUE(report_pars$heatmap$cluster_within_groups %||% TRUE)
# Orientation: which axis carries samples. "samples_y" (default) = samples on
# rows / y-axis, features on columns / x-axis. "samples_x" transposes. "auto"
# puts the larger dimension on the y-axis.
heatmap_orientation        = tolower(report_pars$heatmap$orientation %||% "samples_y")
# Feature-label density: show | small | hide. "small" shrinks the feature-axis
# labels so a large panel stays readable via the class colour annotation bar;
# "hide" drops them entirely.
heatmap_feature_labels     = tolower(report_pars$heatmap$feature_labels %||% "small")
# Root default for the 4.6 comparison-tab ranking metric (top_changing may override).
heatmap_rank_by            = tolower(report_pars$heatmap$rank_by %||% "p_adj")

# Validators: coerce out-of-range values back to the safe default.
.chk_scope  = function(x) if(x %in% c("global","comparisons","all")) x else "all"
.chk_orient = function(x) if(x %in% c("samples_y","samples_x","auto")) x else "samples_y"
.chk_labels = function(x) if(x %in% c("show","small","hide")) x else "small"
heatmap_orientation    = .chk_orient(heatmap_orientation)
heatmap_feature_labels = .chk_labels(heatmap_feature_labels)

# Resolve a sub-block's render settings against the root defaults — the full
# settings list consumed by .draw_heatmap(). Every setting is overridable per
# block (block$x %||% root_x).
.resolve_hm = function(block){
  block = block %||% list()
  list(
    color_samples_by  = block$color_samples_by  %||% heatmap_color_samples_by,
    group_features_by = block$group_features_by %||% heatmap_group_features_by,
    transform         = tolower(block$transform %||% heatmap_transform),
    scale             = tolower(block$scale     %||% heatmap_scale),
    cluster_rows      = isTRUE(block$cluster_rows          %||% heatmap_cluster_rows),
    cluster_within    = isTRUE(block$cluster_within_groups %||% heatmap_cluster_within),
    orientation       = .chk_orient(tolower(block$orientation    %||% heatmap_orientation)),
    feature_labels    = .chk_labels(tolower(block$feature_labels %||% heatmap_feature_labels))
  )
}

# 4.5 general — all features retained.
heatmap_general       = report_pars$heatmap$general %||% list()
heatmap_general_on    = isTRUE(heatmap_general$enabled %||% TRUE)
heatmap_general_scope = .chk_scope(tolower(heatmap_general$scope %||% "all"))
heatmap_general_cfg   = .resolve_hm(heatmap_general)

# 4.6 top-changing — top-N changing features per class. Comparison tabs rank by
# `rank_by` (block %||% root); the Global tab ranks by separation across
# color_samples_by (no single 2-group contrast exists globally).
heatmap_topchg        = report_pars$heatmap$top_changing %||% list()
heatmap_topchg_on     = isTRUE(heatmap_topchg$enabled %||% FALSE)
heatmap_topchg_scope  = .chk_scope(tolower(heatmap_topchg$scope %||% "comparisons"))
heatmap_topchg_n      = as.integer(heatmap_topchg$n_per_class %||% 10)
heatmap_topchg_rank   = tolower(heatmap_topchg$rank_by %||% heatmap_rank_by)
heatmap_topchg_cfg    = .resolve_hm(heatmap_topchg)

# 4.7 by class — one heatmap per subclass_col value, or one COMBINED heatmap
# across the selected values (multi_class). `subclass_values` focuses specific
# values (empty/NULL = all).
heatmap_split         = report_pars$heatmap$split_by_class %||% list()
heatmap_split_on      = isTRUE(heatmap_split$enabled %||% FALSE)
heatmap_split_scope   = .chk_scope(tolower(heatmap_split$scope %||% "global"))
heatmap_subclass_col  = heatmap_split$subclass_col %||% NULL
heatmap_subclass_vals = heatmap_split$subclass_values %||% NULL
heatmap_split_multi   = tolower(heatmap_split$multi_class %||% "combined")
if(!heatmap_split_multi %in% c("combined","tabs")) heatmap_split_multi = "combined"
heatmap_split_cfg     = .resolve_hm(heatmap_split)

# Biological-sample filter (drops blanks/QCs from global summary + global PCA).
# Defaults to sample_type_head ∈ sample_labels. Set biological_filter.head: ~ to disable.
bio_filter        = report_pars$biological_filter %||%
                    list(head = sample_type_head, levels = sample_labels)
bio_filter_head   = bio_filter$head
bio_filter_levels = bio_filter$levels

# General per-compound boxplot/barplot (independent of comparisons)
boxplot_enabled = isTRUE(report_pars$boxplot$enabled %||% FALSE)
boxplot_factors = report_pars$boxplot$factors %||%
                   c(report_pars$boxplot_factor1, report_pars$boxplot_factor2)
boxplot_factors = boxplot_factors[!is.null(boxplot_factors) & nzchar(boxplot_factors)]
boxplot_display = report_pars$boxplot$display %||% "tabs"      # combined | tabs
boxplot_levels  = report_pars$boxplot$levels %||% sample_labels
boxplot_head    = report_pars$boxplot$head   %||% sample_type_head

# Plot type — applies to BOTH section 4.2 (faceted across comparisons) and 4.3
# (general by-factor). Values: "box" | "bar" | "both". When "both", boxplot
# renders first then barplot underneath. Error bars on barplots controlled by
# `boxplot.error_bar`: "SD" | "SE" | "none".
plot_type      = tolower(report_pars$boxplot$plot_type %||%
                          report_pars$plot_type %||% "box")
if(!plot_type %in% c("box", "bar", "both")) plot_type = "box"
plot_error_bar = toupper(report_pars$boxplot$error_bar %||%
                          report_pars$error_bar %||% "SD")
if(!plot_error_bar %in% c("SD", "SE", "NONE")) plot_error_bar = "SD"

# Per-comparison boxplot toggle (faceted across comparisons)
plot_boxplot_per_comparison = isTRUE(report_pars$plot_boxplot_per_compound %||% TRUE)
yaxis_quant    = report_pars$yaxis_quant    %||% 0.95
yaxis_scalar = as.numeric(report_pars$yaxis_scalar %||% 1.3)

# Sample-level summary bar plots; default group/colour by the sample-type classifier.
global_summary_factor = report_pars$global_summary_factor %||% sample_type_head

units = pmp_params$units %||% pmp_params$datatype %||% project_paths$datatype %||% ""

# Y-value label for point tooltips: the datatype units (ng/mL, Response, …),
# or "value" when unset. Coerced to a scalar (datatype may arrive as a list).
value_label = local({
  u = as.character(units)[1]
  if(!is.na(u) && nzchar(u)) u else "value"
})

# Ion ratios (enzyme activity surrogates etc.)
ion_ratios_list    = ((report_pars$ion_ratios %||% list())$ratios) %||% list()
ion_ratios_enabled = length(ion_ratios_list) > 0 &&
                     isTRUE((report_pars$ion_ratios %||% list())$enabled %||% TRUE)

# Interactive plotting (plotly hover tooltips). Default ON. Set
# `stats_report.interactive_plots: False` in YAML to fall back to static png.
interactive_plots = isTRUE(report_pars$interactive_plots %||% TRUE) &&
                    requireNamespace("plotly", quietly = TRUE)

# Render a ggplot — interactive (plotly with `text` tooltip) when
# interactive_plots is TRUE, otherwise standard static print. Use in place of
# print(p) inside results="asis" chunks.
#
# Why the explicit dependency dance: in a `results="asis"` chunk that
# interleaves cat()'d HTML with widgets, `print(htmltools::tagList(w))` does
# NOT reliably register the plotly.js / htmlwidgets.js dependencies into the
# document <head>, so widgets render blank. Building tags explicitly and
# calling knit_meta_add() forces knitr to inject the deps once per render.
.show_plot = function(p, lock_aspect = FALSE, legend_below = FALSE,
                      max_width = NULL, pixel_height = NULL){
  if(interactive_plots){
    w = plotly::ggplotly(p, tooltip = "text", height = pixel_height)
    if(isTRUE(lock_aspect))
      w = plotly::layout(w,
                         yaxis = list(scaleanchor = "x", scaleratio = 1))
    if(isTRUE(legend_below)){
      w = plotly::layout(w,
                         margin = list(b = 170, t = 40),
                         legend = list(orientation = "h",
                                       x = 0.5, xanchor = "center",
                                       y = -0.40, yanchor = "top"))
      # Continuous colorbars live on each trace's `marker.colorbar`, not in
      # the layout legend. Walk the figure data and modify ONLY traces that
      # already have a colorbar (discrete-color traces don't, so they're
      # untouched — avoids the spurious-colorbar bug from earlier).
      for(.i in seq_along(w$x$data)){
        .mk = w$x$data[[.i]]$marker
        if(isTRUE(.mk$showscale) || !is.null(.mk$colorscale)){
          if(is.null(w$x$data[[.i]]$marker$colorbar))
            w$x$data[[.i]]$marker$colorbar = list()
          w$x$data[[.i]]$marker$colorbar$orientation = "h"
          w$x$data[[.i]]$marker$colorbar$x          = 0.5
          w$x$data[[.i]]$marker$colorbar$xanchor    = "center"
          w$x$data[[.i]]$marker$colorbar$y          = -0.30
          w$x$data[[.i]]$marker$colorbar$yanchor    = "top"
          w$x$data[[.i]]$marker$colorbar$len        = 0.7
          w$x$data[[.i]]$marker$colorbar$thickness  = 14
        }
      }
    }
    w = plotly::config(w, displaylogo = FALSE,
                       modeBarButtonsToRemove = c("lasso2d", "select2d"))
    tags = htmltools::as.tags(w, standalone = FALSE)
    deps = htmltools::findDependencies(tags)
    if(length(deps)) knitr::knit_meta_add(list(deps))
    html_str = as.character(tags)
    if(!is.null(max_width))
      html_str = sprintf('<div style="max-width:%s;margin:0 auto;">%s</div>',
                         max_width, html_str)
    cat(html_str)
  } else {
    print(p)
  }
}

# Pre-built theme reused across per-compound loops
boxplot_theme = ggplot2::theme_classic() +
  ggplot2::theme(
    axis.text.x     = ggplot2::element_text(angle = 90, vjust = 0.5, hjust = 1),
    legend.position = "none"
  )

# Add boxplot layers (used when plot_type is "box" or in the "both" loop).
# Text tooltip is bound only on the jitter layer (where you actually want
# per-point hover) — putting `text` on the parent ggplot makes ggplotly()
# render one trace per data row across ALL layers, which produces the
# horizontal-line-through-the-bar artefact in the "bar" layer.
.add_box_layers = function(p){
  p +
    ggplot2::geom_boxplot(alpha = 0.3, outlier.shape = NA) +
    ggplot2::geom_jitter(ggplot2::aes(text = .tooltip),
                         width = 0.15, alpha = 0.7, size = 1.4)
}

# Per-point hover for the boxplot/jitter layers: sample id, y-value (labelled
# with the datatype units — Response / ng/mL / etc.), and the primary study
# factor. `%.3g` = 3 sig figs (report house format); sprintf recycles the
# scalar labels across the vector args, so it fills a whole plot's points.
.point_tooltip = function(sid, gsf_val, yval, ylab = value_label){
  sprintf("%s: %s<br>%s: %.3g<br>%s: %s",
          sample_id_head, sid, ylab, yval, global_summary_factor, gsf_val)
}

# Add barplot layers with SD/SE error bars (used when plot_type is "bar" or
# in the "both" loop). `error_bar` ∈ {"SD","SE","NONE"}. Jitter intentionally
# omitted on bars: the boxplot view next to it already shows individual
# points, and ggplotly() renders per-point jitter as horizontal segments
# under the bar (which is what the dense "lines through the bar" effect was).
.add_bar_layers = function(p, error_bar){
  err_fn = switch(error_bar,
    SD   = function(x){ m = mean(x, na.rm = TRUE); s = stats::sd(x, na.rm = TRUE); c(y = m, ymin = m - s, ymax = m + s) },
    SE   = function(x){ m = mean(x, na.rm = TRUE); n = sum(!is.na(x)); s = stats::sd(x, na.rm = TRUE) / sqrt(max(1, n)); c(y = m, ymin = m - s, ymax = m + s) },
    NONE = NULL
  )
  # ggplotly's translation of geom-width parameters is not 1:1 between bar
  # and errorbar — empirically the errorbar cap renders much wider than the
  # nominal width. To get a cap that's clearly narrower than the bar fill
  # (~1/5 the bar width visually), use a very small absolute width here.
  p = p + ggplot2::stat_summary(geom = "bar", fun = "mean", alpha = 0.4,
                                colour = "black", width = 0.7)
  if(!is.null(err_fn))
    p = p + ggplot2::stat_summary(geom = "errorbar", fun.data = err_fn,
                                  width = 0.05, linewidth = 0.4)
  p
}

# Render one or two plots based on plot_type. Returns a list with named
# entries among "box" / "bar" — caller iterates and prints each.
.compound_plots = function(p_base, plot_type, error_bar){
  out = list()
  if(plot_type %in% c("box", "both")) out$box = .add_box_layers(p_base)
  if(plot_type %in% c("bar", "both")) out$bar = .add_bar_layers(p_base, error_bar)
  out
}
```

# 1. Aim

Statistical analysis of processed `r units` data. Pre-defined comparisons,
correlations, and linear models from the YAML config; multivariate (PCA +
heatmap) inspection per comparison.

> **All per-feature numerical results — comparisons p-values, correlations,
> linear-model coefficients — are written to the stats xlsx (file path
> ends with `_stats.xlsx`) saved alongside this HTML report. Open it in
> Excel to filter / sort the full tables.** This report shows only the
> visualisations and the volcano significant-feature shortlist; the
> long-form tables are intentionally not duplicated here.

**Missing-value policy** (which sections impute and which don't):

| Section | NA handling |
|---|---|
| Comparisons (t-test / Wilcoxon / ANOVA / Kruskal / Tukey) | **No imputation** — raw values used; NAs dropped per test by `t.test()` / `wilcox.test()` / `aov()`. |
| Boxplots (per-compound + by-factor) | **No imputation** — raw values plotted; NAs simply absent from the boxes/jitter. |
| Correlations | **No imputation** — `pairwise.complete.obs` (each pair uses its own complete cases). |
| Linear models | `lm()` drops rows with NA in the feature or any predictor (per feature). |
| PCA (global + per-comparison) | Imputed per `pca.preprocess.impute` (default in this config: `r pca_impute``r if(pca_impute=="frac_min") sprintf(" with frac=%g", pca_impute_frac) else ""`), then transform → centre → scale. |
| Heatmap | After `r heatmap_transform` transform and per-feature z-scoring, residual NAs are set to 0 (= per-feature mean in z-units), then capped at ±2.5. |
| Volcano | Inherits the comparisons table, so no imputation in the underlying p-values; rows missing log2FC or p are dropped from the plot. |

# 2. Data import

```{r import}
importData     = combined_datamatrices
metadata       = importData$sample_meta
data_matrix    = importData$data
compoundInfo   = importData$variable_meta
compound_names = compoundInfo$Compound

# Biological-sample subset: dropped from sections that should not see blanks/QCs
# (global summary, global PCA). Per-comparison sections do their own subsetting
# via subset_dataset(), so they're unaffected.
if(!is.null(bio_filter_head) && bio_filter_head %in% colnames(metadata)){
  bio_de = subset_dataset(importData,
                     conditions = setNames(list(bio_filter_levels),
                                           bio_filter_head))
  bio_meta   = bio_de$sample_meta
  bio_matrix = bio_de$data
} else {
  bio_de     = importData
  bio_meta   = metadata
  bio_matrix = data_matrix
}

# Fail loud here rather than let downstream sections silently degrade:
# smeta[[factor_name]] on a missing column returns NULL (no error), which
# would just produce an uninformative single-group summary bar/heatmap
# instead of surfacing the YAML typo.
if(!global_summary_factor %in% colnames(metadata))
  stop(sprintf(
    "[stats_report] global_summary_factor='%s' is not a column in sample_metadata. Available columns: %s",
    global_summary_factor, paste(colnames(metadata), collapse = ", ")))

DT::datatable(
  metadata %>%
    dplyr::group_by(.data[[sample_type_head]]) %>%
    dplyr::summarize(`Number of samples` = dplyr::n(), .groups = "drop")
)
```

`r if(!is.null(bio_filter_head)) sprintf("_Biological-sample filter applied to global summary + global PCA: \x60%s\x60 ∈ {%s} → %d samples (of %d)._", bio_filter_head, paste(bio_filter_levels, collapse=", "), nrow(bio_meta), nrow(metadata))`

# 3. Global summary {.tabset}

Mean intensity per compound (biological subset: `r nrow(bio_meta)` of
`r nrow(metadata)` samples), with `r global_summary_factor` levels dodged
side-by-side. Compounds in TargetLynx import order. Plot height auto-scales
with the number of features.

```{r global-helpers, include=FALSE}
# Shared qualitative palette so a categorical variable (e.g. Enzymatic_pathway)
# gets IDENTICAL colours in the heatmap feature annotation and the PCA loadings
# bars. Keyed on sorted levels, so a given level maps to the same colour
# regardless of the order features happen to appear in each plot.
.qual_pal = function(n){
  if(n < 1) return(character(0))
  if(requireNamespace("RColorBrewer", quietly = TRUE) && n >= 3 && n <= 9)
    return(RColorBrewer::brewer.pal(n, "Set1"))
  if(requireNamespace("RColorBrewer", quietly = TRUE) && n > 9)
    return(grDevices::colorRampPalette(RColorBrewer::brewer.pal(9, "Set1"))(n))
  grDevices::hcl.colors(max(n, 2), palette = "Dark 3")[seq_len(n)]   # n == 1 or 2
}
# Named colour vector for the sorted levels of x. NA / "NA" -> grey; a
# 2-level variable uses the fixed blue/red pair (matches binary annotations).
.group_fill_colors = function(x){
  x  = as.character(x)
  lv = sort(unique(x[!is.na(x) & x != "NA"]))
  m  = if(length(lv) == 0)      character(0)
       else if(length(lv) == 2) stats::setNames(c("#377eb8", "#e41a1c"), lv)
       else                     stats::setNames(.qual_pal(length(lv)), lv)
  if(any(is.na(x) | x == "NA")) m = c(m, "NA" = "grey70")
  m
}

.feature_bar = function(d, smeta, factor_name, scale, units){
  feats = colnames(d)
  grp   = smeta[[factor_name]]

  rows = lapply(feats, function(f){
    by_grp = tapply(d[[f]], grp, mean, na.rm = TRUE)
    data.frame(feature = f,
               .grp    = names(by_grp),
               value   = unname(by_grp),
               stringsAsFactors = FALSE)
  })
  m = dplyr::bind_rows(rows)

  if(scale == "log2") m$value = suppressWarnings(log2(m$value + 1))
  if(scale == "sqrt") m$value = suppressWarnings(sqrt(pmax(m$value, 0)))

  # Reverse so first imported feature sits at the top of the y-axis
  m$feature = factor(m$feature, levels = rev(feats))

  xlab = switch(scale,
                raw  = units,
                log2 = paste0("log2(", units, " + 1)"),
                sqrt = paste0("sqrt(", units, ")"))

  ggplot2::ggplot(m, ggplot2::aes(value, feature, fill = .grp)) +
    ggplot2::geom_col(position = ggplot2::position_dodge(width = 0.8),
                      width = 0.7) +
    ggplot2::scale_fill_brewer(palette = "Set1", na.value = "grey70") +
    ggplot2::labs(x = xlab, y = "Compound", fill = factor_name) +
    ggplot2::theme_classic(base_size = 11) +
    ggplot2::theme(
      axis.text.y        = ggplot2::element_text(size = 6),
      legend.position    = "bottom",
      panel.grid.major.x = ggplot2::element_line(color = "grey90", linewidth = 0.3)
    )
}
```

## 3.1 Raw (`r units`) {-}

```{r global-feature-raw, fig.width=7*fig_scale, fig.height=max(8, ncol(bio_matrix)*0.18)*fig_scale}
print(.feature_bar(bio_matrix, bio_meta, global_summary_factor, "raw", units))
```

## 3.2 log2 (`r units`) {-}

```{r global-feature-log2, fig.width=7*fig_scale, fig.height=max(8, ncol(bio_matrix)*0.18)*fig_scale}
print(.feature_bar(bio_matrix, bio_meta, global_summary_factor, "log2", units))
```

# 4. Comparisons

```{r comparisons-summary, results="asis"}
if(length(comparisons) == 0){
  cat("_No comparisons defined in YAML._\n")
} else {
  cat("Configured comparisons:\n\n")
  for(comp in comparisons){
    sub_str = if(is.null(comp$subset)) "(all samples)" else
      paste(sprintf("%s = %s", names(comp$subset), unlist(comp$subset)),
            collapse = ", ")
    cat(sprintf("- **%s** — subset: `%s`; compare `%s`: %s\n",
                comp$name, sub_str, comp$compare$factor,
                paste(comp$compare$levels, collapse = " vs ")))
  }
}
```

**Methods.** For each comparison the test method is auto-selected by group
count: 2 levels → t-test + Wilcoxon; ≥3 levels → ANOVA + Kruskal-Wallis,
plus Tukey HSD post-hoc per pair. P-values are Benjamini-Hochberg adjusted
across **all** comparisons × methods × features.

> **Full p-value tables (per comparison + per method, all features) are
> in the `stats` tab of the stats xlsx** written by `run_MRManalyzeR()`
> alongside the rendered HTML reports. The same xlsx also carries the
> `correlations` and `linear_models` tabs.

## 4.1 Per-compound `r switch(plot_type, box="boxplot", bar="barplot", both="boxplot + barplot")` (faceted across comparisons) {-}

`r if(!plot_boxplot_per_comparison) "_Skipped — set_ \x60plot_boxplot_per_compound: True\x60 _in YAML to enable._"`

```{r comp-boxplots, results="asis", eval=plot_boxplot_per_comparison && length(comparisons) > 0, fig.width=10*fig_scale, fig.height=6*fig_scale}
# Build one long data frame across all comparisons, then facet.
# Combine user `subset:` with the comparison level restriction in one
# subset_dataset() call (avoid post-hoc DE mutation which trips SE's assay check).
long_rows = list()
for(comp in comparisons){
  cond = comp$subset
  cond = if(is.null(cond)) list() else as.list(cond)
  cond[[comp$compare$factor]] = comp$compare$levels
  de_sub = subset_dataset(importData, conditions = cond)
  if(nrow(de_sub$sample_meta) == 0) next

  d = de_sub$data
  .sm = de_sub$sample_meta
  d$.sample_id  = if(sample_id_head %in% colnames(.sm))
                    as.character(.sm[[sample_id_head]]) else rownames(d)
  d$.group      = factor(de_sub$sample_meta[[comp$compare$factor]],
                         levels = comp$compare$levels)
  d$.comparison = comp$name
  d$.gsf        = if(global_summary_factor %in% colnames(.sm))
                    as.character(.sm[[global_summary_factor]]) else NA_character_
  long_rows[[length(long_rows) + 1]] = d
}
long = dplyr::bind_rows(long_rows)

make_safe_id = function(x) gsub("[^A-Za-z0-9_\\-]", "_", x)
safe_ids = vapply(compound_names, make_safe_id, character(1))

is_detected = vapply(compound_names, function(cmp){
  cc = which(colnames(long) == cmp)
  length(cc) > 0 && !all(is.na(unlist(long[, cc])))
}, logical(1))
detected_compounds = compound_names[is_detected]
detected_ids       = safe_ids[is_detected]

# Alphabetise dropdown order (T3.5-7); plot blocks below iterate in import order.
.alpha_idx        = order(detected_compounds)
detected_alpha    = detected_compounds[.alpha_idx]
detected_ids_alpha = detected_ids[.alpha_idx]

cat('
<label for="boxplot_select" style="font-weight:600;">Select compound:</label>
<select id="boxplot_select" style="margin: 0 0 12px 8px; padding: 4px 6px;"
        onchange="showBoxplot(this.value)">
')
for(i in seq_along(detected_alpha)){
  sel = if(i == 1) " selected" else ""
  cat(sprintf('<option value="%s"%s>%s</option>\n',
              detected_ids_alpha[i], sel, detected_alpha[i]))
}
cat('</select>
<script>
function showBoxplot(id){
  var plots = document.querySelectorAll(".cmp_boxplot_block");
  for (var i=0; i<plots.length; i++) { plots[i].style.display = "none"; }
  var el = document.getElementById("cmp_boxplot_" + id);
  if (el) {
    el.style.display = "block";
    if (window.Plotly){
      var pls = el.querySelectorAll(".plotly-graph-div");
      for (var j=0; j<pls.length; j++) Plotly.Plots.resize(pls[j]);
    }
  }
}
</script>
')

for(i in seq_along(detected_compounds)){
  compound = detected_compounds[i]
  cid      = detected_ids[i]
  axis_limit = quantile(long[[compound]], prob = yaxis_quant, na.rm = TRUE) * yaxis_scalar

  long$.tooltip = .point_tooltip(long$.sample_id, long$.gsf, long[[compound]])

  p_base = ggplot2::ggplot(long, ggplot2::aes(x = .group, y = .data[[compound]],
                                              fill = .group)) +
            ggplot2::coord_cartesian(ylim = c(0, axis_limit)) +
            ggplot2::labs(x = NULL, y = units, title = compound) +
            ggplot2::facet_wrap(~ .comparison, scales = "free_x", nrow = 1) +
            boxplot_theme
  plots = .compound_plots(p_base, plot_type, plot_error_bar)

  # Show alphabetically-first detected compound by default (matches dropdown)
  display = if(detected_ids[i] == detected_ids_alpha[1]) "block" else "none"
  cat(sprintf('<div class="cmp_boxplot_block" id="cmp_boxplot_%s" style="display:%s;">\n',
              cid, display))
  # When plot_type is "both" we get two separate figures (box first, bar second)
  for(pname in names(plots)){
    if(plot_type == "both")
      cat(sprintf('<div style="margin-top:10px;"><b>%s</b></div>\n',
                  if(pname == "box") "Boxplot" else "Barplot"))
    .show_plot(plots[[pname]])
  }
  cat("</div>\n")
}
```

## 4.2 Per-compound `r switch(plot_type, box="boxplot", bar="barplot", both="boxplot + barplot")` by factor(s) {-}

`r if(!boxplot_enabled) "_Skipped — set_ \x60boxplot.enabled: True\x60 _in YAML to enable._"`

`r if(boxplot_enabled) sprintf("Per-compound boxplot grouped by_ \x60%s\x60_. Display mode:_ \x60%s\x60_._", paste(boxplot_factors, collapse=", "), boxplot_display)`

```{r factor-boxplots, results="asis", eval=boxplot_enabled && length(boxplot_factors) > 0, fig.width=10*fig_scale, fig.height=6*fig_scale}
# Filter to requested sample types
keep = metadata[[boxplot_head]] %in% boxplot_levels
plot_meta = metadata[keep, , drop = FALSE]
plot_data = data_matrix[keep, , drop = FALSE]

# Primary study factor per row (for point tooltips), aligned to plot_data rows.
gsf_fb = if(global_summary_factor %in% colnames(plot_meta))
           as.character(plot_meta[[global_summary_factor]]) else NA_character_

missing_bf = setdiff(boxplot_factors, colnames(plot_meta))
if(length(missing_bf))
  stop(sprintf(
    "[stats_report] boxplot.factors contains column(s) not in sample_metadata: %s. Available columns: %s",
    paste(missing_bf, collapse = ", "), paste(colnames(plot_meta), collapse = ", ")))

# Detect compounds with any data
is_detected_fb = vapply(compound_names, function(cmp){
  cc = which(colnames(plot_data) == cmp)
  length(cc) > 0 && !all(is.na(unlist(plot_data[, cc])))
}, logical(1))
detected_fb     = compound_names[is_detected_fb]
make_safe_id_fb = function(x) gsub("[^A-Za-z0-9_\\-]", "_", x)
detected_ids_fb = vapply(detected_fb, make_safe_id_fb, character(1))

# Build per-tab grouping vectors
tab_groupings = list()
if(boxplot_display == "combined" || length(boxplot_factors) == 1){
  lab = paste(boxplot_factors, collapse = " × ")
  tab_groupings[[lab]] = do.call(paste, c(plot_meta[boxplot_factors], sep = "_"))
} else {
  for(f in boxplot_factors)
    tab_groupings[[f]] = as.character(plot_meta[[f]])
  if(length(boxplot_factors) >= 2){
    lab = paste(boxplot_factors, collapse = " × ")
    tab_groupings[[lab]] = do.call(paste, c(plot_meta[boxplot_factors], sep = "_"))
  }
}

cat('\n### {.tabset .tabset-fade}\n\n')

for(tab_name in names(tab_groupings)){
  cat(sprintf('\n#### %s {-}\n\n', tab_name))

  grouping = tab_groupings[[tab_name]]
  tab_safe = make_safe_id_fb(tab_name)

  # Alphabetised dropdown (T3.5-7); plot blocks below in import order.
  alpha_idx_fb       = order(detected_fb)
  detected_alpha_fb  = detected_fb[alpha_idx_fb]
  detected_ids_alpha_fb = detected_ids_fb[alpha_idx_fb]

  cat(sprintf('
<label for="fbox_select_%s" style="font-weight:600;">Select compound:</label>
<select id="fbox_select_%s" style="margin: 0 0 12px 8px; padding: 4px 6px;"
        onchange="showFactorBoxplot_%s(this.value)">
', tab_safe, tab_safe, tab_safe))
  for(i in seq_along(detected_alpha_fb)){
    sel = if(i == 1) " selected" else ""
    cat(sprintf('<option value="%s"%s>%s</option>\n',
                detected_ids_alpha_fb[i], sel, detected_alpha_fb[i]))
  }
  cat(sprintf('</select>
<script>
function showFactorBoxplot_%s(id){
  var plots = document.querySelectorAll(".fbox_block_%s");
  for (var i=0; i<plots.length; i++) { plots[i].style.display = "none"; }
  var el = document.getElementById("fbox_%s_" + id);
  if (el) {
    el.style.display = "block";
    if (window.Plotly){
      var pls = el.querySelectorAll(".plotly-graph-div");
      for (var j=0; j<pls.length; j++) Plotly.Plots.resize(pls[j]);
    }
  }
}
</script>
', tab_safe, tab_safe, tab_safe))

  for(i in seq_along(detected_fb)){
    compound = detected_fb[i]
    cid      = detected_ids_fb[i]
    axis_limit = quantile(plot_data[[compound]], prob = yaxis_quant, na.rm = TRUE) * yaxis_scalar
    sid_fb = if(sample_id_head %in% colnames(plot_meta))
               as.character(plot_meta[[sample_id_head]]) else rownames(plot_data)
    df_plot = data.frame(group     = grouping,
                         value     = plot_data[[compound]],
                         sample_id = sid_fb,
                         .tooltip  = .point_tooltip(sid_fb, gsf_fb, plot_data[[compound]]),
                         stringsAsFactors = FALSE)

    p_base = ggplot2::ggplot(df_plot,
                             ggplot2::aes(group, value, fill = group)) +
              ggplot2::coord_cartesian(ylim = c(0, axis_limit)) +
              ggplot2::scale_fill_brewer(palette = "Set1", na.value = "grey60") +
              ggplot2::labs(x = NULL, y = units, title = compound) +
              boxplot_theme
    plots = .compound_plots(p_base, plot_type, plot_error_bar)

    # Show alphabetically-first compound by default (matches dropdown)
    display = if(detected_ids_fb[i] == detected_ids_alpha_fb[1]) "block" else "none"
    cat(sprintf('<div class="fbox_block_%s" id="fbox_%s_%s" style="display:%s;">\n',
                tab_safe, tab_safe, cid, display))
    for(pname in names(plots)){
      if(plot_type == "both")
        cat(sprintf('<div style="margin-top:10px;"><b>%s</b></div>\n',
                    if(pname == "box") "Boxplot" else "Barplot"))
      .show_plot(plots[[pname]])
    }
    cat("</div>\n")
  }
}
```

## 4.3 PCA per comparison

`r if(!pca_include) "_Skipped — set_ \x60pca.enabled: True\x60 _in YAML to enable._"`

For each comparison: filter samples by the comparison's `subset:`, run PCA on
that slice, colour points by the comparison's `compare.factor`. The first
tab is a **global** PCA across the biological subset, coloured by
`r global_summary_factor`.

**Components plotted:** PC`r pca_pc_x` × PC`r pca_pc_y`
(set via `stats_report.pca.components: [x, y]`).

**Preprocessing pipeline** (configured in `stats_report.pca.preprocess`; per-run actual steps are echoed inside each tab):

| Step | Setting | Value |
|---|---|---|
| Feature filter | `feat_na_max` | `r if(pca_feat_na_max < 1) sprintf("drop features with > %g%% NA", pca_feat_na_max*100) else "none (only all-NA features removed)"` |
| Sample filter | `sample_na_max` | `r if(pca_sample_na_max < 1) sprintf("drop samples with > %g%% NA", pca_sample_na_max*100) else "none"` |
| Imputation | `impute` | `r pca_impute``r if(pca_impute=="frac_min") sprintf(" (frac = %g)", pca_impute_frac)` |
| Transform | `transform` | `r pca_transform` |
| Centre | `center` | `r pca_center` |
| Scale | `scale` | `r pca_scale` |

```{r pca-helpers, include=FALSE}
# Local helper: outlier flag on chosen PCs
.pca_outliers = function(x, y){
  zx = scale(x)[, 1]; zy = scale(y)[, 1]
  abs(zx) > 2.5 | abs(zy) > 2.5
}

# Build the scores ggplot object. Does NOT print — caller decides.
.pca_scores_plot = function(res_pca, smeta, color_factor,
                            pc_x, pc_y, label_mode, label_factor,
                            ellipse_mode, ellipse_conf){
  pr = res_pca$pr
  if(is.null(pr)) return(NULL)

  # Validate / fall back component selection
  pc_max = ncol(pr$x)
  if(pc_x > pc_max || pc_y > pc_max){
    pc_x = 1; pc_y = min(2, pc_max)
  }
  vexp = (pr$sdev^2) / sum(pr$sdev^2) * 100
  pcx_lab = sprintf("PC%d", pc_x); pcy_lab = sprintf("PC%d", pc_y)

  scores = data.frame(
    pcx   = pr$x[, pc_x],
    pcy   = pr$x[, pc_y],
    group = smeta[[color_factor]],
    stringsAsFactors = FALSE
  )
  scores$label = if(!is.null(label_factor) && label_factor %in% colnames(smeta))
                   as.character(smeta[[label_factor]]) else rownames(scores)
  scores$tooltip = sprintf("%s: %s<br>%s: %s<br>PC%d: %.3g<br>PC%d: %.3g",
                           label_factor, scores$label,
                           color_factor, as.character(scores$group),
                           pc_x, scores$pcx, pc_y, scores$pcy)
  is_num = is.numeric(scores$group)

  p = ggplot2::ggplot(scores, ggplot2::aes(pcx, pcy, color = group, text = tooltip)) +
        ggplot2::geom_point(size = 3, alpha = 0.85) +
        ggplot2::labs(x = sprintf("%s (%.1f%%)", pcx_lab, vexp[pc_x]),
                      y = sprintf("%s (%.1f%%)", pcy_lab, vexp[pc_y]),
                      color = color_factor) +
        ggplot2::theme_classic(base_size = 13) +
        ggplot2::theme(
          legend.position   = "bottom",
          legend.box.margin = ggplot2::margin(t = 12),
          axis.title.x      = ggplot2::element_text(margin = ggplot2::margin(t = 6, b = 0)),
          plot.margin       = ggplot2::margin(4, 6, 4, 6)
        )
  p = if(is_num) p + ggplot2::scale_color_viridis_c()
      else       p + ggplot2::scale_color_brewer(palette = "Set1",
                                                 na.value = "grey60")
  if(ellipse_mode == "group" && !is_num && length(unique(scores$group)) > 1)
    p = p + ggplot2::stat_ellipse(ggplot2::aes(group = group),
                                  type = "norm", level = ellipse_conf,
                                  linewidth = 0.4)
  if(label_mode == "all"){
    p = p + ggplot2::geom_text(ggplot2::aes(label = label),
                               size = 2.5, vjust = -1, show.legend = FALSE)
  } else if(label_mode == "outliers"){
    out_idx = .pca_outliers(scores$pcx, scores$pcy)
    if(any(out_idx))
      p = p + ggplot2::geom_text(data = scores[out_idx, ],
                                 ggplot2::aes(label = label),
                                 size = 2.5, vjust = -1, show.legend = FALSE)
  }
  p
}

# Build the loadings ggplot object. Does NOT print.
.pca_loadings_plot = function(res_pca, vmeta, pc_x, pc_y, loadings_color_by){
  pr = res_pca$pr
  if(is.null(pr)) return(NULL)
  pc_max = ncol(pr$x)
  if(pc_x > pc_max || pc_y > pc_max){
    pc_x = 1; pc_y = min(2, pc_max)
  }
  pcx_lab = sprintf("PC%d", pc_x); pcy_lab = sprintf("PC%d", pc_y)
  feats = rownames(pr$rotation)

  has_color = !is.null(loadings_color_by) && loadings_color_by %in% colnames(vmeta)
  color_vec = if(has_color){
    cv = as.character(vmeta[[loadings_color_by]][match(feats, vmeta$Compound)])
    cv[is.na(cv)] = "NA"; cv
  } else {
    rep("(all)", length(feats))
  }
  groups_in_order = unique(color_vec)

  build_pc_df = function(loading_vec, pc_lab){
    ord_feats = character(0)
    for(g in groups_in_order){
      idx_g = which(color_vec == g)
      if(!length(idx_g)) next
      sorted = idx_g[order(loading_vec[idx_g], decreasing = TRUE)]
      ord_feats = c(ord_feats, feats[sorted])
    }
    df = data.frame(
      feature = feats,
      feat_pc = paste0(feats, "__", pc_lab),
      loading = loading_vec,
      .color  = factor(color_vec, levels = groups_in_order),
      PC      = pc_lab,
      stringsAsFactors = FALSE
    )
    list(df = df, order_levels = rev(paste0(ord_feats, "__", pc_lab)))
  }
  ld_x = build_pc_df(pr$rotation[, pc_x], pcx_lab)
  ld_y = build_pc_df(pr$rotation[, pc_y], pcy_lab)
  ld_long = rbind(ld_x$df, ld_y$df)
  combined_levels = c(ld_x$order_levels, ld_y$order_levels)
  ld_long$feat_pc = factor(ld_long$feat_pc, levels = combined_levels)
  ld_long$PC      = factor(ld_long$PC, levels = c(pcx_lab, pcy_lab))

  ggplot2::ggplot(ld_long,
                  ggplot2::aes(loading, feat_pc, fill = .color)) +
        ggplot2::geom_col() +
        ggplot2::geom_vline(xintercept = 0, color = "grey50", linewidth = 0.3) +
        ggplot2::facet_wrap(~ PC, scales = "free", nrow = 1) +
        ggplot2::scale_y_discrete(labels = function(x) sub("__[^_]+$", "", x)) +
        # Same palette as the heatmap feature annotation (shared .group_fill_colors)
        ggplot2::scale_fill_manual(values = .group_fill_colors(color_vec),
                                   na.value = "grey60", drop = FALSE) +
        ggplot2::labs(x = "Loading", y = NULL,
                      fill = loadings_color_by %||% "") +
        ggplot2::theme_classic(base_size = 10) +
        ggplot2::theme(axis.text.y     = ggplot2::element_text(size = 6),
                       strip.text      = ggplot2::element_text(face = "bold"),
                       legend.position = "bottom") +
        ggplot2::guides(fill = ggplot2::guide_legend(nrow = 3, byrow = TRUE))
}

# Print preprocessing trace + the matrix-size note (used in scores tab).
.pca_preprocess_trace = function(res_pca){
  cat("**Preprocessing pipeline applied:**\n\n")
  for(s in res_pca$steps) cat(sprintf("- %s\n", s))
  cat(sprintf("\n_Final matrix: %d samples × %d features._\n\n",
              res_pca$n_samples, res_pca$n_features))
  pr = res_pca$pr
  if(is.null(pr))
    cat(sprintf("_PCA could not be fit (n_samples=%d, n_features=%d)._\n\n",
                res_pca$n_samples, res_pca$n_features))
}
```

```{r comp-pca-compute, include=FALSE, eval=pca_include}
# Pre-compute every PCA (Global + per-comparison) into a list shared by the
# two rendering chunks below. Each entry holds the PCA result, the metadata
# slices, the colour factor, and a tab heading. We compute once and render
# twice (scores at small canvas, loadings at tall canvas) to give the scores
# plot a sensible aspect ratio independent of feature count.
.pca_runs = list()

# Helper: run the PCA pipeline AND realign sample_meta to the samples that
# actually survived `sample_na_max`. Without this, scores ↔ smeta mismatch
# when sample_na_max < 1 drops any rows.
.pca_run_one = function(de){
  M  = as.matrix(de$data)
  sm = de$sample_meta
  # Anchor rownames so we can subset sm after the pipeline drops rows.
  rn = if(!is.null(rownames(M)) && !anyDuplicated(rownames(M))) rownames(M)
       else paste0("s", seq_len(nrow(M)))
  rownames(M) = rn
  rownames(sm) = rn

  res = run_pca(M,
                         impute        = pca_impute,
                         impute_frac   = pca_impute_frac,
                         transform     = pca_transform,
                         center        = pca_center,
                         scale         = pca_scale,
                         feat_na_max   = pca_feat_na_max,
                         sample_na_max = pca_sample_na_max)
  if(!is.null(res$pr))
    sm = sm[rownames(res$pr$x), , drop = FALSE]

  list(res_pca = res, smeta = sm, vmeta = de$variable_meta)
}

# Global tab
pca_global_de = if(length(pca_global_subset)){
  subset_dataset(bio_de, conditions = as.list(pca_global_subset))
} else {
  bio_de
}
global_subset_lbl = if(length(pca_global_subset)){
  sprintf(" — subset: %s",
          paste(sprintf("%s=%s", names(pca_global_subset),
                        unlist(pca_global_subset)), collapse = ", "))
} else { "" }
.gbl = .pca_run_one(pca_global_de)
.pca_runs[["Global"]] = list(
  tab_heading  = sprintf("Global (%d samples; coloured by %s)%s",
                         nrow(.gbl$smeta),
                         global_summary_factor, global_subset_lbl),
  res_pca      = .gbl$res_pca,
  smeta        = .gbl$smeta,
  vmeta        = .gbl$vmeta,
  color_factor = global_summary_factor,
  insufficient = FALSE
)

# Per-comparison tabs
for(comp in comparisons){
  cond = comp$subset
  cond = if(is.null(cond)) list() else as.list(cond)
  cond[[comp$compare$factor]] = comp$compare$levels
  de_sub = subset_dataset(importData, conditions = cond)

  if(nrow(de_sub$data) < 3 || ncol(de_sub$data) < 2){
    .pca_runs[[comp$name]] = list(tab_heading = comp$name, insufficient = TRUE)
    next
  }
  .one = .pca_run_one(de_sub)
  .pca_runs[[comp$name]] = list(
    tab_heading  = comp$name,
    res_pca      = .one$res_pca,
    smeta        = .one$smeta,
    vmeta        = .one$vmeta,
    color_factor = comp$compare$factor,
    insufficient = FALSE
  )
}

# Cap the loadings canvas height by the largest feature count across runs.
.pca_max_feats = max(c(0, vapply(.pca_runs, function(r){
  if(isTRUE(r$insufficient) || is.null(r$res_pca$pr)) 0L
  else nrow(r$res_pca$pr$rotation)
}, integer(1))))
```

### Scores {.tabset .tabset-fade}

```{r comp-pca-scores, results="asis", eval=pca_include, fig.width=4.5*fig_scale, fig.height=4.5*fig_scale, out.width="450px", out.height="auto"}
for(nm in names(.pca_runs)){
  r = .pca_runs[[nm]]
  cat(sprintf("\n\n#### %s {-}\n\n", r$tab_heading))
  if(isTRUE(r$insufficient)){ cat("_Insufficient data._\n\n"); next }
  .pca_preprocess_trace(r$res_pca)
  p_sc = .pca_scores_plot(r$res_pca, r$smeta, r$color_factor,
                          pca_pc_x, pca_pc_y,
                          pca_label, pca_label_factor,
                          pca_ellipse, pca_ellipse_conf)
  if(!is.null(p_sc)){
    .show_plot(p_sc, lock_aspect = TRUE, legend_below = TRUE,
               max_width = "650px", pixel_height = 650)
    cat("\n\n")
  }
}
```

### Loadings {.tabset .tabset-fade}

```{r comp-pca-loadings, results="asis", eval=pca_include && pca_loadings_show, fig.width=10*fig_scale, fig.height=max(7, (if(exists(".pca_max_feats")) .pca_max_feats else 0) * 0.18) * fig_scale}
for(nm in names(.pca_runs)){
  r = .pca_runs[[nm]]
  cat(sprintf("\n\n#### %s {-}\n\n", r$tab_heading))
  if(isTRUE(r$insufficient) || is.null(r$res_pca$pr)){
    cat("_No loadings available._\n\n"); next
  }
  if(!is.null(pca_loadings_color_by) &&
     pca_loadings_color_by %in% colnames(r$vmeta)){
    cat(sprintf("**Loadings** _(colour =_ \x60%s\x60_; bars grouped by colour, sorted within each group)_\n\n",
                pca_loadings_color_by))
  } else {
    cat("**Loadings**\n\n")
  }
  p_ld = .pca_loadings_plot(r$res_pca, r$vmeta, pca_pc_x, pca_pc_y,
                            pca_loadings_color_by)
  if(!is.null(p_ld)){ print(p_ld); cat("\n\n") }
}
```

## 4.4 Volcano per comparison {.tabset .tabset-fade}

`r if(!volcano_include) "_Skipped — set_ \x60volcano.enabled: True\x60 _in YAML to enable._"`

**Methods.** One volcano per 2-group comparison directly from the t-test
rows. For ≥3-group comparisons we plot one volcano per Tukey HSD pair.
X = log2 fold-change (mean_g2 / mean_g1); Y = -log10 of `r if(volcano_use_adj) "BH-adjusted" else "raw"` p
(toggle via `stats_report.volcano.use_adjusted`).
Top `r volcano_top_n` features by significance are labelled.
Significance threshold dashed line at p `r if(volcano_use_adj) "(BH-adjusted)" else "(raw)"` = `r sig_threshold`.

```{r comp-volcano, results="asis", eval=volcano_include && !is.null(stats_tables) && nrow(stats_tables$stats) > 0, fig.width=5.5*fig_scale, fig.height=5*fig_scale, out.width="600px"}
# Volcano sources:
#   2-group comparisons -> use the t_test rows directly (have log2FC + p_adj)
#   3+-group comparisons -> use the tukey post-hoc rows (one volcano per pair)
.volc_p_col = if(volcano_use_adj) "p_adj" else "p_value"
.volc_p_lbl = if(volcano_use_adj) sprintf("BH-adjusted p (%s)", p_adjust) else "raw p"

volc_data = stats_tables$stats %>%
  dplyr::filter(method %in% c("t_test", "tukey"),
                !is.na(log2FC), !is.na(.data[[.volc_p_col]]))

if(nrow(volc_data) == 0){
  cat("_No volcano-eligible rows._\n")
} else {
  for(comp_name in unique(volc_data$comparison)){
    sub = volc_data %>% dplyr::filter(comparison == comp_name)
    pair_groups = if("tukey" %in% sub$method)
      unique(sub %>% dplyr::filter(method == "tukey") %>%
               dplyr::transmute(pair = paste(group1, "vs", group2)) %>%
               dplyr::pull(pair))
    else NA

    for(pair_lbl in (if(is.na(pair_groups[1])) NA else pair_groups)){
      this = if(is.na(pair_lbl)) sub else
        sub %>% dplyr::filter(method == "tukey",
                              paste(group1, "vs", group2) == pair_lbl)

      cat(sprintf("\n### %s%s {-}\n\n", comp_name,
                  if(is.na(pair_lbl)) "" else paste0(" — ", pair_lbl)))

      this$neglog10p = -log10(this[[.volc_p_col]])
      this$sig = ifelse(this[[.volc_p_col]] < sig_threshold, "sig", "ns")
      this$tooltip = sprintf("%s<br>log2FC: %.3g<br>%s: %.3g",
                             this$feature, this$log2FC,
                             .volc_p_lbl, this[[.volc_p_col]])

      # Always label all significant features; if more than top_n_label sig
      # features exist, label only the top_n_label most significant.
      sig_rows = this %>% dplyr::filter(sig == "sig") %>%
                          dplyr::arrange(.data[[.volc_p_col]])
      to_label = if(nrow(sig_rows) <= volcano_top_n) sig_rows
                 else utils::head(sig_rows, volcano_top_n)

      p = ggplot2::ggplot(this, ggplot2::aes(log2FC, neglog10p, color = sig, text = tooltip)) +
            ggplot2::geom_point(alpha = 0.7, size = 2) +
            ggplot2::scale_color_manual(values = c(sig = "#d62728", ns = "grey60")) +
            ggplot2::geom_hline(yintercept = -log10(sig_threshold),
                                linetype = "dashed", color = "grey40") +
            ggplot2::geom_vline(xintercept = 0, linetype = "dashed", color = "grey40") +
            ggplot2::labs(x = "log2 fold-change",
                          y = sprintf("-log10(%s)", .volc_p_lbl),
                          color = NULL,
                          title = sprintf("%s%s",
                                          comp_name,
                                          if(is.na(pair_lbl)) "" else paste0(" — ", pair_lbl))) +
            ggplot2::theme_classic(base_size = 13) +
            ggplot2::theme(legend.position = "none")

      if(nrow(to_label) > 0){
        # ggrepel is now an Imports dep so it's always available; but keep
        # a geom_text fallback in case of unusual install paths.
        if(requireNamespace("ggrepel", quietly = TRUE)){
          p = p + ggrepel::geom_text_repel(data = to_label,
                                           ggplot2::aes(label = feature),
                                           color = "#d62728",
                                           fontface = "bold",
                                           size = 3.2,
                                           max.overlaps = Inf,
                                           force = 4,
                                           min.segment.length = 0,
                                           segment.color = "#d62728",
                                           segment.size = 0.3,
                                           show.legend = FALSE)
        } else {
          p = p + ggplot2::geom_text(data = to_label,
                                     ggplot2::aes(label = feature),
                                     color = "#d62728", fontface = "bold",
                                     size = 3.2, vjust = -0.7,
                                     show.legend = FALSE)
        }
      }
      .show_plot(p, max_width = "650px")
      cat("\n\n")

      if(nrow(sig_rows) > 0){
        sig_tbl = sig_rows %>%
          dplyr::select(feature, log2FC,
                        dplyr::any_of(c("mean_g1", "mean_g2")),
                        dplyr::all_of(.volc_p_col)) %>%
          dplyr::mutate(dplyr::across(where(is.numeric), ~ signif(., 4)))
        # kable renders reliably in results="asis" chunks that also emit plots.
        # DT htmlwidgets are not captured correctly when mixed with ggplot
        # device output in the same asis chunk.
        cap_html = sprintf(
          '<p style="font-size:13px;margin-top:8px;"><b>%d significant feature(s)</b> &mdash; %s &lt; %.3g</p>',
          nrow(sig_tbl), .volc_p_lbl, sig_threshold)
        tbl_html = knitr::kable(
          sig_tbl, format = "html", row.names = FALSE,
          table.attr = 'class="table table-sm table-striped table-bordered" style="font-size:13px;width:auto;"')
        cat(cap_html, '\n<div style="overflow-x:auto;max-height:340px;overflow-y:auto;">\n',
            tbl_html, '\n</div>\n\n', sep = "")
      }
    }
  }
}
```

## 4.5 Heatmap — general (all features) {.tabset .tabset-fade}

`r if(!heatmap_include) "_Skipped — set_ \x60heatmap.enabled: True\x60 _in YAML to enable._"`

`r if(heatmap_include && !heatmap_general_on) "_Skipped — set_ \x60heatmap.general.enabled: True\x60 _in YAML to enable._"`

`r if(heatmap_include && heatmap_general_on) sprintf("All features retained. Cells show per-feature %s of %s; samples coloured by \x60%s\x60%s. Scope: **%s** (Global and/or one tab per comparison). Feature labels: %s.", heatmap_general_cfg$scale, heatmap_general_cfg$transform, heatmap_general_cfg$color_samples_by, if(!is.null(heatmap_general_cfg$group_features_by)) sprintf(", features grouped by \x60%s\x60", heatmap_general_cfg$group_features_by) else "", heatmap_general_scope, heatmap_general_cfg$feature_labels)`

```{r heatmap, eval=heatmap_include, fig.width=max(10, ncol(bio_matrix)*0.18)*fig_scale, fig.height=max(5, nrow(bio_matrix)*0.22 + 2)*fig_scale}
# Shared heatmap renderer — called for every scope target across 4.5/4.6/4.7.
# M_raw   : samples × features numeric matrix (raw / pre-transform)
# smeta   : sample metadata data.frame, rows aligned to M_raw
# vmeta   : variable metadata data.frame with a 'Compound' column
# title   : plot title string
# cfg     : resolved render settings for the calling section (from .resolve_hm)
#           — transform / scale / color_samples_by / cluster_rows /
#           cluster_within / orientation / group_features_by / feature_labels.
# group_by / feature_labels: explicit overrides (default to the cfg values);
#           the combined split-by-class heatmap overrides group_by to the
#           subclass column.
.draw_heatmap = function(M_raw, smeta, vmeta, title, cfg = heatmap_general_cfg,
                         group_by = cfg$group_features_by,
                         feature_labels = cfg$feature_labels){
  if(!requireNamespace("pheatmap", quietly = TRUE)){
    cat("_Install `pheatmap` to render heatmaps._\n"); return(invisible(NULL))
  }
  if(nrow(M_raw) < 3 || ncol(M_raw) < 2){
    cat("_Insufficient data for heatmap._\n"); return(invisible(NULL))
  }

  M = as.matrix(M_raw)
  if(sample_id_head %in% colnames(smeta))
    rownames(M) = as.character(smeta[[sample_id_head]])

  if(cfg$transform == "log2"){
    M[M < 0 & !is.na(M)] = 0; M = log2(M + 1)
  } else if(cfg$transform == "sqrt"){
    M[M < 0 & !is.na(M)] = 0; M = sqrt(M)
  }

  vars   = apply(M, 2, stats::var, na.rm = TRUE)
  keep_f = is.finite(vars) & vars > 0
  M = M[, keep_f, drop = FALSE]
  if(ncol(M) == 0){ cat("_No variable features._\n"); return(invisible(NULL)) }

  if(cfg$scale == "z")
    M = scale(M, center = TRUE, scale = TRUE)
  M[is.na(M)] = 0
  cap = 2.5; M[M > cap] = cap; M[M < -cap] = -cap

  ann_row = data.frame(row.names = rownames(M))
  ann_row[[cfg$color_samples_by]] = smeta[[cfg$color_samples_by]]

  ann_col  = NULL; gaps_col = NULL
  if(!is.null(group_by) && group_by %in% colnames(vmeta)){
    feats       = colnames(M)
    fgroups     = as.character(vmeta[[group_by]][match(feats, vmeta$Compound)])
    fgroups[is.na(fgroups)] = "NA"
    group_levels = unique(fgroups)
    ordered_cols = character(0)
    for(g in group_levels){
      in_g = feats[fgroups == g]
      if(cfg$cluster_within && length(in_g) > 1){
        d = stats::dist(t(M[, in_g, drop = FALSE]))
        if(all(is.finite(d)))
          in_g = in_g[stats::hclust(d, method = "average")$order]
      }
      ordered_cols = c(ordered_cols, in_g)
    }
    M = M[, ordered_cols, drop = FALSE]
    fgroups2 = as.character(vmeta[[group_by]][match(ordered_cols, vmeta$Compound)])
    fgroups2[is.na(fgroups2)] = "NA"
    ann_col = data.frame(row.names = ordered_cols)
    ann_col[[group_by]] = factor(fgroups2, levels = group_levels)
    blk = rle(fgroups2)$lengths
    gaps_col = cumsum(blk); gaps_col = gaps_col[-length(gaps_col)]
  }

  ord_s   = order(as.character(ann_row[[cfg$color_samples_by]]))
  M       = M[ord_s, , drop = FALSE]
  ann_row = ann_row[ord_s, , drop = FALSE]
  blk_r   = rle(as.character(ann_row[[cfg$color_samples_by]]))$lengths
  gaps_row = cumsum(blk_r); gaps_row = gaps_row[-length(gaps_row)]

  pal = grDevices::colorRampPalette(c("#2166ac", "white", "#b2182b"))(101)
  brk = seq(-cap, cap, length.out = 102)

  # Annotation colours via the shared .group_fill_colors() so the feature-group
  # annotation (e.g. Enzymatic_pathway) matches the PCA loadings bars exactly.
  .ann_colors = function(ann){
    if(is.null(ann)) return(NULL)
    out = lapply(ann, .group_fill_colors)
    names(out) = colnames(ann); out
  }
  ann_colors = c(.ann_colors(ann_row), .ann_colors(ann_col))

  # Orientation: by default samples on y-axis (rows). "samples_x" transposes
  # explicitly. "auto" transposes when there are more samples than features —
  # the long axis goes to x so the canvas isn't tall and narrow with crowded
  # row labels.
  flip = identical(cfg$orientation, "samples_x") ||
         (identical(cfg$orientation, "auto") &&
          nrow(M) > ncol(M))
  if(flip){
    M_use         = t(M)
    ann_row_use   = ann_col
    ann_col_use   = ann_row
    gaps_row_use  = gaps_col
    gaps_col_use  = gaps_row
    cluster_rows_use = FALSE
    cluster_cols_use = cfg$cluster_rows  # original "rows" was samples
  } else {
    M_use         = M
    ann_row_use   = ann_row
    ann_col_use   = ann_col
    gaps_row_use  = gaps_row
    gaps_col_use  = gaps_col
    cluster_rows_use = cfg$cluster_rows
    cluster_cols_use = FALSE
  }

  # Feature-label density: features sit on the columns pre-flip, rows post-flip.
  # Sample labels stay readable; only the (potentially huge) feature axis is
  # shrunk / hidden per heatmap_feature_labels.
  feat_fs   = switch(feature_labels, show = 9, small = 4, 4)
  feat_show = feature_labels != "hide"
  samp_fs   = 9
  if(flip){   # features -> rows, samples -> cols
    show_rn = feat_show; fs_row = feat_fs; show_cn = TRUE; fs_col = samp_fs
  } else {    # features -> cols, samples -> rows
    show_cn = feat_show; fs_col = feat_fs; show_rn = TRUE; fs_row = samp_fs
  }

  ph = pheatmap::pheatmap(
    M_use,
    cluster_rows      = cluster_rows_use,
    cluster_cols      = cluster_cols_use,
    gaps_row          = if(length(gaps_row_use)) gaps_row_use else NULL,
    gaps_col          = if(length(gaps_col_use)) gaps_col_use else NULL,
    annotation_row    = ann_row_use,
    annotation_col    = ann_col_use,
    annotation_colors = if(length(ann_colors)) ann_colors else NA,
    color             = pal, breaks = brk,
    show_rownames     = show_rn, show_colnames = show_cn,
    fontsize_row      = fs_row, fontsize_col = fs_col,
    border_color      = NA,
    legend_breaks     = c(-cap, -1, 0, 1, cap),
    legend_labels     = c(sprintf("≤ -%g", cap), "-1", "0", "1", sprintf("≥ %g", cap)),
    main              = title,
    silent            = TRUE        # suppress auto-draw; we draw explicitly below
  )
  # Drawing approach: wrap the pheatmap gtable in a ggplot via annotation_custom().
  # This fills the full knitr canvas (respects fig.width/fig.height) and is
  # captured correctly in both regular chunks and results="asis" loops.
  # Pitfalls avoided:
  #   - print(ph$gtable)  → print.gtable emits text description, not a plot
  #   - grid.draw()        → not captured in results="asis" chunks by knitr
  #   - cowplot::plot_grid → places gtable at natural size, leaving whitespace
  p_out = ggplot2::ggplot() +
    ggplot2::annotation_custom(ph$gtable,
                               xmin = -Inf, xmax = Inf,
                               ymin = -Inf, ymax = Inf) +
    ggplot2::theme_void()
  print(p_out)
  invisible(NULL)
}

# Build the render targets for a scope selector, shared by 4.5/4.6/4.7. Each
# target is list(label, M = samples x features matrix, smeta, vmeta, comp) —
# comp is NULL for the Global target, else the comparison spec. "comparisons"
# and "all" expand the pre-defined comparisons; empty subsets are dropped.
.heatmap_targets = function(scope){
  g = list(list(label = "Global", M = bio_matrix, smeta = bio_meta,
                vmeta = bio_de$variable_meta, comp = NULL))
  cmps = list()
  if(scope %in% c("comparisons", "all"))
    cmps = Filter(Negate(is.null), lapply(comparisons, function(comp){
      cond = if(is.null(comp$subset)) list() else as.list(comp$subset)
      cond[[comp$compare$factor]] = comp$compare$levels
      de = subset_dataset(importData, conditions = cond)
      if(nrow(de$sample_meta) == 0) return(NULL)
      list(label = comp$name, M = de$data, smeta = de$sample_meta,
           vmeta = de$variable_meta, comp = comp)
    }))
  switch(scope, global = g, comparisons = cmps, all = c(g, cmps))
}

# Option 1: keep only the N most-changing features per class for a comparison.
# Ranks each feature by the chosen stats metric (p_adj / p_value: smaller is
# better; log2fc: larger |value| is better), takes the top N within each
# `group_features_by` class. Falls back to all features if top_changing is off
# or the comparison has no stats.
.top_changing = function(comp_name, feats, vmeta, group_by = heatmap_group_features_by){
  if(!heatmap_topchg_on || is.null(stats_tables) ||
     is.null(stats_tables$stats) || nrow(stats_tables$stats) == 0) return(feats)
  s = stats_tables$stats[stats_tables$stats$comparison == comp_name, , drop = FALSE]
  metric_col = switch(heatmap_topchg_rank,
                      log2fc = "log2FC", p_value = "p_value", "p_adj")
  if(nrow(s) == 0 || !metric_col %in% colnames(s)) return(feats)

  use_abs = metric_col == "log2FC"          # log2FC ranks by magnitude
  agg = tapply(s[[metric_col]], s$feature, function(v){
    v = v[is.finite(v)]
    if(!length(v)) return(NA_real_)
    if(use_abs) max(abs(v)) else min(v)     # best (biggest FC / smallest p)
  })
  score = agg[match(feats, names(agg))]     # aligned to feats; NA if no stats
  grp = if(!is.null(group_by) && group_by %in% colnames(vmeta))
          as.character(vmeta[[group_by]][match(feats, vmeta$Compound)]) else rep("all", length(feats))
  grp[is.na(grp)] = "NA"

  keep = unlist(lapply(unique(grp), function(g){
    idx = which(grp == g)                   # positions in feats for this class
    ord = order(score[idx], decreasing = use_abs, na.last = NA)  # drops NA-score feats
    feats[idx][ord][seq_len(min(heatmap_topchg_n, length(ord)))]
  }), use.names = FALSE)
  if(length(keep)) keep else feats
}

# Global-tab analogue of .top_changing when there is no single comparison to
# rank by: score each feature by how strongly it separates the `color_by` groups
# (the section's color_samples_by / primary study factor). On the transformed
# values, z-score each feature across samples, then score = variance of the
# per-group means; keep the top-N per `group_by` class. Falls back to all feats
# if the factor is absent or has < 2 usable groups.
.top_changing_global = function(feats, M, smeta, vmeta, color_by, group_by, transform){
  if(!heatmap_topchg_on) return(feats)
  if(is.null(color_by) || !color_by %in% colnames(smeta)) return(feats)
  g_all = as.character(smeta[[color_by]]); ok = !is.na(g_all)
  if(length(unique(g_all[ok])) < 2) return(feats)

  X = as.matrix(M[, feats, drop = FALSE])
  if(identical(transform, "log2")){ X[X < 0 & !is.na(X)] = 0; X = log2(X + 1) }
  else if(identical(transform, "sqrt")){ X[X < 0 & !is.na(X)] = 0; X = sqrt(X) }
  X = scale(X)[ok, , drop = FALSE]; g = g_all[ok]     # per-feature z, keep labelled rows
  score = apply(X, 2, function(col){
    m = tapply(col, g, mean, na.rm = TRUE)            # per-group means
    stats::var(as.numeric(m), na.rm = TRUE)           # spread across groups
  })
  names(score) = feats

  cls = if(!is.null(group_by) && group_by %in% colnames(vmeta))
          as.character(vmeta[[group_by]][match(feats, vmeta$Compound)]) else rep("all", length(feats))
  cls[is.na(cls)] = "NA"
  keep = unlist(lapply(unique(cls), function(cc){
    idx = which(cls == cc)
    ord = order(score[idx], decreasing = TRUE, na.last = NA)
    feats[idx][ord][seq_len(min(heatmap_topchg_n, length(ord)))]
  }), use.names = FALSE)
  if(length(keep)) keep else feats
}

# Figure-canvas sizing shared by 4.5/4.6/4.7. Each asis section renders many
# heatmaps into ONE knitr canvas (single fig size per chunk), so size to the
# largest target using that block's orientation. .hm_scope_nrow = max sample
# count across a scope's targets (used for the height/width formula).
.hm_fig_dims = function(max_nrow, n_feat, orientation, fig_scale = 1){
  flip = orientation == "samples_x" || (orientation == "auto" && max_nrow > n_feat)
  w = if(flip) max_nrow * 0.30 else n_feat  * 0.20
  h = if(flip) n_feat  * 0.18 else max_nrow * 0.27
  list(w = max(10, w) * fig_scale, h = max(6, h + 2) * fig_scale)
}
.hm_scope_nrow = function(scope){
  n = vapply(.heatmap_targets(scope), function(tg) nrow(tg$M), integer(1))
  if(length(n)) max(n) else nrow(bio_matrix)
}
```

```{r heatmap-general-precompute, include=FALSE, eval=heatmap_include && heatmap_general_on}
.tg45 = .heatmap_targets(heatmap_general_scope)
.d45  = .hm_fig_dims(.hm_scope_nrow(heatmap_general_scope), ncol(bio_matrix),
                     heatmap_general_cfg$orientation, fig_scale)
```

```{r heatmap-general, results="asis", eval=heatmap_include && heatmap_general_on, fig.width=(if(exists(".d45")) .d45$w else 10), fig.height=(if(exists(".d45")) .d45$h else 6), out.width="100%"}
if(!length(.tg45))
  cat(sprintf("_No render targets for scope '%s'._\n\n", heatmap_general_scope))
for(tg in .tg45){
  cat(sprintf("\n### %s {-}\n\n", tg$label))
  cat(sprintf("_%d samples × %d features_\n\n", nrow(tg$M), ncol(tg$M)))
  .draw_heatmap(tg$M, tg$smeta, tg$vmeta,
                sprintf("%s of %s — %s", heatmap_general_cfg$scale,
                        heatmap_general_cfg$transform, tg$label),
                cfg = heatmap_general_cfg)
  cat("\n\n")
}
```

## 4.6 Heatmap — top-changing per class {.tabset .tabset-fade}

`r if(heatmap_include && !heatmap_topchg_on) "_Skipped — set_ \x60heatmap.top_changing.enabled: True\x60 _in YAML to enable._"`

`r if(heatmap_include && heatmap_topchg_on) sprintf("Top %d features per \x60%s\x60 class. Comparison tabs rank by %s; the Global tab ranks by separation across \x60%s\x60 (no single contrast). Scope: **%s**. Feature labels: %s.", heatmap_topchg_n, heatmap_topchg_cfg$group_features_by %||% "class", heatmap_topchg_rank, heatmap_topchg_cfg$color_samples_by, heatmap_topchg_scope, heatmap_topchg_cfg$feature_labels)`

```{r heatmap-topchg-precompute, include=FALSE, eval=heatmap_include && heatmap_topchg_on}
.tg46 = .heatmap_targets(heatmap_topchg_scope)
.gf46 = heatmap_topchg_cfg$group_features_by
.n_cls46 = if(!is.null(.gf46) && .gf46 %in% colnames(bio_de$variable_meta)){
  v = as.character(bio_de$variable_meta[[.gf46]]); length(unique(v[!is.na(v)]))
} else 1L
.d46 = .hm_fig_dims(.hm_scope_nrow(heatmap_topchg_scope),
                    max(2, min(ncol(bio_matrix), heatmap_topchg_n * .n_cls46)),
                    heatmap_topchg_cfg$orientation, fig_scale)
```

```{r heatmap-topchg, results="asis", eval=heatmap_include && heatmap_topchg_on, fig.width=(if(exists(".d46")) .d46$w else 10), fig.height=(if(exists(".d46")) .d46$h else 6), out.width="100%"}
if(!length(.tg46))
  cat(sprintf("_No render targets for scope '%s'._\n\n", heatmap_topchg_scope))
for(tg in .tg46){
  cat(sprintf("\n### %s {-}\n\n", tg$label))
  kf = if(is.null(tg$comp))
         .top_changing_global(colnames(tg$M), tg$M, tg$smeta, tg$vmeta,
                              heatmap_topchg_cfg$color_samples_by,
                              heatmap_topchg_cfg$group_features_by,
                              heatmap_topchg_cfg$transform)
       else
         .top_changing(tg$comp$name, colnames(tg$M), tg$vmeta,
                       group_by = heatmap_topchg_cfg$group_features_by)
  Mk = tg$M[, kf, drop = FALSE]
  rank_desc = if(is.null(tg$comp)) sprintf("%s separation", heatmap_topchg_cfg$color_samples_by) else heatmap_topchg_rank
  cat(sprintf("_%d samples × %d features (top %d per class by %s)_\n\n",
              nrow(Mk), ncol(Mk), heatmap_topchg_n, rank_desc))
  .draw_heatmap(Mk, tg$smeta, tg$vmeta,
                sprintf("%s of %s — %s (top-changing)", heatmap_topchg_cfg$scale,
                        heatmap_topchg_cfg$transform, tg$label),
                cfg = heatmap_topchg_cfg)
  cat("\n\n")
}
```

## 4.7 Heatmap — by class {.tabset .tabset-fade}

`r if(heatmap_include && (!heatmap_split_on || is.null(heatmap_subclass_col))) "_Skipped — set_ \x60heatmap.split_by_class.enabled: True\x60 _and a_ \x60subclass_col\x60 _in YAML to enable._"`

`r if(heatmap_include && heatmap_split_on && !is.null(heatmap_subclass_col)) sprintf("Split by \x60%s\x60 (%s). Scope: **%s**. %s Feature labels: %s.", heatmap_subclass_col, if(length(heatmap_subclass_vals)) paste(heatmap_subclass_vals, collapse=", ") else "all values", heatmap_split_scope, if(heatmap_split_multi=="combined") "Selected values combined in one heatmap (grouped by class)." else "One heatmap per value.", heatmap_split_cfg$feature_labels)`

```{r heatmap-split-precompute, include=FALSE, eval=heatmap_include && heatmap_split_on && !is.null(heatmap_subclass_col)}
.tg47 = .heatmap_targets(heatmap_split_scope)
.d47  = .hm_fig_dims(.hm_scope_nrow(heatmap_split_scope), ncol(bio_matrix),
                     heatmap_split_cfg$orientation, fig_scale)
```

```{r heatmap-split, results="asis", eval=heatmap_include && heatmap_split_on && !is.null(heatmap_subclass_col), fig.width=(if(exists(".d47")) .d47$w else 10), fig.height=(if(exists(".d47")) .d47$h else 6), out.width="100%"}
# Per scope target, split features by `subclass_col`. multi_class == "combined"
# draws all selected values in one heatmap (grouped by the subclass column);
# "tabs" draws one heatmap per value. `subclass_values` (if set) focuses values.
split_col = heatmap_subclass_col
for(tg in .tg47){
  vmeta_t = tg$vmeta
  if(!split_col %in% colnames(vmeta_t)){
    cat(sprintf("\n### %s {-}\n\n_subclass_col '%s' not in feature_metadata — skipped._\n\n",
                tg$label, split_col)); next
  }
  vals = sort(unique(as.character(vmeta_t[[split_col]])))
  vals = vals[!is.na(vals) & nzchar(vals)]
  if(length(heatmap_subclass_vals))
    vals = intersect(vals, as.character(heatmap_subclass_vals))
  if(!length(vals)){
    cat(sprintf("\n### %s {-}\n\n_no values match subclass_values._\n\n", tg$label)); next
  }
  if(heatmap_split_multi == "combined"){
    feats_v = intersect(vmeta_t$Compound[as.character(vmeta_t[[split_col]]) %in% vals],
                        colnames(tg$M))
    if(length(feats_v) < 2){
      cat(sprintf("\n### %s {-}\n\n_too few features._\n\n", tg$label)); next
    }
    cat(sprintf("\n### %s {-}\n\n", tg$label))
    cat(sprintf("_%d samples × %d features across %s: %s_\n\n",
                nrow(tg$M), length(feats_v), split_col, paste(vals, collapse = ", ")))
    .draw_heatmap(tg$M[, feats_v, drop = FALSE], tg$smeta, vmeta_t,
                  sprintf("%s — %s", tg$label, paste(vals, collapse = " / ")),
                  cfg = heatmap_split_cfg, group_by = split_col)
    cat("\n\n")
  } else {
    for(v in vals){
      feats_v = intersect(vmeta_t$Compound[as.character(vmeta_t[[split_col]]) == v],
                          colnames(tg$M))
      if(length(feats_v) < 2) next
      cat(sprintf("\n### %s — %s: %s {-}\n\n", tg$label, split_col, v))
      cat(sprintf("_%d samples × %d features_\n\n", nrow(tg$M), length(feats_v)))
      .draw_heatmap(tg$M[, feats_v, drop = FALSE], tg$smeta, vmeta_t,
                    sprintf("%s — %s = %s", tg$label, split_col, v),
                    cfg = heatmap_split_cfg)
      cat("\n\n")
    }
  }
}
```

# 5. Ion ratios

User-defined compound ratios (e.g. EpOME/DiHOME as sEH activity proxy). Each
ratio is `numerator / denominator` per sample; denominator ≤ 0 gives NA.
Wilcoxon p-value shown per comparison panel (Kruskal–Wallis for ≥3 groups).
Ratios are **not** carried into heatmap, PCA, volcano, or the comparisons table.

`r if(!ion_ratios_enabled) "_No ion ratios defined — add an_ \x60ion_ratios:\x60 _block with a_ \x60ratios:\x60 _list to_ \x60stats_report\x60 _in the YAML to enable this section._"`

```{r ion-ratios, results="asis", eval=ion_ratios_enabled, fig.width=max(8, length(comparisons)*3.5)*fig_scale, fig.height=5.5*fig_scale}
has_comps_ir  = length(comparisons) > 0
has_factor_ir = boxplot_enabled && length(boxplot_factors) > 0

for(ratio_cfg in ion_ratios_list){
  num_col   = as.character(ratio_cfg$num)
  den_col   = as.character(ratio_cfg$den)
  ratio_lbl = as.character(ratio_cfg$label %||% paste0(num_col, " / ", den_col))

  ok_num   = num_col %in% colnames(data_matrix)
  ok_den   = den_col %in% colnames(data_matrix)
  use_tabs = has_comps_ir && has_factor_ir

  if(use_tabs) cat(sprintf("\n\n## %s {.tabset}\n\n", ratio_lbl))
  else         cat(sprintf("\n\n## %s {-}\n\n", ratio_lbl))
  cat(sprintf("_Ratio:_ `%s` / `%s`\n\n", num_col, den_col))

  if(!ok_num || !ok_den){
    if(!ok_num) cat(sprintf("_Numerator `%s` not found in data._\n\n", num_col))
    if(!ok_den) cat(sprintf("_Denominator `%s` not found in data._\n\n", den_col))
    next
  }

  ratio_all                        = as.numeric(data_matrix[[num_col]]) /
                                     as.numeric(data_matrix[[den_col]])
  ratio_all[!is.finite(ratio_all)] = NA
  names(ratio_all)                 = rownames(data_matrix)

  # ---- By comparison -------------------------------------------------------
  if(has_comps_ir){
    if(use_tabs) cat("\n\n### By comparison {-}\n\n")

    long_rows_ir = list()
    anno_rows_ir = list()

    for(comp in comparisons){
      if(!isTRUE(comp$enabled %||% TRUE)) next
      cond = if(is.null(comp$subset)) list() else as.list(comp$subset)
      cond[[comp$compare$factor]] = comp$compare$levels
      de_sub = tryCatch(subset_dataset(importData, conditions = cond),
                        error = function(e) NULL)
      if(is.null(de_sub) || nrow(de_sub$sample_meta) == 0) next

      sub_rn    = rownames(de_sub$sample_meta)
      sub_ratio = ratio_all[sub_rn]
      sub_grp   = factor(de_sub$sample_meta[[comp$compare$factor]],
                         levels = comp$compare$levels)
      sub_sid   = if(sample_id_head %in% colnames(de_sub$sample_meta))
                    as.character(de_sub$sample_meta[[sample_id_head]]) else sub_rn
      sub_gsf   = if(global_summary_factor %in% colnames(de_sub$sample_meta))
                    as.character(de_sub$sample_meta[[global_summary_factor]]) else NA_character_

      long_rows_ir[[length(long_rows_ir) + 1]] = data.frame(
        .comparison = comp$name,
        .group      = sub_grp,
        .sample_id  = sub_sid,
        ratio_value = sub_ratio,
        .tooltip    = .point_tooltip(sub_sid, sub_gsf, sub_ratio, ylab = "ratio"),
        stringsAsFactors = FALSE
      )

      ngrp = length(comp$compare$levels)
      pval = tryCatch({
        if(ngrp == 2){
          g1 = sub_ratio[sub_grp == comp$compare$levels[1]]
          g2 = sub_ratio[sub_grp == comp$compare$levels[2]]
          wilcox.test(g1, g2, exact = FALSE)$p.value
        } else {
          kruskal.test(sub_ratio, sub_grp)$p.value
        }
      }, error = function(e) NA_real_)

      plab = if(is.na(pval))  "p = NA"    else
             if(pval < 0.001) "p < 0.001" else
             sprintf("p = %.3f", pval)
      if(ngrp > 2) plab = paste0("KW ", plab) else plab = paste0("W ", plab)

      anno_rows_ir[[length(anno_rows_ir) + 1]] = data.frame(
        .comparison = comp$name,
        .group      = comp$compare$levels[length(comp$compare$levels)],
        label       = plab,
        stringsAsFactors = FALSE
      )
    }

    if(length(long_rows_ir) == 0){
      cat("_No valid comparisons for this ratio._\n\n")
    } else {
      long_ir        = dplyr::bind_rows(long_rows_ir)
      long_ir$.group = factor(long_ir$.group, levels = unique(long_ir$.group))
      anno_ir        = dplyr::bind_rows(anno_rows_ir)
      anno_ir$.group = factor(anno_ir$.group, levels = levels(long_ir$.group))

      y_lim = quantile(long_ir$ratio_value, prob = yaxis_quant, na.rm = TRUE) * yaxis_scalar
      y_ann = y_lim * 0.97

      p_base = ggplot2::ggplot(long_ir,
                               ggplot2::aes(x = .group, y = ratio_value,
                                            fill = .group)) +
        ggplot2::coord_cartesian(ylim = c(0, y_lim)) +
        ggplot2::labs(x = NULL, y = "ratio", title = ratio_lbl) +
        ggplot2::facet_wrap(~ .comparison, scales = "free_x", nrow = 1) +
        ggplot2::geom_text(
          data        = anno_ir,
          ggplot2::aes(x = .group, label = label),
          y           = y_ann,
          hjust       = 0.5, vjust = 1,
          size        = 3.2, fontface = "italic", color = "grey30",
          inherit.aes = FALSE
        ) +
        boxplot_theme

      plots_ir = .compound_plots(p_base, plot_type, plot_error_bar)
      for(pname in names(plots_ir)){
        if(plot_type == "both")
          cat(sprintf('<div style="margin-top:10px;"><b>%s</b></div>\n',
                      if(pname == "box") "Boxplot" else "Barplot"))
        .show_plot(plots_ir[[pname]])
      }
      cat("\n\n")
    }
  }

  # ---- By factor -----------------------------------------------------------
  if(has_factor_ir){
    if(use_tabs) cat("\n\n### By factor {-}\n\n")

    keep_ir  = metadata[[boxplot_head]] %in% boxplot_levels
    ir_meta  = metadata[keep_ir, , drop = FALSE]
    ir_ratio = ratio_all[rownames(ir_meta)]

    for(f in boxplot_factors){
      if(!f %in% colnames(ir_meta)) next
      sid_ir = if(sample_id_head %in% colnames(ir_meta))
                 as.character(ir_meta[[sample_id_head]]) else rownames(ir_meta)
      df_f = data.frame(group     = as.character(ir_meta[[f]]),
                        value     = ir_ratio,
                        sample_id = sid_ir,
                        gsf       = if(global_summary_factor %in% colnames(ir_meta))
                                      as.character(ir_meta[[global_summary_factor]]) else NA_character_,
                        stringsAsFactors = FALSE)
      df_f = df_f[!is.na(df_f$group), ]
      if(nrow(df_f) == 0) next

      df_f$.tooltip = .point_tooltip(df_f$sample_id, df_f$gsf, df_f$value, ylab = "ratio")

      y_lim_f = quantile(df_f$value, prob = yaxis_quant, na.rm = TRUE) * yaxis_scalar
      p_f = ggplot2::ggplot(df_f, ggplot2::aes(group, value, fill = group)) +
        ggplot2::coord_cartesian(ylim = c(0, y_lim_f)) +
        ggplot2::scale_fill_brewer(palette = "Set1", na.value = "grey60") +
        ggplot2::labs(x = NULL, y = "ratio",
                      title = sprintf("%s — by %s", ratio_lbl, f)) +
        boxplot_theme

      plots_f = .compound_plots(p_f, plot_type, plot_error_bar)
      for(pname in names(plots_f)){
        if(plot_type == "both")
          cat(sprintf('<div style="margin-top:10px;"><b>%s</b></div>\n',
                      if(pname == "box") "Boxplot" else "Barplot"))
        .show_plot(plots_f[[pname]])
      }
      cat("\n\n")
    }
  }
}
```

# 6. Correlations (feature × feature) {.tabset}

Pairwise feature correlations within each subset, one heatmap per
(correlation × subset × method).

> **Long table of all pairs (estimates, statistics, p-values, p_adj) is
> in the `correlations` tab of the stats xlsx.**

## Heatmaps {-}

```{r corr-heatmaps, results="asis", eval=!is.null(stats_tables) && nrow(stats_tables$correlations) > 0, fig.width=8*fig_scale, fig.height=7*fig_scale}
# One heatmap per (correlation × subset × method).
# Optional clustering controlled by `correlations.entries[i].cluster:` in
# YAML; `correlations` (unpacked at the top) is the entries list.
corr_yaml = correlations
get_yaml_opt = function(name, opt, default){
  for(c in corr_yaml) if(identical(c$name, name) && !is.null(c[[opt]])) return(c[[opt]])
  default
}

cdf = stats_tables$correlations
combos = unique(cdf[, c("correlation", "subset", "method")])

for(i in seq_len(nrow(combos))){
  cn = combos$correlation[i]; sb = combos$subset[i]; mt = combos$method[i]
  cluster_yn   = isTRUE(get_yaml_opt(cn, "cluster",      TRUE))
  show_values  = isTRUE(get_yaml_opt(cn, "show_values",  FALSE))

  cat(sprintf("\n### %s — %s — %s {-}\n\n", cn, sb, mt))

  sub = cdf[cdf$correlation == cn & cdf$subset == sb & cdf$method == mt, ,
            drop = FALSE]
  if(nrow(sub) == 0){ cat("_(empty)_\n\n"); next }

  # Reshape lower-triangle long → full square matrix.
  feats = unique(c(sub$feature_a, sub$feature_b))
  M = matrix(NA_real_, nrow = length(feats), ncol = length(feats),
             dimnames = list(feats, feats))
  for(k in seq_len(nrow(sub))){
    a = sub$feature_a[k]; b = sub$feature_b[k]; r = sub$estimate[k]
    M[a, b] = r; M[b, a] = r
  }
  # Self-self cells are dropped from the long table by default — leave them
  # as NA so the heatmap renders the diagonal in `na.value` grey, signalling
  # "self-correlation, omitted". (When include_self: True they're filled.)

  # Feature order: group by pathway (contiguous blocks) and cluster within
  # each group, mirroring the general heatmap. Falls back to global clustering
  # when no group_features_by column is available. `cluster:` (per-entry YAML)
  # toggles the within-group (or global) clustering step.
  grp_col     = heatmap_group_features_by
  vmeta_here  = if(exists("bio_de")) bio_de$variable_meta else NULL
  have_groups = !is.null(grp_col) && !is.null(vmeta_here) &&
                grp_col %in% colnames(vmeta_here) && "Compound" %in% colnames(vmeta_here)

  clust_within = function(fs){
    if(cluster_yn && length(fs) > 2){
      dg = tryCatch(stats::as.dist(1 - abs(M[fs, fs, drop = FALSE])),
                    error = function(e) NULL)
      hg = if(!is.null(dg)) tryCatch(stats::hclust(dg, method = "average"),
                                     error = function(e) NULL) else NULL
      if(!is.null(hg)) fs = fs[hg$order]
    }
    fs
  }

  if(have_groups){
    fgroups = as.character(vmeta_here[[grp_col]][match(feats, vmeta_here$Compound)])
    fgroups[is.na(fgroups)] = "NA"
    feats = unlist(lapply(unique(fgroups),
                          function(g) clust_within(feats[fgroups == g])),
                   use.names = FALSE)
  } else {
    feats = clust_within(feats)
  }

  # Feature -> pathway colour (shared palette, NA-safe): drives the axis
  # annotation blocks and the legend.
  grp_vals = if(have_groups){
    gv = as.character(vmeta_here[[grp_col]][match(feats, vmeta_here$Compound)])
    gv[is.na(gv)] = "NA"; gv
  } else rep("NA", length(feats))
  pal       = if(have_groups) .group_fill_colors(grp_vals) else NULL
  feat_cols = if(have_groups) unname(pal[grp_vals]) else rep("grey70", length(feats))
  names(feat_cols) = feats

  # Cut the (symmetric) square to a single triangle: keep each unordered pair
  # once (feature index a < b), plus the diagonal only when self-correlations
  # are present (include_self). The x-axis feature order is reversed ("flipped").
  # A dedicated strip level carries the pathway colour blocks along each axis
  # (like the pheatmap annotation bars), so axis text can stay black.
  strip_lvl = "__grp__"                                 # label blanked below
  xlev = if(have_groups) c(strip_lvl, rev(feats)) else rev(feats)
  ylev = if(have_groups) c(strip_lvl, rev(feats)) else rev(feats)

  long = expand.grid(feature_a = feats, feature_b = feats, stringsAsFactors = FALSE)
  long$r = M[cbind(long$feature_a, long$feature_b)]
  ia = match(long$feature_a, feats); ib = match(long$feature_b, feats)
  keep_tri = ia > ib | (ia == ib & !is.na(long$r))      # flip to `ia < ib` for the other half
  long = long[keep_tri, , drop = FALSE]
  long$feature_a = factor(long$feature_a, levels = xlev)
  long$feature_b = factor(long$feature_b, levels = ylev)

  blank_strip = function(l) ifelse(l == strip_lvl, "", l)

  p = ggplot2::ggplot(long, ggplot2::aes(feature_a, feature_b, fill = r)) +
        ggplot2::geom_tile(color = "white", linewidth = 0.1) +
        ggplot2::scale_fill_gradient2(low = "#2166ac", mid = "white",
                                      high = "#b2182b", midpoint = 0,
                                      limits = c(-1, 1), na.value = "grey90") +
        ggplot2::scale_x_discrete(labels = blank_strip, drop = FALSE, na.translate = FALSE) +
        ggplot2::scale_y_discrete(labels = blank_strip, drop = FALSE, na.translate = FALSE) +
        ggplot2::coord_fixed() +
        ggplot2::theme_classic(base_size = 10) +
        ggplot2::theme(
          axis.text.x = ggplot2::element_text(angle = 90, vjust = 0.5,
                                              hjust = 1, size = 6),
          axis.text.y = ggplot2::element_text(size = 6),
          axis.ticks  = ggplot2::element_blank(),
          legend.position = "right"
        ) +
        ggplot2::labs(x = NULL, y = NULL, fill = mt,
                      title = sprintf("%s — %s — %s", cn, sb, mt))

  # Pathway colour blocks: a one-cell strip per axis (left column + bottom row),
  # filled directly (bypasses the correlation gradient). Invisible points carry
  # the discrete colour scale so the pathway legend still appears.
  if(have_groups){
    x_strip = data.frame(feature_a = factor(feats,     levels = xlev),
                         feature_b = factor(strip_lvl, levels = ylev),
                         stringsAsFactors = FALSE)
    y_strip = data.frame(feature_a = factor(strip_lvl, levels = xlev),
                         feature_b = factor(feats,     levels = ylev),
                         stringsAsFactors = FALSE)
    # Invisible points (positions dropped as NA) carry only the colour scale so
    # the pathway legend renders. Numeric NA_real_ is used deliberately — a
    # factor NA would make the discrete axes render a phantom "NA" row/column.
    grp_legend = data.frame(x = NA_real_, y = NA_real_,
                            grp = factor(names(pal), levels = names(pal)))
    p = p +
      ggplot2::geom_tile(data = x_strip, ggplot2::aes(feature_a, feature_b),
                         fill = feat_cols[feats], inherit.aes = FALSE) +
      ggplot2::geom_tile(data = y_strip, ggplot2::aes(feature_a, feature_b),
                         fill = feat_cols[feats], inherit.aes = FALSE) +
      ggplot2::geom_point(data = grp_legend,
                          ggplot2::aes(x = x, y = y, colour = grp),
                          inherit.aes = FALSE, na.rm = TRUE, size = 3) +
      ggplot2::scale_colour_manual(values = pal, name = grp_col, drop = FALSE) +
      ggplot2::guides(colour = ggplot2::guide_legend(
        override.aes = list(size = 3, shape = 15)))
  }

  if(show_values && length(feats) <= 30){
    p = p + ggplot2::geom_text(ggplot2::aes(label = sprintf("%.2f", r)),
                               size = 2)
  }
  print(p); cat("\n\n")
}
```

# 7. Linear models

For each feature we fit `feature ~ formula` (one `lm()` per feature).
Each row of the table is **one coefficient × feature**.

**Reading the terms** (using a model like `~ Treatment + Sex` with reference
levels `Treatment = Ctrl`, `Sex = Female`):

- **`Intercept (ref: Treatment=Ctrl; Sex=Female)`** — the model's expected mean
  value of the feature for a sample at all reference levels (a Ctrl female).
  Often biologically uninteresting on raw intensity data; mostly a baseline.
- **`Treatment: HDM_Curdlan (vs Ctrl)`** — how much higher (positive estimate)
  or lower (negative) the feature is in HDM_Curdlan vs Ctrl, **after adjusting
  for Sex**. This is the treatment effect controlling for sex imbalance.
- **`Sex: Male (vs Female)`** — how much the feature differs in males vs females,
  **after adjusting for Treatment**.
- **Continuous predictors** (e.g. `~ Treatment + Age`) get a single term —
  the per-unit slope of the feature on that variable.
- **Interactions** (e.g. `~ Treatment * Sex`) add a `Treatment:Sex` term that
  asks "does the treatment effect itself differ between males and females".

**`significant` (p_adj < `r sig_threshold`)**: the coefficient is statistically
distinguishable from zero after Benjamini-Hochberg correction across all
features × terms in this run.

`term_raw` keeps the original `lm()` coefficient name (e.g.
`TreatmentHDM_Curdlan`) for joining with downstream scripts.

> **Per-model coefficient tables (estimate, std_error, statistic, p_value,
> p_adj, significant) are in the `linear_models` tab of the stats xlsx.**

# 8. Session info

```{r session}
tryCatch(sessionInfo(), error = function(e){
  cat("sessionInfo() failed (likely a malformed DESCRIPTION in an installed package):\n")
  cat(conditionMessage(e), "\n\n")
  cat("R version:", R.version$version.string, "\n")
  cat("Platform :", R.version$platform, "\n")
  cat("Loaded packages:\n")
  cat(paste(" ", sort(loadedNamespaces()), collapse = "\n"), "\n")
})
```
